
# Financial data

```{r include=FALSE}
rm(list = ls())
library(vembedr)
library(tidyquant)
library(dplyr)
library(tidyr)
library(ggplot2)
library(quantmod)
library(lubridate)
```

Financial modeling starts with a data question: which observations are needed, where do they come from, and how should they be transformed before any model is estimated? Finance has always been data-intensive.[^3] Prices, volume, interest rates, exchange rates, macroeconomic indicators, balance-sheet items, and market indexes are the raw material for valuation, trading, portfolio construction, and risk measurement.

[^3]: According to market intelligence company IDC, the 'Global Datasphere' in 2018 reached 18 zettabytes. The vast majority of the world's data has been created in the last few years and this astonishing growth of data shows no sign of slowing down. In fact, IDC predicts the world's data will grow to 175 zettabytes in 2025. One zettabyte is 1000000000000000000000 bytes. In scientific notation this is 1e+21, or 1 with 21 zeros at the right, also called one sextillion.

The chapter builds the first part of the book's workflow. It downloads market and economic data, inspects the structure of the returned objects, visualizes prices, and introduces simple indicators such as trading range, moving averages, and Bollinger Bands. The emphasis is practical: before returns, signals, beta regressions, allocation, or VaR can be computed, the data must be obtained and understood.

Some financial data is public and easy to download; other data is private, delayed, licensed, incomplete, or available only as summaries. That distinction matters because a model inherits the limitations of its data source. In this chapter, we work with public examples that can be reproduced directly in R.

<!-- ## Financial statements. -->

<!-- Financial data comes in a wide variety of formats, and is reported and stored in numerous electronic sources. Financial statements represent a great source to analyze and understand the financial health of a firm, individuals and even governments. Financial statements are standard and well-organized reports that summarize important financial accounting information of a business. The traditional approach to analyze financial statements is looking at those in the firm's annual reports, or downloading them in PDF or Excel format in the firm's investor website site. -->

<!-- In R, we can download public firm's financial statements from the US Securities and Exchange Commission using the *finreportr* package (there are others like *finstr*). This is useful to evaluate a firm performance and conduct financial analysis as we do not need to visit the firm site. In fact, we could replicate or extend the full analysis by running the code instead of downloading newer financial statements. -->

<!-- In case you are not familiar with R packages, you should know you need to download the packages you need before using them. This is an easy process in RStudio. First, select the "Tools" menu, and then "Install Packages..." option. Now, type the name of the package you need, in this case type *finreportr* for accessing financial statements (later we will need *tidyquant* as well and others), leave the default options as they are, and click "Install". You will see a typical bar progress of this download process and R will let you know when this is done. You only need to download packages once, but you have to load the package (using the function *library*) every time you will use the package in your script.

<!-- If you are interested in working with financial statements, you should look for the *finreportr* package documentation online or users' examples available in Internet. Perhaps the main drawback of this package is that it may be slow, you normally need to wait for R to download the financial reports. Let's start with some examples here. First, we load the previously downloaded packages. -->

<!-- ```{r} -->
<!-- # Load the tidyquant package to manipulate the data and more. -->
<!-- library(tidyquant) -->
<!-- library(dplyr) -->
<!-- library(tidyr) -->
<!-- ``` -->





<!-- This is why we are interested firms to perform well in the economy. If firms are well managed, they are expected to take good strategic decisions to increase the value of the firm. In particular, the firm looks for funds from banks or owners, then invests in assets to produce or provide a service to the market. If things go well, sales increase and if revenues are greater than costs, the firm makes profits. When the firm increases their assets, they might also need more people to produce or manage those assets, then employment increases. More jobs mean more families with available income, and families will spend their income and stimulate the whole economy further. Even government may increase their tax revenues to (hopefully) spend them into high-quality public goods to increase society's welfare. Not only that, as a firm performs well, the whole value chain straightens as the firm requires inputs and services from other firms to operate. We are interested firms to perform well in the economy because that is good for the population in general and to other firms as well. Private investment stimulates the whole economy. -->

<!-- Strictly speaking, firms (regardless of their size) do not own what they have. Firm's assets were originally bought or financed using others' money (liabilities) and/or owners' money (equity). In simple terms, people are interested to finance the firm activities because the firm is supposed to know how to invest those funds in productive assets to make profits as in the case of Meta. When the financial statements show that firms take bad decisions, their assets do not produce profits and people will not be very interested in financing a non-profitable firm. In good times, people who finance productive firms receive interests, returns, or dividends as a compensation for putting their money at risk. -->

<!-- Firms have incentives to be and keep being financially healthy in the future to attract financing more easily. Public firms (as those participating in the stock market) are subject to strict regulations and public scrutiny so everybody can constantly evaluate the firm performance. Regulations do not prevent all frauds or bankruptcies, but they are intended to minimize irregular and illegal practices. We were able to freely download financial statements because regulators ask public firms to do so given some deadlines and specific guidelines. Yearly and quarterly financial statements are analyzed in detail to evaluate firms as projects, and the main interest is the relationship between risk and return, so people can make better investment decisions. The decision of financing a firm operation is risky, but the reward is supposed to be attractive enough to compensate that risk. In finance we are interested to measure how risky are the investment opportunities in the form of firms, projects, and financial assets in general. When we do a financial analysis, we also compare the performance of one company not only with respect to previous years, but also with respect to other firms, and also with respect to their own industry or sector. This is not difficult as the regulators ask public firms to publish their financial statements in a similar manner, so comparisons among different firms are valid. -->

<!-- Finance is about people taking decisions under uncertainty with the help of data analysis. Think in two kind of people in the world: those who have money but no ideas (investors), and those who have ideas but no money (firms or entrepreneurs). We refer to ideas as business ideas, innovations, business projects that require funds to get started. Those with ideas and money are rare in this world, and those without ideas and without money are in trouble. When the financial system works well, productive projects will succeed at attracting funds, firms will invest in productive assets, private investment grows, and we should expect economic growth, and hopefully economic development improves. When the financial system fails, good projects might fail to get funding and probably bad projects get financing. This could be frustrating. -->

<!-- People with money and no ideas are looking for opportunities for their money to grow. They basically face two alternatives: to save their money (put it in the bank and receive a small but certain return), or invest their money in risky projects offered by firms or entrepreneurs. The final decision is taken based on a risk-return analysis, with the help of data analysis and financial models. Imagine the bank offers a secure 4%, and the firm offers a risky 3.5% return (assume it could be as low as −1% or as high as 4%). In that case, most of us might agree (if you are risk averse as me) that the funds are better in a bank. Now assume a different situation. The bank offers a secure 4%, and the firm offers a risky 6% return (it could be as low as 3.5% and as high as 8%). In that case, there might be some investors that after doing some analysis finally decides to put their money at risk. Things become interesting and complex because the same project can be attractive for some investors and bad for others as they might have different tools to analyze risky projects and they might also have different risk appetite levels. -->

<!-- People with ideas but no money are looking for funds to run their projects. If the project is impressively good, then the firm could ask for funds in exchange of a reduced return. The project is then not too risky and the return is also low. On the other hand, if the project looks OK but entails a high risk, the firm or entrepreneur should be willing to pay high returns to investors to compensate for the associated risk. This is why risk and return move together, the higher the risk the higher the return and vice-versa. If this rule fails, then we are in trouble. -->

<!-- Interestingly, an excess of private investment is not always good for the economy. Imagine the interest rate (return of savings, and the cost of asking for a loan) is low and the economic perspectives look good. In that case, we have good conditions for the private investment to grow as it is cheap to ask for financing. In the extreme, families now have more money and they start spending a lot. Aggregate demand for all goods and services increase and consequently the prices of all goods and services increase as well. The country is experiencing an increase in the inflation levels. Then, even though people have money in their pockets, they are not able to buy as prices increase faster than families' income. In order to reduce inflation levels, the monetary authority increases the interest rates, making savings more attractive and investment less attractive. If they succeed, people will spend less and save more. This decreases the aggregate demand for goods and services and prices could decrease again to desired levels. To add more drama, imagine that this increase in interest rates is so high that now economic growth is slowing down and the risk of falling into a recession increase. If that happens, then the monetary authority should decrease the interest rates again to incentive people to stop saving and start investing again. This is why you see that interest rates go up and down, although we have had low interest rates for a while. -->

<!-- Let's go back to finance and illustrate the process of a firm looking for funds, the difference between private and public firms, and see how a firm can grow or go bankrupt in a simple but illustrative example. -->

<!-- Firms can sell bonds or stocks as a way to get funds. Let's focus on the stocks. By selling stocks, the firm can increase their equity and consequently increase their assets. People who buy stocks expect the firm to make wise investment decisions to get a return in the future. Initially, in time zero $t=0$, assume the firm is private and has the following balance sheet: -->

<!-- ```{r} -->
<!-- # The firm is private now.  -->
<!-- private_firm_bs <- c(assets = 100, liabilities = 40, equity = 60) -->
<!-- # Print results. -->
<!-- private_firm_bs -->
<!-- ``` -->

<!-- Private firms cannot sell stocks in the stock market, only public firms can. Selling stocks in the stock market is attractive for some firms because it is a way for them to grow. Stock markets allow more participant firms as long as they fulfill some strict requirements. Given the balance sheet above, the owner of this private firm invested 60 USD. When the firm goes 100% public, then the owner will sell the 60 USD investment to others. -->

<!-- In $t=1$, with the objective of getting fresh funds, the firm decides to go 100% public and sell 60 stocks at 1 USD each. Imagine that the firm succeeded at selling all 60 stocks at 1 USD. This is because investors who participate in the stock market consider that this is a fair value for the firm. These investors become new owners; this is why we say the firm is now public. So, by the end of $t=1$ we have: -->

<!-- ```{r} -->
<!-- # The firm is public now.  -->
<!-- public_firm_bs <- c(assets = 100, liabilities = 40,  -->
<!--                     equity_60x1 = 60) -->
<!-- # Print results. -->
<!-- public_firm_bs -->
<!-- ``` -->

<!-- In $t=2$ more investors realize that the firm is well managed, its market is growing and there is no close competition. Now, investors actually think that paying 1 USD per stock is actually a bargain, so they are willing to pay even more than 1 USD to participate in this project. So, the demand for those stocks increases, and as a result the price goes from 1 to 2 USD per share. By the end of $t=2$ we have: -->

<!-- ```{r} -->
<!-- # Stock price increased from 1 to 2. -->
<!-- public_firm_bs <- c(assets = 100+60, liabilities = 40,  -->
<!--                     equity_60x2 = 120) -->
<!-- # Print results. -->
<!-- public_firm_bs -->
<!-- ``` -->

<!-- Now the firm has grown by going public and by showing good perspectives. -->

<!-- However, in $t=3$ we sadly get to know that the firm's financial statements were overestimating profits, and a big fraud was detected today. Big scandal and all newspapers talk about it. Then, investors are now interested in selling all their stocks as quickly as they can. Investors are so desperate to sell their stocks that they drop the stock price to almost zero. By the end of $t=3$ we have: -->

<!-- ```{r} -->
<!-- # Stock price decrease to 0. -->
<!-- public_firm_bs <- c(assets = 40, liabilities = 40,  -->
<!--                     equity_60x0 = 0) -->
<!-- # Print results. -->
<!-- public_firm_bs -->
<!-- ``` -->

<!-- There are still 60 stocks in the market with a unit price of zero, so the market value of equity is null. This hypothetical firm is virtually bankrupt as they need to sell all their assets to pay the debt. In simple terms, a firm is bankrupt when the liabilities are greater than the assets. The firm will hardly get fresh funds as investors realize this was a bad project. -->

<!-- Many things can happen when a firm faces this difficult situation. Sometimes it is worthwhile to try to improve the firm's financial condition, hiring professional managers, and hope to recover the investor's confidence. Before bankruptcy, firms could also negotiate with their debt holders. If negotiation succeeds, the firm could get a longer debt maturity, a reduced debt to help the firm remain alive and hope to recover its operations, or a more innovative solution. There are also other firms that look for firms in trouble to buy them for a wide variety of reasons. In the US, you may hear about "Chapter 11", which is a form of bankruptcy that involves a reorganization of a debtor's business affairs, debts, and assets, and for that reason is known as "reorganization" bankruptcy. Corporations generally file Chapter 11 if they require time to restructure their debts. This version of bankruptcy gives the debtor a fresh start. However, the terms are subject to the debtor's fulfillment of its obligations under the plan of reorganization. -->

<!-- As we show before, the stock price evolution is generally a good indicator about how the investors interpret the future of the firm. When the stock price increases, then investors are demanding more stocks. A higher stock price is associated with a more valuable equity and this is good for the firm. When the stock price decreases, then there are more people trying to sell stocks compared to those trying to buy so the stock price falls. There might be some cases in which a drop in the stock price is associated with a firm good performance. Consider the stock price has increased in the last week, investors bought at 10 USD and today's price is 15 USD. Those investors might be interested to sell now to make a 5 USD profit. This increase in the supply of stocks could be significant enough to drop prices in the short run. -->

<!-- Remember this firm needed to go public before issuing and selling stocks in the stock market. Normally, these firms are big, although there are some stock markets for medium-sized firms. Small firms can hardly access these kinds of markets to get funding in a direct way. This might be problematic because small firms are the ones which require more alternatives to get financing so they can grow. The literature regarding small firms show that one of the most important problems they face is precisely the lack of financing or the limited alternatives to get funds. This remains an open challenge for the financial services and an opportunity for fintech companies. In the world there are significantly more small firms compared with big firms, so any improvement in the credit conditions of small firms could have a significant impact on the income of many people. -->

<!-- In the following section we move from a financial statement analysis to the analysis of stock prices. Note that both are linked by the equity and overall performance of public firms. -->

## Stock prices and other financial data

Financial markets facilitate transactions between buyers and sellers and generate a rich source of financial data. The examples below show how R can download and organize stock price data for analysis. The goal at this stage is modest and important: create clean objects whose dates, variables, units, and source can be checked before any return or model is calculated.

Packages such as <tt>`quantmod`</tt> and <tt>`tidyquant`</tt> reduce the friction of downloading market data. <tt>`quantmod`</tt> provides classic tools for quantitative financial analysis, while <tt>`tidyquant`</tt> connects financial data retrieval with tidy data structures. The chapter uses both packages as data-access tools, leaving model evaluation and strategy testing for later chapters.

The following optional resource shows the mechanics of installing packages in R.

```{r}
embed_url("https://youtu.be/JBcVi-fAT_k")
```

In the past, financial analysis often required visiting a market-data site, downloading files in Excel or text format, and converting them into a compatible structure. Earlier workflows relied on printed sources. R packages now make this process faster, reproducible, and easier to document.

The first market-data example downloads Apple stock prices in one step with the <tt>`tq_get()`</tt> function from <tt>`tidyquant`</tt>.

```{r}
# Get stock prices for Apple stock from Yahoo! finance site.
aapl_stock_prices <- tq_get("AAPL")
```

The object <tt>`aapl_stock_prices`</tt> now contains Apple prices and related market information. The call to <tt>`tq_get()`</tt> comes from <tt>`tidyquant`</tt>, connects to the selected data provider, retrieves the "AAPL" symbol, and stores the result in a reproducible R object.

The structure of the object shows the available variables and their data types.

```{r}
# str function is a way to display the structure of an R object.
str(aapl_stock_prices)
```

According to the <tt>`str()`</tt> output, <tt>`aapl_stock_prices`</tt> contains daily observations and 7 variables. The object includes prices and volume. A compact inspection of the first and last rows helps verify the date range and the available columns.

The first rows show how the time series begins.

```{r}
# See the first observations of aapl_stock_prices.
head(aapl_stock_prices)
```

The last rows show the most recent observations retrieved by the data provider.

```{r}
tail(aapl_stock_prices)
```

By default, the <tt>`tq_get()`</tt> function downloads the latest set of data available. The last date of <tt>`aapl_stock_prices`</tt> should approximately correspond to the date when the code is executed. Differences can occur when the market is closed, when the code is run during a weekend, or when the data provider updates with a delay.

The same data can be inspected graphically. A price plot is usually the first diagnostic check: it reveals the date span, large splits or jumps, missing segments, and long-run price scale.

```{r}
#| label: fig-asp
#| fig-cap: "Apple stock prices."
plot(aapl_stock_prices$date, aapl_stock_prices$adjusted,
     type = "l")
```

Adding labels, a title, and a line color makes the plot easier to read.

```{r}
#| label: fig-aspif
#| fig-cap: "Apple stock prices, improved format."
plot(aapl_stock_prices$date, aapl_stock_prices$adjusted,
     type = "l", xlab = "Date", ylab = "Adjusted price",
     main = "Apple stock price", col = "blue")
```

Apple can also be compared with a broad market index such as the S&P 500. The comparison is useful because individual stock prices are often interpreted relative to a broad market benchmark.

```{r}
SP <- tq_get("^GSPC")
```

The two series are plotted together with separate vertical axes because Apple and the S&P 500 are measured on different price scales. Later chapters use indexed prices and returns to make comparisons more systematic.

```{r}
#| label: fig-aspas
#| fig-cap: "Apple stock prices and SP500."
par(mar = c(5, 5, 2, 5))
plot(SP$date, SP$adjusted, type = "l", col = "red",
     ylab = "Standard and Poors 500 index",
     xlab = "Date")
par(new = T)
plot(aapl_stock_prices$date, aapl_stock_prices$adjusted,
     type = "l", axes = F, xlab = NA, ylab = NA, cex = 1.2)
axis(side = 4)
mtext(side = 4, line = 3, "Apple stock price")
legend("topleft",
       legend=c("SP500", "Apple"),
       lty = 1, col = c("red", "black"))
```

The Apple example can also be used to study intraday variation. Stock prices change throughout the trading day, so the open, high, low, and close prices summarize different pieces of the trading session. For day $t$, the intraday range is

$$
\Delta_t = \text{High}_t - \text{Low}_t.
$$

The code creates <tt>`aapl_diff`</tt>, the daily high-low range.

```{r}
aapl_diff = aapl_stock_prices$high - aapl_stock_prices$low
```

The daily range can be plotted directly.

```{r}
#| label: fig-astv
#| fig-cap: "Apple daily high-low range."
plot(aapl_stock_prices$date, aapl_diff, type = "h", xlab = "Date",
     ylab = "Difference between high and low prices", col = "blue")
```

The difference between high and low prices shows that the stock price changes during a trading day. Larger bars identify days with wider intraday movement. The next calculation locates the date with the largest high-low range without sorting the full data set.

Mathematically, the task is to find the observation index where the daily range is largest.

$$
t^{\star}=\operatorname*{arg\,max}_{t}\Delta_t.
$$

```{r}
highest_change = which.max(aapl_diff)
highest_change
```

The value of <tt>`highest_change`</tt> is the row index containing the highest <tt>`aapl_diff`</tt> value. This index extracts the corresponding date and range value.

The date and the actual value of <tt>`aapl_diff`</tt> are extracted from the same row index.

```{r}
# The brackets [] are used to extract a single row value.
when = aapl_stock_prices$date[highest_change]
when
```


```{r}
top = aapl_diff[highest_change]
top
```

The output above identifies the date <tt>`when`</tt> Apple had its largest daily high-low range <tt>`top`</tt>. These values can be added to the plot.

```{r}
#| label: fig-astvafa
#| fig-cap: "Apple daily high-low range and maximum observation."
plot(aapl_stock_prices$date, aapl_diff, type = "h", xlab = "Date",
     ylab = "Difference between high and low prices", col = "blue")
# Here, we add the red point.
points(when, top, pch = 19, col = "red", cex = 1.5)
```

So far, the examples have used default download options. A call such as <tt>`tq_get("AAPL")`</tt> returns the default historical range available through the data source. The next example changes the start date to request a longer Apple history, following examples developed by Matt Dancho, the author of <tt>`tidyquant`</tt>.

The call below requests Apple stock prices starting in 1990.

```{r}
aapl_prices  <- tq_get("AAPL", get = "stock.prices",
                       from = " 1990-01-01")
# Show results.
aapl_prices
```

Sometimes the analysis requires aggregation from daily to monthly frequency. Aggregating from a higher-frequency series to a lower-frequency series is straightforward; the reverse direction would require additional assumptions. FANG is a dataset containing the daily historical stock prices for the FANG tech stocks, META, AMZN, NFLX, and GOOG, spanning from the beginning of 2013 through the end of 2016.

For month $m$, the monthly adjusted price used here is the last available adjusted price in that month:

$$
P_m^{monthly}=P_{\max(t \in m)}^{adjusted}.
$$

The pre-loaded data are then aggregated from daily to monthly frequency.

```{r}
data("FANG")
FANG |>
    group_by(symbol) |>
    tq_transmute(select = adjusted, mutate_fun = to.monthly,
                 indexAt = "lastof")
```

The <tt>`tidyquant`</tt> package can also access other kinds of data from sources such as Federal Reserve Economic Data (FRED). FRED is maintained by the Research division of the Federal Reserve Bank of St. Louis and contains US and international time series. The next example uses WTI crude oil prices to show that economic data can contain unusual market episodes.

The call below downloads WTI oil prices from FRED. For reproducibility, the code also includes a compact stored FRED excerpt around the April 2020 episode, which keeps the example executable when the data provider is temporarily unavailable during rendering.

```{r}
# See https://fred.stlouisfed.org/series/DCOILWTICO
wti_price_usd <- tryCatch({
  wti_price_raw <- quantmod::getSymbols("DCOILWTICO",
                                        src = "FRED",
                                        auto.assign = FALSE)

  data.frame(
    date = as.Date(rownames(as.data.frame(wti_price_raw))),
    price = as.numeric(wti_price_raw[, 1])
  )
}, error = function(e) {
  # Stored excerpt from FRED/ALFRED DCOILWTICO around April 2020.
  read.csv(text = "date,price
2020-02-21,53.36
2020-02-24,51.36
2020-02-25,49.78
2020-02-26,48.67
2020-02-27,47.17
2020-02-28,44.83
2020-03-02,46.78
2020-03-03,47.27
2020-03-04,46.78
2020-03-05,45.90
2020-03-06,41.14
2020-03-09,31.05
2020-03-10,34.47
2020-03-11,33.13
2020-03-12,31.56
2020-03-13,31.72
2020-03-16,28.96
2020-03-17,26.96
2020-03-18,20.48
2020-03-19,25.09
2020-03-20,19.48
2020-03-23,23.33
2020-03-24,21.03
2020-03-25,20.75
2020-03-26,16.60
2020-03-27,15.48
2020-03-30,14.10
2020-03-31,20.51
2020-04-01,20.28
2020-04-02,25.18
2020-04-03,28.36
2020-04-06,26.21
2020-04-07,23.54
2020-04-08,24.97
2020-04-09,22.90
2020-04-13,22.36
2020-04-14,20.15
2020-04-15,19.96
2020-04-16,19.82
2020-04-17,18.31
2020-04-20,-36.98
2020-04-21,8.91
2020-04-22,13.64
2020-04-23,15.06
2020-04-24,15.99
2020-04-27,12.17
2020-04-28,12.40
2020-04-29,15.04
2020-04-30,19.23
2020-05-01,19.72
2020-05-04,20.47
2020-05-05,24.56
2020-05-06,23.88
2020-05-07,23.68
2020-05-08,24.73
2020-05-11,24.02
2020-05-12,25.76
2020-05-13,25.37
2020-05-14,27.40
2020-05-15,29.44
2020-05-18,31.83")
})

wti_price_usd$date <- as.Date(wti_price_usd$date)
wti_price_usd <- wti_price_usd[complete.cases(wti_price_usd), ]
# Show results.
wti_price_usd
```

Oil prices can be plotted directly after the data are retrieved. The series includes the unusual episode of negative oil prices, a rare event that can occur in commodity markets under extreme storage and delivery pressure.

The line plot shows the evolution of oil prices through time.

```{r}
#| label: fig-op
#| fig-cap: "Oil prices."
ggplot(wti_price_usd, aes(date, price)) +
  geom_line() + theme_bw() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red")
```

The exact day with a negative oil price is found by filtering observations that satisfy:

$$
Price_t < 0.
$$

```{r}
subset(wti_price_usd, price < 0)
```

Without going into technical arguments, negative prices can be interpreted through storage pressure and contract delivery. Imagine, as it actually happened, that storing oil is expensive and producers have no further physical space to store production. At the same time, buyers see weak economic conditions and expect lower demand for fuel. Producers want to sell, while many buyers already have enough inventory or lack storage capacity. Under extreme pressure, producers may be willing to pay others to take oil delivery. This is why commodity prices can be negative.

There are other explanations, for example one related to the maturity of oil futures contracts. The price that went negative on Monday 2020-04-20 was for futures contracts to be delivered in May. Those contracts expired on Tuesday 2020-04-21. Upon expiration of the futures contract, the clearinghouse matches the holder of a long contract against the holder of a short position. The short position delivers the underlying asset to the long position. So, on Monday, traders --- who were not equipped to take physical deliveries --- were rushing to sell them to buyers who have booked storage.

Exchange rates are another core data type in financial modeling because foreign assets must often be converted into a common currency before returns, portfolio values, or risk measures are computed. The <tt>`quantmod`</tt> package can retrieve exchange rates as well.

```{r}
#| label: fig-uer
#| fig-cap: "USD/MXN exchange rate."
# Download USD/MXN exchange rate from Oanda site.
exchange_rate <- getSymbols("USD/MXN", src = "oanda",
                            auto.assign = FALSE)
# Plot the results.
plot(exchange_rate)
```

The same exchange-rate series can be summarized with a density plot. The density shows where the exchange rate has spent more time historically and gives a compact view of the empirical distribution.

```{r}
#| label: fig-ueradp
#| fig-cap: "USD/MXN exchange rate, a density plot."
# Density plots.
ggplot(exchange_rate, aes(x = USD.MXN)) +
  geom_density(alpha = 0.4, fill = "red") +
  geom_hline(yintercept = 0) +
  labs(title = "USD.MXN", x = "USD.MXN", y = "Density") +
  theme(legend.position = "none", legend.title = element_blank()) +
  scale_fill_tq()
```

Macroeconomic series provide context for financial analysis because asset prices react to economic conditions, policy expectations, and business-cycle risk. The next example visualizes US employment together with NBER recession periods shown as shaded areas. Employment growth is computed as a three-period percentage change:

$$
g_t^{(3)} = 100\left(\frac{E_t}{E_{t-3}}-1\right).
$$

The code below prepares the recession shading and the employment series.

```{r}
# US nonfarm payroll employment
df <- tq_get("PAYEMS", get = "economic.data", from = "1948-01-01")
# recession df (for plotting)
recessions.df = read.table(textConnection(
  "Peak, Trough
  1945-02-01, 1945-10-01
  1948-11-01, 1949-10-01
  1953-07-01, 1954-05-01
  1957-08-01, 1958-04-01
  1960-04-01, 1961-02-01
  1969-12-01, 1970-11-01
  1973-11-01, 1975-03-01
  1980-01-01, 1980-07-01
  1981-07-01, 1982-11-01
  1990-07-01, 1991-03-01
  2001-03-01, 2001-11-01
  2007-12-01, 2009-06-01
  2020-02-01, 2021-04-30"), sep = ',',
  colClasses = c('Date', 'Date'), header = TRUE)
rec3 <- filter(df, df$symbol == "PAYEMS")
my_trans <- function(in.data,transform = "pctdiff3") {
  switch(transform, logdiff  = c(NA, diff(log(in.data))),
         pctdiff3 = 100 * Delt(in.data, k = 3),
        logdiff3 = c(rep(NA, 3), diff(log(in.data), 3)))
}

df41 <- df |>
  transmute(date, PAYEMS = my_trans(price)) |>
  filter(year(date) > 1945)
```

With the data prepared, employment growth can be plotted together with recession periods.

```{r}
ggplot(data = df41, aes(x = date, y = PAYEMS)) +
  geom_rect(data = recessions.df, inherit.aes = FALSE,
            aes(xmin = Peak, xmax = Trough,
                ymin = -Inf, ymax = +Inf),
            fill = 'black', alpha = 0.3) +
  theme_minimal() +
  geom_line(color = "red", size = 1, alpha = 0.8) +
  labs(x = "", y = "",
       title = "US employment growth and recessions",
       subtitle = "Shaded areas are NBER recessions",
       caption = "Source: US Bureau of Labor Statistics.
       \nRetrieved from FRED, Federal Reserve Bank of St. Louis.") +
  geom_hline(yintercept = 0) +
  theme(plot.caption = element_text(hjust = 0),
        plot.subtitle = element_text(face = "italic", size = 9),
        plot.title = element_text(face = "bold", size = 14))
```

The shaded areas identify NBER recession periods, and the red line shows employment growth. The plot makes the economic-data workflow useful: a time series becomes more informative when it is aligned with an external event indicator. Recessions are associated with visible pressure on employment growth, although the magnitude and timing differ across episodes.

Access to financial and economic data is only the first step. The analyst also has to align dates, check definitions, transform units, and communicate the result in a way that supports a financial or economic question. The same discipline will be used in later chapters when prices become returns, returns become signals, and portfolio returns become risk measures.

## Technical analysis

Technical analysis studies price and volume patterns in an attempt to describe trend, momentum, and possible turning points. It can be applied to stocks, indexes, commodities, futures, and other traded instruments. The approach is useful in this book for a specific reason: it converts raw prices into indicators, and those indicators can later become inputs for explicit trading rules. Its limitations should remain visible. A chart pattern does not control for fundamentals, macroeconomic shocks, liquidity, transaction costs, or model validation.

The chapter now returns to stock prices and builds a few basic indicators.

```{r}
# Get AAPL and AMZN stock prices.
AAPL <- tq_get("AAPL", get = "stock.prices", from = "2015-09-01",
               to = "2016-12-31")
```

The first plot shows Apple closing prices. This line chart is the baseline against which the later indicators will be compared.

```{r}
#| label: fig-alc
#| fig-cap: "Apple line chart."
AAPL |>
    ggplot(aes(x = date, y = close)) + geom_line() +
    labs(y = "Closing price", x = "") +
    theme_tq()
```

The previous plot represents closing prices. The data also contain open, high, low, and close prices for each day. A candlestick chart uses that information in a compact format: the vertical range shows the daily high and low, while the body summarizes the open-close movement. Blue candles indicate days when the close is above the open, and red candles indicate days when the close is below the open.

```{r}
#| label: fig-acc
#| fig-cap: "Apple candlestick chart."
AAPL |>
  ggplot(aes(x = date, y = close)) +
  geom_candlestick(aes(open = open, high = high, low = low,
                       close = close)) +
  labs(y = "Closing Price", x = "") +
  theme_tq()
```

There are some blank spaces. This is simply because the close price of yesterday is not always exactly the same as the open price of today. Local stock markets close on weekends and holidays.

These charts are difficult to read when there are many observations. A zoomed view helps inspect the recent pattern.

```{r}
#| label: fig-accz
#| fig-cap: "Apple candlestick chart, zoom."
AAPL |>
  ggplot(aes(x = date, y = close)) +
  geom_candlestick(aes(open = open, high = high, low = low,
                       close = close)) +
  coord_x_date(xlim = c("2016-12-15", "2016-12-31"),
               ylim = c(28.75, 29.5)) +
  labs(y = "Closing Price", x = "") +
  theme_tq()
```

Candlesticks are descriptive signals. A sequence of large blue candles followed by smaller blue candles may suggest fading upward momentum, while a sequence of large red candles followed by smaller red candles may suggest fading downward pressure.[^4] These readings are useful for organizing visual evidence, but they require formal validation before becoming a trading rule.

[^4]: The analogy is similar to *flatten the curve* in the context of the 2020 pandemic. As the curve increases, the high speed of increase is captured by big vertical blue bars. A slowdown appears as smaller vertical blue bars before the curve becomes flat and eventually decreases.

A short sequence of alternating blue and red bars is difficult to convert into a reliable forecast. This indicator alone, without complementary analysis, does not deliver a clear trading signal.

Moving averages are often used as trend-following indicators. The black line is the original price series, and the blue line is the moving average. In this case, the indicator is a 30-day simple moving average, calculated as the average of the last 30 closing prices:

$$
SMA_{t,n}=\frac{1}{n}\sum_{j=0}^{n-1}P_{t-j}.
$$

In the code below, this value is easily modified by changing the value of $n$, in this case $n=30$. A crossover can be read as a possible change in momentum, but it is a heuristic signal and can generate false positives.

If the black line crosses the blue line from below, the chart suggests a possible increase in momentum. If it crosses from above, the chart suggests weakening momentum. In practice, the value of $n$ should be calibrated and validated with historical data before using the indicator as part of a trading rule.

```{r}
#| label: fig-a3dmac
#| fig-cap: "Apple 30-day moving average chart."
AAPL |>
  ggplot(aes(x = date, y = close)) +
  geom_line()  +
  geom_ma(ma_fun = SMA, n = 30, linetype = 1, size = 1.25) +
  labs(y = "Closing Price", x = "") +
  theme_tq()
```

In this historical window, some price declines occurred after the price crossed below the moving average, and some increases occurred after the reverse pattern. This is an ex-post reading of the chart. It does not establish forecasting power by itself. The period around July 2016 is less clear. By the end of the time-series, the black line is above the blue, which suggests recent upward momentum.

```{r}
AAPL_recent <- tq_get("AAPL", get = "stock.prices",
                      from = "2016-12-30", to = "2017-01-12")
AAPL_recent
```

Bollinger Bands are envelopes plotted at a standard deviation level above and below a simple moving average of the price. Because the distance of the bands is based on standard deviation, they adjust to volatility swings in the underlying price. With $k=2$ and a rolling standard deviation $s_{t,n}$, the bands are:

$$
Upper_{t,n}=SMA_{t,n}+k s_{t,n},
\qquad
Lower_{t,n}=SMA_{t,n}-k s_{t,n}.
$$

Wider bands indicate a wider recent volatility envelope. Narrower bands indicate a quieter recent price range. The indicator describes the size of recent price fluctuations and should be tested before being interpreted as a forecast.

```{r}
#| label: fig-abbasma
#| fig-cap: "Apple Bollinger bands and simple moving average."
AAPL |>
  ggplot(aes(x = date, y = close, open = open,
             high = high, low = low, close = close)) +
  geom_line() +
  geom_bbands(ma_fun = SMA, sd = 2, n = 30,
              linetype = 2, size = 0.5, alpha = 0.2,
              fill        = palette_light()[[1]],
              color_bands = palette_light()[[1]],
              color_ma    = palette_light()[[2]]) +
  labs(y = "Closing Price", x = "") +
  theme_tq()
```

The last observation has the price line above the moving average, now shown as a red dotted line. The last price sits above the moving average and near the upper band. This suggests recent upward momentum and a price near the upper part of its rolling volatility envelope. The bands are descriptive; using them as a trading rule would require formal backtesting.

The lesson is methodological. Technical indicators can be useful descriptive tools, but a visual reading is fragile when it depends on one or two signals by eye. A stronger workflow defines the signal, fixes the timing, evaluates the rule on historical data, and accounts for trading costs. Packages such as <tt>`quantstrat`</tt> provide infrastructure for formal backtesting of signal-based strategies.

The simple moving average for Meta is computed in the same way.

```{r}
#| label: fig-f6dmac
#| fig-cap: "Meta 60-day moving average chart."
FANG |>
  filter(symbol == "META") |>
  ggplot(aes(x = date, y = close)) +
  geom_line()  +
  geom_ma(ma_fun = SMA, n = 60, linetype = 1, size = 1.25) +
  labs(y = "Closing Price", x = "") +
  theme_tq()
```

This produces a bearish-looking moving-average signal. The realized price path can be compared with the signal as an ex-post check.

```{r}
FANG |>
  filter(symbol == "META") |>
  filter(date >= "2014-01-01" & date <= "2014-02-01")
```

In this historical example, the realized price path moved in the direction suggested by the moving-average signal. The next chapter turns this idea into an explicit trading workflow: prices are transformed into indicators, indicators become predictors, and predictions are evaluated as trading positions.
