# Prices and returns

```{r include=FALSE}
rm(list = ls())
library(tidyquant)
library(ggplot2)
library(dplyr)
library(lubridate)
```

An asset is any resource owned by a business or an economic entity that can produce value in the future. In this book, stocks are the main working example, but the same return logic can be applied to commodities, currencies, bonds, derivatives, and portfolio values.

This chapter is the bridge between raw financial data and the modeling chapters that follow. The previous chapter showed how to obtain and visualize prices. Here, we convert those prices into returns, cumulative values, distributional summaries, and risk-return indicators. These objects become the inputs for trading rules, asset pricing regressions, portfolio allocation, and Value at Risk.

A return is the percentage change of an asset price. Returns allow assets with different price levels to be compared on the same scale. For example, buying at \$100 and selling at \$110 produces a \$10 gain and a 10% return. Buying at \$10 and selling at \$12 produces a \$2 gain and a 20% return. The dollar gain is larger in the first case; the relative performance is larger in the second case.

The previous returns were calculated as a simple percentage change: $\frac{\$110-\$100}{\$100}$ and $\frac{\$12-\$10}{\$10}$ respectively. We can also calculate log-returns. They are close to simple returns for small price changes, and they are especially useful because multi-period log-returns add over time. For larger price changes, simple and log returns differ.

In this case the log-returns are $\log\left(\frac{\$110}{\$100}\right)$ and $\log\left(\frac{\$12}{\$10}\right)$ which are 9.53% and 18.23% respectively.

```{r}
log(110/100)
log(12/10)
```


Both absolute and relative valuations (\$10 versus \$2, and 10% versus 20%) are important and valid depending on what we want to report. However, the good thing about the relative approach (percentages) is that we can compare among different alternatives more easily. Returns also have some relevant statistical properties (like stationarity) that are needed to implement some quantitative financial and econometrics models.

Returns can also be expressed relative to risk. If two assets have the same expected return of 5%, but the first has risk of 5% and the second has risk of 10%, the first asset has 1 unit of return per unit of risk and the second has 0.5. This is the logic behind many risk-return comparisons in the rest of the book.

There are several ways to measure and report asset returns: daily, monthly, yearly, simple, log, cumulative, and compounded. The chapter introduces the versions needed for the later modeling workflow.

## From prices to returns

We need time series of asset prices to calculate time series of asset returns. The first plot shows daily adjusted prices in the <tt>`FANG`</tt> database.

```{r}
#| label: fig-dspnsetr
#| fig-cap: "Daily stock prices - not so easy to read."
FANG_daily_all <- FANG |>
  group_by(symbol) |>
  ggplot(aes(x = date, y = adjusted, color = symbol)) +
  geom_line(size = 1) +
  labs(x = "", y = "Adjusted prices", color = "") +
  scale_y_continuous(labels = scales::dollar) +
  theme_tq() +
  scale_color_tq()

FANG_daily_all
```

See how stock prices fluctuate in different ranges. It would be a mistake to prefer Amazon or Google simply because they have larger price values compared with Meta or Netflix. Taking investment decisions based on prices could be misleading because investors are most of the time looking for returns, and in particular for an attractive risk-return combination. The vertical axis is also problematic in the plot as it is difficult to see the price evolution clearly for each stock. In sum, we do not take investment decisions based on this kind of visual representation. If we are interested to evaluate the stock performance, we need a risk-return analysis, not a price chart.

A faceted view makes the individual price paths easier to inspect.

```{r}
#| label: fig-dspetr
#| fig-cap: "Daily Stock Prices - easier to read."
FANG_daily <- FANG_daily_all +
  facet_wrap(~ symbol, ncol = 2, scales = "free_y") +
  theme(legend.position = "none", legend.title = element_blank())

FANG_daily
```

Now it is clear that all stock prices have a positive trend over time. It is also clear that 2014 was a bad year for Amazon, Google and even Netflix. This is because the stock price decreased in the period of 2014-2015. The year 2015 was good for Amazon and Google. The main point is that this new visualization of the data can help us to compare the evolution of price per stock. Remember we still have prices and not returns. Stock returns are easily calculated in R.

Here, we first select adjusted prices, then transform into yearly returns. For an asset with adjusted price $P_t$, the simple return over one period is:

$$
R_t = \frac{P_t}{P_{t-1}} - 1.
$$

The code below applies the same expression after grouping the data by stock and changing the periodicity to yearly.

```{r}
# Create yearly returns.
FANG_annual_returns <- FANG |>
    group_by(symbol) |>
    tq_transmute(select     = adjusted,
                 mutate_fun = periodReturn,
                 period     = "yearly",
                 type       = "arithmetic")
FANG_annual_returns
```

This is what we call tidy data because (1) each variable forms a column; (2) each observation forms a row; and (3) each type of observational unit forms a table. These yearly returns are expressed in decimal notation. This means that 0.95 is in reality 95%.

The next chunk of code shows how to visualize the annual returns in a nice plot.

```{r}
#| label: fig-far
#| fig-cap: "FANG: annual returns."
FANG_annual_returns |>
  ggplot(aes(x = date-365, y = yearly.returns, fill = symbol)) +
  geom_col() +
  geom_hline(yintercept = 0, color = "black") +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Annual returns", x = "") +
  facet_wrap(~ symbol, ncol = 2, scales = "free_y") +
  theme_tq() +
  scale_fill_tq() +
  theme(legend.position = "none", legend.title = element_blank())
```

Risky stocks like these are risky because the returns change drastically from time to time. Note that here, it is easy to see that 2014 was indeed a bad year for Amazon, Google and Netflix as we anticipate when looking at the price chart. These three assets confirm what we anticipate before because we see negative returns in 2014. The year 2015 was good for Amazon and Google as we said before.

The question that arises now is: What would be the result of investing money in each stock during these years (starting the first day of 2013 and finishing the last day of 2016)? Yearly returns show what happens from one year to another, but we cannot tell which stock would represent the best investment strategy in these years looking at this past information. Cumulative returns can be useful to answer this question.

## Cumulative returns

Working with cumulative percentage changes requires multiplying growth factors. If an initial value is \$100 and it increases by 5%, the new value is $\$100 \times (1+0.05) = \$105$. If the value later falls by 20%, the current value is $\$100 \times (1+0.05) \times (1-0.2) = \$84$.

Now consider a \$100 investment in Amazon over this period. The value of the investment is calculated by compounding the four yearly returns listed above. Amazon's yearly returns from 2013 to 2016 are: 54.98%, -22.18%, 117.78%, 10.95% in that order.

```{r}
# Investment in USD.
investment = 100
# Amazon's cumulative return in percentage.
AMZN_cum_percentage = (1 + 0.54984265) * (1 - 0.22177086) *
  (1 + 1.17783149) * (1 + 0.10945565) - 1
# Amazon's cumulative return in USD.
AMZN_cum_usd = investment * (1 + AMZN_cum_percentage)
# Show results.
AMZN_cum_percentage
AMZN_cum_usd
```

The dollar value of \$100 invested in Amazon at the end of the 4 years is \$291.4267. The cumulative return is 191.4267%. For the sake of clarity, we can check these values are correct by applying this return over the \$100: $\$100 \times (1+1.914267) = \$291.4267$.

The 191.4267% cumulative return is usually not reported as it is here because it accumulates 4 years of returns. We normally express mean returns, in this case mean yearly return. The arithmetic mean of the returns, calculated by taking the sum of the returns and dividing by 4 is:

```{r}
# Arithmetic mean return.
AMZN_arith_mean_retun = (0.54984265 - 0.22177086 +
                         1.17783149 + 0.10945565) / 4
AMZN_arith_mean_retun
```

The arithmetic mean can be misleading when it is used as if it were a compounded growth rate. In this example, the arithmetic mean return is 40.38397% per year. Applying that value as a constant annual growth rate gives:

The long way.

```{r}
investment * (1 + AMZN_arith_mean_retun) *
  (1 + AMZN_arith_mean_retun) *
  (1 + AMZN_arith_mean_retun) * (1 + AMZN_arith_mean_retun)
```

The short way.
```{r}
investment * (1 + AMZN_arith_mean_retun) ^ 4
```

The result is inconsistent with the actual compounded value. The dollar value of \$100 invested in Amazon at the end of the 4 years is \$291.4267, while applying the arithmetic mean as a growth rate gives \$388.3919. The arithmetic mean is useful as a sample summary, but it should be distinguished from the rate that reproduces the multi-period investment value.

We reconcile this difference with the geometric mean return. A geometric mean is the $n$-th root of the product of $n$ numbers. In particular:

```{r}
# Geometric mean return.
AMZN_geom_mean_retun = ((1 + 0.54984265) * (1-0.22177086) *
                        (1 + 1.17783149) * (1 + 0.10945565)) ^ (1/4) - 1
AMZN_geom_mean_retun
```

The geometric mean return of 30.65689% per year reproduces the compounded investment value: \$100 grows to \$291.4267 by the end of the fourth year. The next chunk verifies the calculation.

The long way.

```{r}
investment * (1 + AMZN_geom_mean_retun) *
  (1 + AMZN_geom_mean_retun) *
  (1 + AMZN_geom_mean_retun) * (1 + AMZN_geom_mean_retun)
```

The short way.

```{r}
investment * (1 + AMZN_geom_mean_retun) ^ 4
```

This phenomenon is an example of a result that is well known in mathematics. For positive gross returns, the geometric mean is less than or equal to the arithmetic mean (30.6 \< 40.3). Although our arithmetic mean return of 40.38397% can lead to an overestimation of the investment dollar returns, we usually report mean returns. This is a common issue when reporting mutual funds returns as it is tempting to report arithmetic mean returns because they are higher. This is why in some jurisdictions, regulations require fund managers to report geometric mean returns, 30.65689% in this case.

For now, we work with arithmetic returns. The next plot shows how a \$100 investment grows each year for each stock. The plotted values are in USD.

```{r}
#| label: fig-ycurafv
#| fig-cap: "Yearly cumulative USD returns, a facet view."
# Yearly returns.
FANG_annual_returns <- FANG |>
    group_by(symbol) |>
    tq_transmute(select     = adjusted,
                 mutate_fun = periodReturn,
                 period     = "yearly",
                 type       = "arithmetic")
# Cumulative returns.
FANG_annual_cum_returns <- FANG_annual_returns |>
  mutate(cr = 100*cumprod(1 + yearly.returns)) |>
# Plot the results.
  ggplot(aes(x = date-365, y = cr, fill = symbol)) + geom_col() +
  labs(subtitle = "100 USD investment growth from 2013-01-01 to 2016-12-31.",
  x = "", y = "USD value at the end of the year", color = "") +
  scale_y_continuous(labels = scales::dollar) +
  facet_wrap(~ symbol, ncol = 2, scales = "free_y") +
  theme_tq() +
  scale_color_tq() +
  theme(legend.position = "none", legend.title = element_blank())

FANG_annual_cum_returns
```

Please verify that the dollar value of \$100 invested in Amazon at the end of the 4 years is \$291.4267, as we discussed before.

We can also modify the code above to show monthly returns instead of yearly.

```{r}
#| label: fig-fmrafv
#| fig-cap: "FANG: monthly returns, a facet view."
# Monthly returns.
FANG_monthly_returns <- FANG |>
    group_by(symbol) |>
    tq_transmute(select     = adjusted,
                 mutate_fun = periodReturn,
                 period     = "monthly",
                 type       = "arithmetic")
# Plot the results.
ggplot(FANG_monthly_returns, aes(x = date-12,
       y = monthly.returns, fill = symbol)) +
  geom_col() +
  geom_hline(yintercept = 0, color = "black") +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Monthly returns", x = "") +
  facet_wrap(~ symbol, ncol = 2, scales = "free_y") +
  theme_tq() +
  scale_fill_tq() +
  theme(legend.position = "none", legend.title = element_blank())
```

Moving from yearly to monthly returns gives 12 observations per year. Note that this new plot reveals how volatile these stock returns are. There is a negative return just the month after GOOG reached a positive 20% return. It looks hard to anticipate what will happen the next month since there is evidence that these returns change from positive to negative quite frequently.

Monthly returns allow a finer view of the same \$100 investment example. If the initial investment is $V_0=100$, the investment value after $t$ monthly returns is:

$$
V_t = 100 \prod_{j=1}^{t}(1+R_j).
$$

The code inserts the initial value $V_0=100$ for every stock and then appends the cumulative monthly values. This makes the comparison start from the same dollar base for all four assets.

```{r}
#| label: fig-mcur
#| fig-cap: "Monthly cumulative USD returns."
# Calculate monthly cumulative returns.
FANG_monthly_cum_returns <- FANG_monthly_returns |>
  group_by(symbol) |>
  mutate(cr = 100 * cumprod(1 + monthly.returns))
FANG_monthly_cum_returns <- bind_rows(
  FANG_monthly_returns |>
    distinct(symbol) |>
    mutate(date = as.Date("2012-12-31"),
           monthly.returns = 0,
           cr = 100),
  FANG_monthly_cum_returns
) |>
  arrange(symbol, date)
# Plot results.
FANG_monthly_all <- FANG_monthly_cum_returns |>
ggplot(aes(x = date, y = cr, color = symbol)) +
  geom_line(size = 1) +
  geom_point(data = FANG_monthly_cum_returns |>
               filter(date == as.Date("2012-12-31")),
             size = 4, col = "black", alpha = 0.5) +
  labs(subtitle = "100 USD investment growth from 2013-01-01 to 2016-12-31.",
         x = "", y = "USD value at the end of the month", color = "") +
  scale_y_continuous(labels = scales::dollar) +
  theme_tq() +
  scale_color_tq()

FANG_monthly_all
```

Each stock now begins at the same $100 base value. The plot shows that a common starting investment can lead to very different terminal values, even when the original price series have different price levels.

Alternatively, we can split the plot in four panels. This facilitates the reading of the plot.


```{r}
#| label: fig-mcurafv
#| fig-cap: "Monthly cumulative USD returns, a facet view."
FANG_monthly <- FANG_monthly_all +
  facet_wrap(~ symbol, ncol = 2, scales = "free_y") +
  theme(legend.position = "none", legend.title = element_blank())

FANG_monthly
```

The dollar value of \$100 invested in Amazon at the end of the 4 years is again \$291.4267. The monthly view shows the path of the investment value through time. The faceted view confirms the same message with cleaner scales: every panel starts at $100, and Netflix reaches the largest terminal value in this historical window.

Remember this plot?

```{r}
#| label: fig-dspnsetra
#| fig-cap: "Daily stock prices - not so easy to read (again)."
FANG_daily_all
```

Netflix stock price remains down all the time. However, \$100 invested in Netflix during these 4 year would lead to the highest value of almost \$1,000. This is why we should conduct a return analysis to make investment decisions.

Plotting prices and returns over time is revealing because we can see the evolution of the stock performance over time. It is useful because we could analyze what happens on specific dates to better understand the behaviour of stocks. Remember 2014 was a bad year for Amazon and Google, it would be interesting to investigate what went wrong then so we can better understand the determinants of the firm's value.

## Distribution of returns

Although time series plots represent a useful tool for financial analysis, it is not the only way we can show stock returns. Density plots forget about time and show the distribution of values. The vertical axis is density. Areas under the curve represent probability mass, so taller regions indicate where returns are more concentrated. This approach is useful when we are interested to know the most likely return that the stock can have, how unlikely is to expect determined return values, how risky a specific stock return is.

The next figure uses density plots to compare the distribution of monthly returns in the <tt>`FANG`</tt> data.

```{r}
#| label: fig-fmradv
#| fig-cap: "FANG: monthly returns, a density view."
ggplot(FANG_monthly_returns, aes(x = monthly.returns, fill = symbol)) +
  geom_density(alpha = 0.5) +
  geom_hline(yintercept = 0, color = palette_light()[[1]]) +
  labs(x = "Monthly returns", y = "Density") +
  xlim(-0.3, 0.9) +
  theme_tq() +
  scale_fill_tq()
```

Note that monthly returns around zero are frequent, whereas monthly returns above 50% are very infrequent. Apparently, this 50% value is red, so this should be the case of Meta. It is also noteworthy to mention that Netflix has the higher monthly return dispersion of all. If this is not clear enough, we can add one line to the code to make it clearer.

```{r}
#| label: fig-fmradafv
#| fig-cap: "FANG: monthly returns, a density and facet view."
# Density plots.
ggplot(FANG_monthly_returns, aes(x = monthly.returns, fill = symbol)) +
  geom_density(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "black") +
  labs(x = "Monthly returns", y = "Density") +
  xlim(-0.3, 0.9) +
  theme_tq() +
  scale_fill_tq() +
  facet_wrap(~ symbol, ncol = 2) +
  theme(legend.position = "none", legend.title = element_blank())
```
The faceted density plot separates the four stocks and makes dispersion easier to compare.

-   AMZN. Note the mode is around zero, this suggests that the most frequent return value is around zero. The majority of observations are within roughly -25% and 25%, and there are only a very few returns higher than 25%.
-   GOOG. As in the case of AMZN, the most frequent return value is around zero. Contrary to AMZN, there are no monthly returns higher than 25%. GOOG shows a less dispersed distribution so it is less risky than AMZN.
-   META. The mode is slightly above zero. The sample mean should be verified from summary statistics because the mode and the mean can differ. META exhibits a greater dispersion of returns as there are a few around 50%.
-   NFLX. This is the riskiest asset compared with the rest as the monthly returns have a wide dispersion.

Density distributions are useful to have an idea about the overall risk and return of assets as they summarize 48 observations (in this case 12 observations per year, $12 \times 4$). Some may argue 48 observations are not enough to conduct a financial analysis.

Here is another way to show the returns.

```{r}
#| label: fig-fmrdpy
#| fig-cap: "FANG: Monthly return distribution per year."
FANG_monthly_returns |>
  mutate(year = as.factor(year(date))) |>
  ggplot(aes(x = monthly.returns, fill = symbol)) +
  geom_density(alpha = 0.5) +
  geom_hline(yintercept = 0) +
  labs(x = "Monthly returns", y = "Density") + xlim(-0.3, 0.6) +
  theme_tq() +
  scale_fill_tq() +
  facet_wrap(~ year, ncol = 2)
```

This view shows that return dispersion changes across years. Netflix contributes several extreme monthly observations, whereas Google remains comparatively concentrated around zero in this sample.

We now use statistical indicators to complement the previous analysis.

For a sample of returns $R_1,\ldots,R_n$, the statistics used below are:

$$
\bar{R}=\frac{1}{n}\sum_{t=1}^{n}R_t,
\qquad
s_R=\sqrt{\frac{1}{n-1}\sum_{t=1}^{n}(R_t-\bar{R})^2},
\qquad
sr=\frac{\bar{R}}{s_R}.
$$

The interquartile range is $IQR=Q_{0.75}-Q_{0.25}$.

```{r}
# Calculate relevant statistics.
FANG_stats <- FANG_monthly_returns |>
summarise(mean = mean(monthly.returns), sd = sd(monthly.returns),
          sr = mean/sd, iqr = IQR(monthly.returns))
FANG_stats
```

The sample mean is an estimate of expected return of the stock. The stock with the lowest mean is GOOG and the highest is NFLX. This is consistent with our previous graphical analysis. GOOG expected return is very close to zero, whereas NFLX expected return is higher. The standard deviation <tt>`sd`</tt> is a measure of how disperse are the returns or in financial terms, how risky are the returns. The stock with less risk is GOOG and the stock with highest risk is NFLX. This is also consistent with our previous graphical analysis. <tt>`IQR`</tt> is the interquartile range which is the first quartile subtracted from the third quartile.

Comparing return and risk, or mean and standard deviation of stock returns can be troublesome. This is because the relationship between risk and return is not perfectly proportional.

The risk-return comparison can also be shown graphically.

```{r}
#| label: fig-fthtrthtr
#| fig-cap: "FANG: the higher the risk, the higher the return."
# Mean variance plot.
FANG_stats |>
  ggplot(aes(x = sd, y = mean, color = symbol)) +
  geom_point(size = 5) +
  geom_text(aes(label = paste0(round(sr, 3))),
            vjust = 2, color = "black", size = 3.5) +
  xlim(0.04, 0.18) +
  ylim(0.01, 0.06) +
  labs(subtitle = "Numerical values represent return per unit of risk.",
       x = "Risk", y = "Expected return") +
  theme_tq()
```

According to our results, Netflix is the stock with the highest return and risk. Does this mean Netflix has the highest return per unit of risk? This is easy to find out by dividing the mean by the standard deviation. This indicator is called <tt>`sr`</tt> in the table above. According to this, Netflix has the highest return per unit of risk compared with the rest of the stocks. This is clearly consistent with our previous analysis of cumulative returns.

The plot above measures returns in $y$-axis and risk in $x$-axes. In finance, we call this a mean-variance approach as mean is the measure of return and variance, or its square root the standard deviation, is the measure of risk. The mean-variance approach was a breakthrough in finance because we can clearly compare any kind of assets under the basis of risk and return.

Imagine stock 1 has a return of 5% and a risk of 4%, whereas stock 2 has a return of 6% and a risk of 8%. Although stock 2 has a higher return, this return does not compensate for the increase in the risk. In particular stock 1 has a 1.25 return per unit of risk (5/4), and stock 2 has a 0.75 return per unit of risk. In this case, we would prefer to invest our money in stock 1. Again, the relative values are useful to make decisions. Here the relative value is stock return per unit of risk.

Return distributions help separate lower-risk and higher-risk assets. The next example simulates two artificial return series with the same expected return of 0% and different volatility levels. In notation:

$$
R^{low}_t \sim N(0,0.01^2),
\qquad
R^{high}_t \sim N(0,0.02^2).
$$

```{r}
#| label: fig-ri
#| fig-cap: "Risk illustration."
# Create random assets.
set.seed(13)
less_risky <- data.frame(return = rnorm(10000, 0, 0.01))
more_risky <- data.frame(return = rnorm(10000, 0, 0.02))
# Name random assets.
less_risky$asset <- 'Less risky asset'
more_risky$asset <- 'Riskier asset'
assets <- rbind(less_risky, more_risky)
# Plot the assets.
ggplot(assets, aes(return, fill = asset)) +
  geom_density(alpha = 0.5, adjust = 3) +
  geom_hline(yintercept = 0) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Simulated return", y = "Density")
```

The riskier asset can take a wider range of returns. Most simulated observations for the less risky asset are closer to zero; the riskier asset has wider tails. In the code above we ask R to generate 10,000 random values following a normal distribution with mean zero, and standard deviations of 1% and 2% for the less risky and riskier asset respectively.

The next chapter uses this same return language in a trading workflow. We move from describing return distributions to building lagged indicators that become inputs for a simple robo trader. The goal is to see how prices and returns can be transformed into trading signals while keeping the timing of the information explicit.
