# Asset allocation

```{r include=FALSE}
rm(list = ls())
library(PortfolioAnalytics)
library(ROI)
library(ROI.plugin.glpk)
library(ROI.plugin.quadprog)
library(dplyr)
library(tbl2xts)
library(tidyquant)
library(ggplot2)
library(MASS)
```

The previous chapter estimated return exposures one asset at a time. Asset allocation turns those individual assets into a portfolio decision. The practical question is how much capital should be assigned to each asset when expected returns, risk, correlations, and investment constraints are considered together.

Optimization chooses the best feasible solution under a set of constraints. In this chapter, the feasible solution is a vector of portfolio weights. The objective can be low risk, high expected return, a strong return-risk ratio, or a benchmark comparison. The constraints can require full investment, long-only positions, maximum weights, minimum weights, or rebalancing rules. R is useful because the calculation combines data transformation, matrix operations, optimization, and performance evaluation.

Diversification is the central idea. If all assets move together, portfolio risk remains close to single-asset risk. If assets respond differently to common shocks, portfolio volatility can fall even when expected return remains attractive. The correlation examples from earlier chapters prepare this point: co-movement determines how much risk can be reduced by combining assets.

We call the combined investment a portfolio, and each percentage invested in an individual asset is a portfolio weight. Asset allocation models estimate or test these weights. The chapter starts with a single-period FANG portfolio, then compares optimization methods, examines an equally weighted benchmark, introduces synthetic assets to isolate diversification, and finishes with rebalancing and performance evaluation.


## The single-period problem

The single-period problem is the simplest framework for asset allocation. At $t=0$, the investor observes historical information, chooses portfolio weights, and then realizes a portfolio return at $t=1$.

We use the <tt>`PortfolioAnalytics`</tt> package to distribute 100% of available capital across four stocks. For convenience, the example uses monthly returns from the <tt>`FANG`</tt> database.

If $w_i$ is the portfolio weight of asset $i$ and $R_i$ is its return, the portfolio return is:

$$
R_p=\sum_{i=1}^{N}w_iR_i.
$$

The portfolio expected return and variance are:

$$
E(R_p)=w'\mu,
\qquad
\sigma_p^2=w'\Sigma w.
$$

```{r}
FANG_monthly_returns <- FANG |>
    group_by(symbol) |>
    tq_transmute(select     = adjusted,
                 mutate_fun = periodReturn,
                 period     = "monthly",
                 type       = "arithmetic")

FANG_monthly_returns
```

The format of <tt>`FANG_monthly_returns`</tt> is not compatible with the <tt>`PortfolioAnalytics`</tt> package as the data is currently tidy. So, we have to transform the way our database looks. Originally, we have a column called <tt>`symbol`</tt> and we need each stock to have their corresponding name in different columns. We can implement this change easily.

```{r}
fang <- FANG_monthly_returns |>
  tbl_xts(spread_by = "symbol")

head(fang)
```

The <tt>`FANG`</tt> information is the same, but the format now matches the requirements of the portfolio functions. The next step is to calculate the correlation matrix for the 4 stocks.

```{r}
cor(fang)
```

The lowest correlation is between META and AMZN (0.1846197). This means that both firms exhibit a weak linear relationship. On the other hand, we have GOOG and AMZN with a correlation of 0.6171376, which suggest a stronger linear relationship between these stock returns. In principle, the lower the correlation of our assets, the greater the diversification possibilities when forming a portfolio. Remember the ice cream and the hot chocolate shop example, we assumed a correlation of −0.9 and we were supposed to decrease the volatility of our sales. In practice, it is not very easy to find negative correlated assets. However, we can achieve diversification gains as long as the correlation value is less than +1.

The investment recommendation is a vector of weights: how much to invest in each of the four assets.

We first define a portfolio specification with the help of <tt>`portfolio.spec()`</tt> function.

```{r}
# Create the portfolio specification
port_spec <- portfolio.spec(colnames(fang))
port_spec
```

Given the portfolio specification, we are interested in four assets. Now, add a full investment constraint such that the weights sum to 1. In practical terms, a sum exactly equal to 1 is too restrictive for the optimization, so we allow weights to sum between 0.99 and 1.01. The constraint is:

$$
0.99 \leq \sum_{i=1}^{N}w_i \leq 1.01.
$$

We use the <tt>`add.constraint()`</tt> function.

```{r}
port_spec <- add.constraint(portfolio = port_spec, type = "weight_sum",
                            min_sum = 0.99, max_sum = 1.01)
port_spec
```

Add a box constraint such that the portfolio optimal weights are between 0 and 1. This is equivalent as a *long only* constraint:

$$
0 \leq w_i \leq 1.
$$

```{r}
port_spec <- add.constraint(portfolio = port_spec, type = "box",
                            min = 0, max = 1)
port_spec
```

We are interested in risk and return. The optimization problem can be read as searching for weights that improve the risk-return trade-off under the constraints above:

$$
\max_w \frac{E(R_p)}{\sigma_p}
\quad \text{subject to the weight constraints.}
$$

```{r}
# Add an objective to minimize portfolio standard deviation
port_spec <- add.objective(portfolio = port_spec,
                           type = "risk",
                           name = "StdDev")
port_spec <- add.objective(portfolio = port_spec,
                           type = "return",
                           name = "mean")
port_spec
```

The portfolio specification indicates that we have four assets available, these are: META, AMZN, NFLX and GOOG. Constraints indicates that we are expected to spend 100% of the funds available, or very close to that. The constraint long refers that individual portfolio weights have to be positive.

The alternative of the box constraint is to allow the model to deliver negative portfolio weights. Negative portfolio weights represent short sales. Short selling is an investment strategy that speculates on the decline in a stock. Sometimes short selling is not allowed so the box constraint will deliver only positive portfolio weights.

The <tt>`optimize.portfolio()`</tt> function can solve the portfolio problem with several methods, constraints, and objectives. The next examples compare the random and ROI methods.

```{r}
# Solve the optimization problem.
set.seed(13)
opt_rand <- optimize.portfolio(fang, portfolio = port_spec,
                               optimize_method = "random", trace = TRUE)
opt_roi <- optimize.portfolio(fang, portfolio = port_spec,
                              optimize_method = "ROI", trace = TRUE)
# Show optimization results.
opt_rand
opt_roi
```



The output above shows the portfolio weights, return and risk of the two optimal portfolios. Before plotting the optimized results, we inspect the initial situation: the individual assets before considering any portfolio.

First, the relevant <tt>`FANG`</tt> data.

```{r}
FANG_monthly_returns <- FANG |>
    group_by(symbol) |>
    tq_transmute(select     = adjusted,
                 mutate_fun = periodReturn,
                 period     = "monthly",
                 type       = "arithmetic")

FANG_stats <- FANG_monthly_returns |>
summarise(mean = mean(monthly.returns),
          sd = sd(monthly.returns),
          sr = mean/sd)

FANG_stats
```


```{r}
#| label: fig-ftfia
#| fig-cap: "FANG, the four individual assets."
ggplot(FANG_stats, aes(x = sd, y = mean, color = symbol)) +
  geom_point(size = 5) +
  geom_text(aes(label = paste0(symbol)),
  vjust = 2, color = "black", size = 3.5) +
  labs(x = "Risk", y = "Return") +
  scale_y_continuous(limits = c(0.01, 0.06),
                     labels = scales::percent) +
  scale_x_continuous(labels = scales::percent) +
  theme_tq() +
  theme(legend.position = "none", legend.title = element_blank())
```

This plot shows the four individual assets in mean-variance space. Asset allocation adds a portfolio perspective: combine assets so the joint risk-return profile can improve over a single-asset choice. The optimization process searches for the allocation that best satisfies the objective and constraints.

For example, a portfolio with 3.5% monthly expected return and 15% standard deviation is dominated by alternatives with similar return and lower risk. A portfolio with 4% expected return and 5% risk would have a stronger return per unit of risk. This is the kind of comparison the optimizer formalizes.

The random technique (Random Portfolios Optimization) evaluates many investment recommendation alternatives, each one is one portfolio. In principle, the technique should recommend a portfolio which leads to a more attractive return per unit of risk alternative than investing in individual assets.

```{r}
#| label: fig-frmop
#| fig-cap: "FANG, random method optimal portfolio."
# Plot results.
chart.RiskReward(opt_rand, risk.col = "StdDev",
                 return.col = "mean", chart.assets = TRUE,
                 xlim = c(0, 0.10))
abline(v = 0.05941703, lty = 2)
abline(h = 0.02566966, lty = 2)
legend("bottomright", legend = c("opt_rand", "equally_weighted"),
       col = c("blue", "orange"), pch = 19, cex = 0.9)
```

According to our portfolio specification and assets, gray circles represent feasible portfolios in the sense that they are possible combinations given the individual assets. Note that there are some available portfolios with a lower risk than GOOG. In fact, the algorithm suggests the blue portfolio as the optimal portfolio. This blue portfolio is attractive because it has the same risk than GOOG, but it has higher expected return. Moreover, our optimal blue portfolio has a similar return as AMZN, but with less risk. So, apparently the blue alternative is an attractive investment recommendation. Note that the gray portfolios form a kind of frontier in this mean-variance plot. This frontier suggests that it is not possible to invest in a portfolio at the left of this frontier. The optimal portfolio lies just in the frontier.

See the blue optimal portfolio again. We would prefer any other portfolio located in the top (higher return) or at the left (lower risk), but that is impossible given the data. On the other hand, it is possible to achieve a portfolio located below (low return) or at the right (high risk), but that is not optimal. This is why optimal portfolios are those that lie in the frontier, those are the portfolios located in the extreme high and left of this mean-variance plot.

Then, what can we do to invest in the blue optimal portfolio?


```{r}
#| label: fig-frmow
#| fig-cap: "FANG, random method optimal weights."
chart.Weights(opt_rand)
```

According to the <tt>`opt_rand`</tt> portfolio, we have to invest:

```{r}
extractWeights(opt_rand)
```

```{r}
opt_rand_mean <- as.numeric(extractObjectiveMeasures(opt_rand)$mean)
opt_rand_sd <-
  as.numeric(extractObjectiveMeasures(opt_rand)$StdDev[1,1])
opt_rand_sr <- opt_rand_mean/opt_rand_sd

opt_rand_summary <-data.frame(mean = opt_rand_mean,
                              sd = opt_rand_sd, sr = opt_rand_sr)
row.names(opt_rand_summary) <- "opt_rand"

FANG_stats |>
  add_row(symbol = "opt_rand", opt_rand_summary) |>
  arrange(-sr)
```

The optimal portfolio <tt>`opt_rand`</tt> has a 0.4555797 return per unit of risk. This is clearly a better alternative compared with investing in individual assets as we show below.

A second criterion is to optimize according to ROI (R Optimization Infrastructure for linear and quadratic programming solvers).


```{r}
#| label: fig-frmop2
#| fig-cap: "FANG, ROI method optimal portfolio."
# Plot results.
chart.RiskReward(opt_roi, risk.col = "StdDev",
                 xlim = c(0.05, 0.18),
                 return.col = "mean", rp = TRUE, chart.assets = TRUE)
points(0.05807856, 0.02645942, pch = 19, cex = 1.5, col = "black")
legend("bottomright", legend = c("ROI", "opt_rand"),
       col = c("blue", "black"), pch = 19, cex = 0.9)
```

A different optimization criterion leads to a different optimal portfolio. However, they are both optimal as they are in the efficient frontier.


```{r}
#| label: fig-frmow2
#| fig-cap: "FANG, ROI method optimal weights."
chart.Weights(opt_roi)
```

According to the <tt>`opt_roi`</tt> portfolio, we have to invest:

```{r}
extractWeights(opt_roi)
```


```{r}
opt_roi_mean <- as.numeric(extractObjectiveMeasures(opt_roi)$mean)
opt_roi_sd <-
  as.numeric(extractObjectiveMeasures(opt_roi)$StdDev)
opt_roi_sr <- opt_roi_mean/opt_roi_sd

opt_roi_summary <-data.frame(mean = opt_roi_mean,
                              sd = opt_roi_sd, sr = opt_roi_sr)
row.names(opt_roi_summary) <- "opt_roi"

FANG_stats |>
  add_row(symbol = "opt_rand", opt_rand_summary) |>
  add_row(symbol = "opt_roi", opt_roi_summary) |>
  arrange(-sr)
```
The optimal portfolio <tt>`opt_roi`</tt> has a 0.4244362 return per unit of risk. This is higher than the individual assets but lower than the <tt>`opt_rand`</tt> portfolio. According to this return-risk ratio, <tt>`opt_rand`</tt> is the strongest portfolio in this comparison, with 0.4555797 return per unit of risk. The optimization process therefore produces a portfolio that improves on the individual assets in this historical sample. In practice, an allocation such as 28.2% in META, 16.2% in AMZN, 7.4% in NFLX, and 47.2% in GOOG would still require additional checks before implementation: estimation stability, transaction costs, liquidity, constraints, and investor suitability.

The orange portfolio represents an equally weighted benchmark: 25% in each asset. With $N=4$, the weight is $1/N=1/4$. This benchmark requires no optimization and still offers a diversified comparison point. In this example, it is close to the optimized alternatives, while the optimized portfolios remain stronger under the chosen criterion.

We now define a benchmark portfolio specification <tt>`port_naive`</tt> to force each weight to be exactly 0.25.

```{r}
port_naive <- portfolio.spec(colnames(fang))
port_naive <- add.constraint(portfolio = port_naive,
                             type = "full_investment")
port_naive <- add.constraint(portfolio = port_naive,
                             type = "long_only")
port_naive <- add.objective(portfolio = port_naive,
                           type = "risk",
                           name = "StdDev")
port_naive <- add.objective(portfolio = port_naive,
                           type = "return",
                           name = "mean")
```

Now, compare the naïve portfolio with the rest.

```{r}
opt_naive <- equal.weight(fang, portfolio = port_naive)
opt_naive_mean <- as.numeric(extractObjectiveMeasures(opt_naive)$mean)
opt_naive_sd <-
  as.numeric(extractObjectiveMeasures(opt_naive)$StdDev)
opt_naive_sr <- opt_naive_mean/opt_naive_sd
opt_naive_summary <-data.frame(mean = opt_naive_mean,
                              sd = opt_naive_sd, sr = opt_naive_sr)
row.names(opt_naive_summary) <- "opt_naive"

FANG_stats |>
  add_row(symbol = "opt_rand", opt_rand_summary) |>
  add_row(symbol = "opt_roi", opt_roi_summary) |>
  add_row(symbol = "opt_naive", opt_naive_summary) |>
  arrange(-sr)
```
The benchmark portfolio <tt>`port_naive`</tt> has the highest Sharpe ratio in this comparison, including the optimized alternatives. This result is informative: optimization depends on the objective, constraints, method, and sample estimates. We can add a target return constraint to search for an optimized portfolio that targets the benchmark portfolio's expected return.

```{r}
port_spec2 <- add.constraint(portfolio = port_spec, type = "return",
                             return_target = 0.03414333)
set.seed(13)
opt_roi2 <- optimize.portfolio(fang, portfolio = port_spec2,
                               optimize_method = "ROI", maxSR = TRUE,
                               rf = 0, trace = TRUE)
```


```{r}
#| label: fig-frmowtttnper
#| fig-cap: "FANG, ROI method optimal weights that targets the naïve portfolio expected return."
chart.Weights(opt_roi2)
```

According to the <tt>`opt_roi2`</tt> portfolio, we have to invest:

```{r}
extractWeights(opt_roi2)
```

```{r}
#| label: fig-frmtttnper
#| fig-cap: "FANG, ROI method that targets the naïve portfolio expected return."
# Plot results.
chart.RiskReward(opt_roi2, risk.col = "StdDev",
xlim = c(0.07, 0.073),
ylim = c(0.032, 0.036),
return.col = "mean", chart.assets = TRUE, cex = 2)
points(0.07176544, 0.03414333, pch = 19, col = "orange", cex = 2)
legend("bottomright", legend = c("opt_roi", "equally_weighted"),
col = c("blue", "orange"), pch = 19, cex = 0.8)
```

Now, compare the <tt>`opt_roi2`</tt> portfolio with the rest.

```{r}
opt_roi2_mean <- as.numeric(extractObjectiveMeasures(opt_roi2)$mean)
opt_roi2_sd <-
  as.numeric(extractObjectiveMeasures(opt_roi2)$StdDev)
opt_roi2_sr <- opt_roi2_mean/opt_roi2_sd
opt_roi2_summary <-data.frame(mean = opt_roi2_mean,
                              sd = opt_roi2_sd, sr = opt_roi2_sr)
row.names(opt_roi2_summary) <- "opt_roi2"

FANG_stats |>
  add_row(symbol = "opt_rand", opt_rand_summary) |>
  add_row(symbol = "opt_roi", opt_roi_summary) |>
  add_row(symbol = "opt_naive", opt_naive_summary) |>
  add_row(symbol = "opt_roi2", opt_roi2_summary) |>
  arrange(-sr)
```


The expected return in $t=0$ of the <tt>`opt_rand`</tt> portfolio is 0.02645942. The value of 0.02606294 is the sum of the <tt>`opt_rand`</tt> portfolio weights multiplied by the individual asset returns. The realized return at $t=1$ remains uncertain, so the next step uses a simple simulation to illustrate possible outcomes.

In this case, $t=0$ is 2016-12-30 and $t=1$ is 2017-01-31. The simulation uses the historical mean and standard deviation information available at $t=0$.

The simulation generates 1,000 observations for each individual asset and portfolio. This simulation is a simple teaching device. It treats each investment alternative as normally distributed using the historical mean and standard deviation shown above. It ignores time dependence, changing volatility, tail events, and the full correlation structure. The result is useful for intuition and should be read as a simplified risk model.

```{r}
# Number of simulations.
sim = 1000
set.seed (7)
# Simulation per stock and portfolio.
s_AMZN <- rnorm(sim, 0.02566966, 0.08172605)
s_FB <- rnorm(sim, 0.03412852, 0.09894324)
s_GOOG <- rnorm(sim, 0.01757620, 0.05941703)
s_NFLX <- rnorm(sim, 0.05919893, 0.16656346)
s_ran <- rnorm(sim, opt_rand$opt_values$mean, opt_rand$opt_values$StdDev)
s_roi <- rnorm(sim, opt_roi$opt_values$mean, opt_roi$opt_values$StdDev)
```

The simulation is done. A boxplot summarizes the simulated outcomes by showing the minimum, maximum, median, first quartile, and third quartile.

```{r}
#| label: fig-fsroia
#| fig-cap: "FANG, simulated returns of investment alternatives."
# The boxplot.
b = cbind(s_FB, s_AMZN, s_NFLX, s_GOOG, s_ran, s_roi)
boxplot(b, las = 2)
abline(h = 0, lty = 2)
```

The <tt>`opt_rand`</tt> portfolio appears well diversified and <tt>`opt_roi`</tt> is more risky under this simplified simulation. Real returns can have heavier tails, changing volatility, and co-movement across assets, so this should be read as a first approximation. Because the historical period is known, the realized returns on 2017-01-31 can be compared with the simulated distribution.

First, we need the 2017-01-31 actual or realized returns. The <tt>`FANG`</tt> database ends at 2016-12-30, and these historical returns are fixed observations. We store them directly here to avoid depending on the Meta ticker change or on current data-provider availability for old symbols.

```{r}
# 2017-01-31 realized individual asset returns.
r_stocks <- c(META = 0.1208283,
              AMZN = 0.101782,
              NFLX = 0.1076947,
              GOOG = 0.02058163)

r_stocks
```

These are the realized monthly returns at $t=1$. Please note that we have not introduced these returns before. Our model and portfolios ignores these returns. Now, we need to evaluate the realized return of both optimal portfolios. This is done by adding the multiplication of weights and individual realized returns. This approach is commonly known as out-of-sample evaluation.

```{r}
# Calculate realized portfolio returns.
realized_ret <- c(opt_rand = sum(opt_rand$weights * r_stocks),
                  opt_roi = sum(opt_roi$weights * r_stocks))

expected_ret <- c(opt_rand = opt_rand$opt_values$mean,
                  opt_roi = opt_roi$opt_values$mean)

data.frame(cbind(realized_ret, expected_ret))
```

In sum, the expected monthly return in $t=0$ of <tt>`opt_rand`</tt> portfolio is 0.02606294, and the realized monthly return in $t=1$ is 0.06958923. Not bad. The expected monthly return in $t=0$ of <tt>`opt_roi`</tt> is 0.04999645, and the realized monthly return in $t=1$ is 0.11390269.

The next plot combines the simulated distribution with the realized returns shown in red.

```{r}
#| label: fig-fsroiawrhb2ir
#| fig-cap: "FANG, simulated returns of investment alternatives. What really happened by 2017-01-30 in red."
# The simulation.
boxplot(b, las = 2)
abline(h = 0, lty = 2)
# The realized returns in red.
points(1, 0.1208283,  col = "red", pch = 19, cex = 2)
points(2, 0.101782,   col = "red", pch = 19, cex = 2)
points(3, 0.1076947,  col = "red", pch = 19, cex = 2)
points(4, 0.02058163, col = "red", pch = 19, cex = 2)
points(5, sum(opt_rand$weights * r_stocks), col = "red",
       pch = 19, cex = 2)
points(6, sum(opt_roi$weights * r_stocks),  col = "red",
       pch = 19, cex = 2)
```

The realized returns can be quite different from the simulated or expected returns. This difference depends on the method, the model, the database length, the ability of the portfolio designer, and also depends on a random component. Portfolio models produce estimates under assumptions; realized returns can land far away from those estimates. In our <tt>`FANG`</tt> example, the optimal investment recommendations were good alternatives.

The same allocation workflow can be replicated with the 10-stock data used in the previous asset-pricing example. This shows the value of reproducible code: once the data structure is consistent, the same modeling sequence can be reused with a different asset universe.

The data.
The set keeps the same broad sector logic used before: <tt>`VTR`</tt> represents healthcare real estate exposure and <tt>`VOD`</tt> represents telecommunications exposure.

```{r}
# Download individual asset returns.
R_stocks <- c("NEM", "AMCR", "CLX", "VTR", "KR", "TXN", "F", "TXT",
              "KLAC", "VOD") |>
  tq_get(get = "stock.prices", from = "2010-01-01",
         to = "2015-12-31") |>
  group_by(symbol) |>
  tq_transmute(select     = adjusted,
               mutate_fun = periodReturn,
               period     = "monthly",
               col_rename = "R_stocks")
R_stocks
```

The portfolio specification.

```{r}
capm_stocks <- R_stocks |>
  tbl_xts(spread_by = "symbol") |>
  na.fill(fill = 0.00)

port_spec <- portfolio.spec(colnames(capm_stocks))
port_spec <- add.constraint(portfolio = port_spec, type = "weight_sum",
                            min_sum = 0.99, max_sum = 1.01)
port_spec <- add.constraint(portfolio = port_spec, type = "box",
                            min = 0, max = 1)
port_spec <- add.objective(portfolio = port_spec,
                           type = "risk",
                           name = "StdDev")
port_spec <- add.objective(portfolio = port_spec,
                           type = "return",
                           name = "mean")
port_spec
```

The random method.

```{r}
#| label: fig-tarmop
#| fig-cap: "10-assets, random method optimal portfolio."
set.seed(13)
opt_rand <- optimize.portfolio(capm_stocks, portfolio = port_spec,
                          optimize_method = "random", trace = TRUE)
chart.RiskReward(opt_rand, risk.col = "StdDev",
                 return.col = "mean", chart.assets = TRUE)
legend("bottomleft", legend = c("opt_rand", "equally_weighted"),
       col = c("blue", "orange"), pch = 19, cex = 0.8)
abline(0, opt_rand$objective_measures$mean /
         opt_rand$objective_measures$StdDev, lwd = 2, col = "red")
```

Here, the slope of the red line corresponds to the optimal portfolio return per unit of risk. This means that the optimal portfolio has a similar return per unit of risk compared with KR. Note how the rest of the assets have a lower return per unit of risk.

```{r}
extractWeights(opt_rand)
```

```{r}
#| label: fig-tarmopw
#| fig-cap: "10-assets, random method optimal weights."
chart.Weights(opt_rand)
```

Finally the ROI method.

```{r}
#| label: fig-tarmop2
#| fig-cap: "10-assets, ROI method optimal portfolio."
opt_roi <- optimize.portfolio(capm_stocks, portfolio = port_spec,
                          optimize_method = "ROI", trace = TRUE)
chart.RiskReward(opt_roi, risk.col = "StdDev",
                 return.col = "mean", rp = TRUE, chart.assets = TRUE)
legend("bottomleft", legend = c("opt_roi"),
       col = c("blue"), pch = 19, cex = 0.8)
abline(0, opt_roi$objective_measures$mean /
         opt_roi$objective_measures$StdDev, lwd = 2, col = "red")

```
This solution is concentrated in KR, with a very small KLAC weight caused by the full-investment tolerance.

```{r}
extractWeights(opt_roi)
```

```{r}
#| label: fig-tarmow2
#| fig-cap: "10-assets, ROI method optimal weights."
chart.Weights(opt_roi)
```

## Diversification

In finance, diversification is the process of allocating capital (or creating a portfolio) in a way that reduces the exposure to any one particular asset or risk factor. We usually recommend to invest in a variety of assets to achieve an overall risk reduction of our portfolio. However, if those assets are as correlated as Visa and Mastercard (as discussed before), then our diversification efforts would not be so effective. In this section, we propose an experiment to illustrate the role of diversification in asset allocation. The experiment is to artificially generate two assets and add them to the <tt>`FANG`</tt> database, so we end up with 6 assets. The special characteristic of these two new assets $X$ and $Y$ is that they both have a very extreme negative correlation value. Once we add these two assets, we will repeat the portfolio optimization and see if we could do better than before.

First, we generate the returns of these two assets. The new assets $X$ and $Y$ do not exist in the real world; they are created with a simulation technique to isolate the diversification mechanism.

```{r}
set.seed(13)
data = mvrnorm(n = 48, mu = c(0.2, 0.5),
               Sigma = matrix(c(1, -1.4, -1.4, 2), nrow = 2),
               empirical = TRUE)/10
xy = as.data.frame(data)
X = xy$V1
Y = xy$V2
# Add X and Y to the fang database.
fang_xy <- fang
fang_xy$X <- X
fang_xy$Y <- Y
head(fang_xy)
```

The next calculation verifies that the two new assets are negatively correlated.

```{r}
cor(fang_xy)
```

The correlation of $X$ and $Y$ is −0.98994949, this is very close to −1, so they are strongly negatively correlated. This is just what we wanted. In fact, the correlation of these two new assets and the rest are also negative. In principle, we should expect a new optimal portfolio with a greater return per unit of risk.

Some may argue that investing in two inversely correlated assets would lead to a zero expected return. This is not the case as shown below.

The next step creates a sequence of $X$ and $Y$ portfolio weights.

```{r}
w_x <- seq(0, 1, 0.01)
w_y <- 1 - w_x
```

Then, the two assets portfolio expected return and risk.

```{r}
ret_xy <- w_x * mean(fang_xy$X) + w_y * mean(fang_xy$Y)
sd_xy <- (w_x^2 * sd(fang_xy$X)^2 +
          w_y^2 * sd(fang_xy$Y)^2 +
          2 * w_x * w_y * sd(fang_xy$X) * sd(fang_xy$Y) *
            cor(fang_xy$X, fang_xy$Y))^0.5
```

```{r}
#| label: fig-taef
#| fig-cap: "2-assets, efficient frontier."
plot(sd_xy, ret_xy, type = "l", xlim = c(0, 0.15), ylim = c(0, 0.05),
     xlab = "Risk", ylab = "Expected return")
abline(v = 0, lty = 2)
abline(h = 0, lty = 2)
points(sd(fang_xy$X), mean(fang_xy$X), cex = 2, pch = 19, col = "red")
points(sd(fang_xy$Y), mean(fang_xy$Y), cex = 2, pch = 19, col = "blue")
legend("bottomright", legend = c("X", "Y"),
       col = c("red", "blue"), pch = 19, cex = 1)
```
Which is the minimum portfolio risk?

```{r}
min(sd_xy)
```
Which is the observation that match the minimum portfolio risk?

```{r}
which.min(sd_xy)
```

```{r}
#| label: fig-taefamvp
#| fig-cap: "2-assets, efficient frontier and minimum variance portfolio."
plot(sd_xy, ret_xy, type = "l", xlim = c(0, 0.15), ylim = c(0, 0.05),
     xlab = "Risk", ylab = "Expected return")
abline(v = 0, lty = 2)
abline(h = 0, lty = 2)
points(sd(fang_xy$X), mean(fang_xy$X), cex = 2, pch = 19, col = "red")
points(sd(fang_xy$Y), mean(fang_xy$Y), cex = 2, pch = 19, col = "blue")
points(sd_xy[60], ret_xy[60], cex = 2, pch = 19)
legend("bottomright", legend = c("X", "Y"),
       col = c("red", "blue"), pch = 19, cex = 1)
```
Which are the minimum variance portfolio weights?

```{r}
w_x[60]
w_y[60]
```

Now consider the <tt>`FANG`</tt> database plus the two artificial assets $X$ and $Y$. The portfolio specification and optimization are rebuilt with this expanded asset universe.

```{r}
# Create the portfolio specification
port_spec <- portfolio.spec(colnames(fang_xy))
port_spec <- add.constraint(portfolio =
                              port_spec, type = "full_investment")
port_spec <- add.constraint(portfolio = port_spec, type = "long_only")
port_spec <- add.objective(portfolio = port_spec,
                           type = "risk",
                           name = "StdDev")
port_spec <- add.objective(portfolio = port_spec,
                           type = "return",
                           name = "mean")
# Optimization.
set.seed(13)
opt_xy <- optimize.portfolio(fang_xy, portfolio = port_spec,
                          optimize_method = "random", trace = TRUE)
```

The results can now be visualized.

```{r}
#| label: fig-fpxayrp
#| fig-cap: "FANG plus X and Y, random portfolio."
# Plot results.
chart.RiskReward(opt_xy, risk.col = "StdDev",
                 xlim = c(0, 10),
                 return.col = "mean", chart.assets = TRUE)
points(0.05807126, 0.02606294, pch = 19, cex = 1.5, col = "black")
points(0.1170653,0.04960399, pch = 19, cex = 1.5, col = "red")
legend("topleft", legend = c("opt_xy (FANG+X+Y)",
       "opt_rand (FANG)", "opt_roi (FANG)", "opt_naive (FANG)"),
       col = c("blue", "black", "red", "orange"), pch = 19, cex = 0.9)
```

This looks great as the new assets contribute to the optimization process to deliver a less risky portfolio. Now, the optimal portfolio has almost the same return as Meta but with a significant reduced risk. Then, correlation value plays a determinant role in the diversification process. This is why investors are looking for low, or even better, negatively correlated assets.

The details of the new portfolio show how the optimizer uses the artificial diversification opportunity.

```{r}
# Extract weight, risk and return.
extractWeights(opt_xy)

opt_xy_mean <- as.numeric(extractObjectiveMeasures(opt_xy)$mean)
opt_xy_sd <-
  as.numeric(extractObjectiveMeasures(opt_xy)$StdDev)
opt_xy_sr <- opt_xy_mean/opt_xy_sd

opt_xy_summary <-data.frame(mean = opt_xy_mean,
                              sd = opt_xy_sd, sr = opt_xy_sr)
row.names(opt_xy_summary) <- "opt_xy"

FANG_stats |>
  add_row(symbol = "opt_rand", opt_rand_summary) |>
  add_row(symbol = "opt_roi", opt_roi_summary) |>
  add_row(symbol = "opt_naive", opt_naive_summary) |>
  add_row(symbol = "opt_roi2", opt_roi2_summary) |>
  add_row(symbol = "opt_xy", opt_xy_summary) |>
  arrange(-sr)
```

Note that the portfolio assigns high weights to $X$ and $Y$ with values of 52.2% and 43.4%. The rest is 0.2% in META, 2.8% in AMZN, 0% in NFLX, and 1.4% in GOOG.


The <tt>`opt_xy`</tt> portfolio shows a large improvement because the artificial assets combine high average returns with a very strong negative correlation. This example is designed to isolate the diversification mechanism, so it should be interpreted as a controlled experiment. In practice, investors evaluate return together with volatility, correlation, liquidity, estimation error, and implementation constraints. Low or negatively correlated assets can reduce portfolio volatility and improve the return per unit of risk, while realized returns remain uncertain.

The next step for a dissatisfied investor would be to search for assets with different exposures to risk factors, estimate their joint behavior, and rerun the optimization with realistic constraints. Sector, geography, currency, liquidity, and macroeconomic exposure are natural places to start that search.

## Rebalancing portfolio and evaluation

In the previous section we calculate portfolio weights once. We changed the optimization criteria and we added new assets, but we only calculate the portfolio weights once. Rebalancing is something very common in finance, it means to calculate portfolio weights as time passes. This makes sense because as time passes we have access to new information (stock returns) and we should re-balance our portfolio in order to take into account this new information. Evaluation is also a very common task in finance. We are interested to know what is the annualized return of an investment strategy in a period of time. In this section we are going to extend the asset allocation problem to incorporate rebalancing portfolio and evaluation. We are going to use a different database to illustrate our results.

Consider the following investment process. At $t=0$, the investor has access to 60 months of historical information for a set of individual assets. The data from $t=-60$ to $t=0$ is used to estimate portfolio weights and form the portfolio at $t=0$. Returns are then realized for $t=1$, $t=2$, and $t=3$. At $t=3$, the portfolio is re-estimated using the rolling window from $t=-57$ to $t=3$. The same procedure continues through the investment period. The evaluation question is whether the rebalanced strategy produces stronger annualized return and return per unit of risk than a benchmark.

This process requires repeated data transformation, optimization, implementation, and evaluation. R makes the workflow reproducible because each rebalance date follows the same script. That automation is important in finance: the analyst can focus on assumptions, constraints, risk controls, and interpretation while the computer performs the repeated calculations.

We start with the data needed for the rebalancing exercise.


```{r}
# Get the data.
data(indexes)
returns <- indexes[, 1:4]
tail(returns)
```

The database goes from 1980-01-31 to 2009-12-31. The set of assets are: US Bonds, US Equities, Int'l Equities and Commodities. This means that the first investment recommendation is calculated with information from 1980-01-31 to 1985-01-31 (60 monthly observations), the second from 1980-05-31 to 1985-09-31, and so on until the last period that goes from 2009-09-31 to 2009-12-31. The first monthly return will be the one on 1985-01-31 and the last on 2009-12-31, these are 300 monthly portfolio returns.

Before calculating optimal portfolios, we calculate a benchmark portfolio. This allows the optimized portfolio to be compared with a simple alternative. In this case, the benchmark is an equally weighted portfolio, investing 25% in each asset for all periods. With $N=4$ assets:

$$
w_i=\frac{1}{4}=0.25.
$$

This equally weighted portfolio implies that we will not conduct any optimization. It is like investing 25% in each asset (or index in this case) and doing nothing until the end of the investment period.

```{r}
# Equal weight benchmark.
n <- ncol(returns)
equal_weights <- rep(1 / n, n)
benchmark_returns <- Return.portfolio(R = returns,
                                      weights = equal_weights,
                                      rebalance_on = "quarters")
colnames(benchmark_returns) <- "benchmark"
# Benchmark performance.
table.AnnualizedReturns(benchmark_returns)
```

We are interested in a benchmark because it would facilitate our evaluation task.

Now we define the portfolio specification as we did in the section before.

```{r}
# Base portfolio specification.
base_port_spec <- portfolio.spec(assets = colnames(returns))
#base_port_spec <- add.constraint(portfolio = base_port_spec,
#                                 type = "full_investment")
#base_port_spec <- add.constraint(portfolio = base_port_spec,
#                                 type = "long_only")
base_port_spec <- add.constraint(portfolio = base_port_spec,
                                 type = "weight_sum",
                                 min_sum = 0.99, max_sum = 1.01)
base_port_spec <- add.constraint(portfolio = base_port_spec,
                                 type = "box", min = 0, max = 1)
base_port_spec <- add.objective(portfolio = base_port_spec,
                                type = "risk", name = "StdDev")
```

We are ready to implement the investment process described before.

```{r}
# Run the optimization with periodic rebalancing.
opt_base <- optimize.portfolio.rebalancing(R = returns,
            optimize_method = "ROI", portfolio = base_port_spec,
            rebalance_on = "quarters", training_period = 60,
            rolling_window = 60)
# Calculate portfolio returns.
base_returns <- Return.portfolio(returns, extractWeights(opt_base))
colnames(base_returns) <- "base"
```

The rebalancing and evaluation calculations are complete. The next output shows the results.

```{r}
#| label: fig-iow
#| fig-cap: "Indexes, optimal weights."
# Chart the optimal weights.
chart.Weights(opt_base)
```

The chart shows the contribution of each asset through the rebalancing process. US bonds dominate the allocation in this run, which reflects the optimization objective and historical inputs. The next output reports annualized performance.

```{r}
# Merge benchmark and portfolio returns.
ret <- cbind(benchmark_returns, base_returns)
# Annualized performance.
table.AnnualizedReturns(ret)
```

We did better than our benchmark portfolio, that is good.

Investment mandates often impose limits on individual assets. Suppose the portfolio is not allowed to invest more than 40% in US bonds. This constraint can be incorporated directly before rerunning the rebalancing exercise.

```{r}
# Make a copy of the portfolio specification.
box_port_spec <- base_port_spec
# Update the constraint.
box_port_spec <- add.constraint(portfolio = box_port_spec,
                                type = "box", min = 0.05, max = 0.4,
                                indexnum = 2)
# Backtest.
opt_box <- optimize.portfolio.rebalancing(R = returns,
                                          optimize_method = "ROI",
                                          portfolio = box_port_spec,
                                          rebalance_on = "quarters",
                                          training_period = 60,
                                          rolling_window = 60)
# Calculate portfolio returns.
box_returns <- Return.portfolio(returns, extractWeights(opt_box))
colnames(box_returns) <- "box"
```

In principle, more constraints will lead to a worse solution. The optimization algorithm works better as long as it can freely choose portfolio weights.

```{r fig.cap="Indexes, optimal weights with box constraint."}
# Chart the optimal weights.
chart.Weights(opt_box)
```

Now, the rest of the assets have a more significant role in our portfolio.

```{r}
# Merge box portfolio returns.
ret <- cbind(ret, box_returns)
# Annualized performance.
table.AnnualizedReturns(ret)
```

As expected, more constraints negatively impact the possibilities to get a higher annualized return.

Most of these activities including providing investment recommendations in a regular manner are services offered by private firms. People looking to invest part of their money are supposed to receive assistance and guidance and many times it is in the form of an investment recommendation like the ones we calculate here. In the past, these services were provided only by those firms with access to the knowledge, capital, experience, and technology. Nowadays, there are an increasing number of fintechs that provide financial services because now we have access to powerful software and improved computational capabilities. In any case, functions like <tt>`optimize.portfolio()`</tt> are expected to be used as a tool by professionals and not as a pure source of investment recommendation.

```{r}
#| label: fig-ipp
#| fig-cap: "Indexes, portfolio performance."
charts.PerformanceSummary(R = ret)
```

Portfolio allocation gives us weights, realized returns, and benchmark comparisons. The next step is to express the same portfolio problem in terms of possible losses. Value at Risk provides a compact way to ask how large a one-day portfolio loss could be under a chosen confidence level and a specific modeling assumption.
