# Blockchain

```{r blockchain-setup, include=FALSE}
rm(list = ls())

library(digest)
library(dplyr)
library(ggraph)
library(ggplot2)
library(igraph)
library(openssl)
library(tibble)
library(tidygraph)

set.seed(13)
theme_set(theme_minimal(base_size = 12))
```

The previous chapters worked with market data, returns, trading rules, asset
pricing, portfolio allocation, and market risk. This chapter changes the object
of analysis. A blockchain is treated here as a financial-computing mechanism:
a way to maintain a shared record of transfers, validate changes to that
record, and coordinate incentives without requiring a single central ledger.

The financial question is direct: how can a system record transfers of value
when many participants can read the ledger, submit transactions, and verify the
history? The answer combines several ideas:

* a ledger that records transactions;
* blocks that group transactions;
* cryptographic hashes that link blocks;
* digital signatures that authorize transactions;
* a consensus rule that tells nodes which history to accept;
* incentives that reward participants who help maintain the ledger.

The chapter uses Bitcoin as the motivating case, then builds the technical
pieces needed to understand it. Bitcoin is useful because it turns a ledger
mechanism into an economic system with wallets, private keys, miners, fees,
scarce issuance, public verification, and costly history rewriting. Once those
pieces are visible, tokenization becomes a natural financial extension: a
ledger can also record ownership units and cash-flow allocation rules for
claims whose value may depend on assets outside the ledger.

The route is therefore:

$$
\mathrm{Bitcoin\ as\ motivation}
\longrightarrow
\mathrm{generic\ ledger\ mechanism}
\longrightarrow
\mathrm{Bitcoin\ proof\ of\ work\ case}
\longrightarrow
\mathrm{tokenized\ financial\ claims}.
$$

The examples are intentionally small teaching models. They show the logic of
the system at a scale where each moving part can be inspected.

A useful way to read the chapter is through four frictions: a user may try to
spend the same value twice; block space can become scarce; settlement becomes
more reliable as confirmations accumulate; tokenized claims need a connection
between the ledger record and the legal or operational asset. These frictions
make blockchain a financial topic, because they connect technology to scarcity,
incentives, custody, and settlement.

## Bitcoin as motivating case

Students usually meet blockchain through Bitcoin prices, wallets, exchanges,
and transaction confirmations. Those entry points are useful, but they hide the
mechanism. A Bitcoin payment is a signed instruction that spends previous
outputs, moves through a peer-to-peer network, competes for block space, and
becomes harder to reverse as proof-of-work accumulates above it.

The financial modeling question is therefore richer than "what is the price of
Bitcoin?" We also need to ask:

| Object | Modeling question | Financial meaning |
|---|---|---|
| Private key | Who can authorize spending? | Custody and operational control |
| Transaction | Which value is being transferred, and under what fee? | Payment, settlement, and fee pressure |
| Block | Which transactions enter the next ledger update? | Scarce inclusion and audit trail |
| Proof-of-work | How costly is it to replace recent history? | Settlement confidence |
| Network | How do transactions and blocks propagate? | Latency, monitoring, and public observability |
| Token rule | What claim does a ledger entry represent? | Ownership, cash-flow allocation, and legal linkage |

The path from familiar market objects to protocol mechanics can be drawn as a
short workflow.

```{r}
#| label: fig-bitcoin-motivation-map
#| fig-cap: "From market-facing Bitcoin objects to protocol mechanics."
#| fig-width: 8.4
#| fig-height: 2.3
motivation_nodes <- tibble(
  id = c("price", "account", "wallet", "transaction",
         "mempool", "block", "confirmations"),
  label = c("Price\nquote", "Exchange\naccount", "Wallet\nkey",
            "Signed\ntransaction", "Mempool", "Block",
            "Confirmations"),
  layer = c("Market", "Market", "Custody", "Protocol",
            "Protocol", "Protocol", "Settlement"),
  x = seq(0.6, 7.8, length.out = 7),
  y = 1
)

motivation_edges <- motivation_nodes |>
  arrange(x) |>
  mutate(xend = lead(x), yend = lead(y)) |>
  filter(!is.na(xend))

ggplot() +
  geom_segment(
    data = motivation_edges,
    aes(x = x + 0.38, y = y, xend = xend - 0.38, yend = yend),
    arrow = grid::arrow(length = grid::unit(0.11, "inches")),
    color = "#6C7A89", linewidth = 0.7
  ) +
  geom_rect(
    data = motivation_nodes,
    aes(xmin = x - 0.45, xmax = x + 0.45,
        ymin = y - 0.26, ymax = y + 0.26, fill = layer),
    color = "white", linewidth = 0.8
  ) +
  geom_text(
    data = motivation_nodes,
    aes(x = x, y = y, label = label),
    size = 2.9, lineheight = 0.88
  ) +
  scale_fill_manual(values = c(
    "Market" = "#FDEBD0",
    "Custody" = "#D5F5E3",
    "Protocol" = "#D6EAF8",
    "Settlement" = "#E8DAEF"
  )) +
  coord_cartesian(xlim = c(0.05, 8.35), ylim = c(0.62, 1.38),
                  expand = FALSE, clip = "off") +
  guides(fill = guide_legend(title = "Layer")) +
  theme_void() +
  theme(
    legend.position = "bottom",
    plot.margin = margin(2, 4, 2, 4)
  )
```

The figure sets the chapter's stance. A price quote or an exchange account is
one visible surface of the system. The modeling work begins when custody,
authorization, pending settlement, block inclusion, and confirmations are tied
to the same record of value.

The rest of the chapter builds these objects in small pieces. Each teaching
model corresponds to a production concern: double spending, block construction,
signatures, custody, network propagation, confirmations, and tokenized claims.

## Generic blockchain model

The generic model is a compact laboratory for the Bitcoin case. It uses an
account-based ledger because balances are easy to read: each wallet has a
balance, transactions move value across wallets, fees compensate the block
producer, and the final ledger state can be checked directly. Bitcoin uses a
different accounting model, based on unspent transaction outputs, and that
difference is introduced after the generic mechanism is in place.

The generic model answers four questions:

| Component | Question | Financial role |
|---|---|---|
| Ledger state | Who owns what before and after transfers? | Records balances or ownership claims |
| Transaction rule | Which transfers are valid? | Prevents invalid spending and malformed transfers |
| Block structure | Which transactions enter the next ledger update? | Groups accepted transfers into an auditable batch |
| Consensus rule | Which history should nodes extend? | Coordinates participants around one accepted ledger |

The sequence to keep in view is:

$$
\mathrm{current\ ledger}
\rightarrow
\mathrm{valid\ transactions}
\rightarrow
\mathrm{candidate\ block}
\rightarrow
\mathrm{accepted\ block}
\rightarrow
\mathrm{updated\ ledger}.
$$

**A shared ledger problem**

A conventional financial ledger records who owns what and how balances change
after transactions. In a centralized system, one institution updates the
ledger, rejects invalid transfers, and resolves conflicts. A blockchain moves
part of that logic into a shared protocol: many participants can hold copies of
the ledger, check the same rules, and converge on an accepted history.

The most basic financial state is a table of balances. Suppose four wallets
start with the following balances.

```{r}
balances <- tibble(
  wallet = c("Alice", "Bob", "Carol", "Dave", "Miner"),
  balance = c(100, 35, 15, 5, 0)
)

balances
```

A transaction changes that state. If wallet $A$ sends amount $x$ to wallet
$B$ and pays fee $f$, the accounting update is:

$$
\mathrm{Balance}_A^{\prime} = \mathrm{Balance}_A - x - f,
\qquad
\mathrm{Balance}_B^{\prime} = \mathrm{Balance}_B + x,
\qquad
\mathrm{Balance}_{\mathrm{miner}}^{\prime}
= \mathrm{Balance}_{\mathrm{miner}}+f.
$$

The transfer is valid only if the sender has enough balance and the transaction
format is acceptable. The rule can be written as:

$$
\mathrm{Balance}_A \geq x+f,
\qquad
x>0,
\qquad
f \geq 0.
$$

This accounting rule is the starting point for the blockchain workflow. The
ledger state gives balances at one moment; the block history explains how those
balances were reached. For finance, that audit path matters because ownership
claims are stronger when they can be traced through accepted transactions.

**Double-spending pressure**

The first interesting problem is conflict. If Alice has 100 units and submits
two transactions that each spend about 70 units, both transactions cannot settle
inside the same ledger history:

$$
x_1+f_1+x_2+f_2 > \mathrm{Balance}_A.
$$

The system needs a rule for deciding which transaction enters the accepted
history. In a centralized ledger the institution decides. In a blockchain
network, the accepted history comes from transaction validity, block inclusion,
and consensus.

```{r}
double_spend_attempt <- tibble(
  tx_id = c("tx_A", "tx_B"),
  sender = c("Alice", "Alice"),
  receiver = c("Bob", "Carol"),
  amount = c(70, 70),
  fee = c(0.20, 0.40)
)

apply_ledger_order <- function(pool, balances, miner = "Miner") {
  balances_after <- balances
  outcome <- pool |>
    mutate(total_spent = amount + fee, status = NA_character_)

  for (k in seq_len(nrow(outcome))) {
    tx <- outcome[k, ]
    sender_balance <- balances_after$balance[
      balances_after$wallet == tx$sender
    ]

    valid <- sender_balance >= tx$amount + tx$fee

    if (valid) {
      outcome$status[k] <- "Accepted"
      balances_after$balance[balances_after$wallet == tx$sender] <-
        balances_after$balance[balances_after$wallet == tx$sender] -
        tx$amount - tx$fee
      balances_after$balance[balances_after$wallet == tx$receiver] <-
        balances_after$balance[balances_after$wallet == tx$receiver] +
        tx$amount
      balances_after$balance[balances_after$wallet == miner] <-
        balances_after$balance[balances_after$wallet == miner] + tx$fee
    } else {
      outcome$status[k] <- "Rejected"
    }
  }

  outcome
}

double_spend_outcome <- bind_rows(
  apply_ledger_order(double_spend_attempt, balances) |>
    mutate(ordering = "tx_A arrives first"),
  apply_ledger_order(arrange(double_spend_attempt, desc(fee)), balances) |>
    mutate(ordering = "Higher fee arrives first")
)

double_spend_outcome
```

```{r}
#| label: fig-double-spend-conflict
#| fig-cap: "Conflicting transactions under two ordering rules."
#| fig-width: 7.2
#| fig-height: 3.1
ggplot(
  double_spend_outcome,
  aes(x = tx_id, y = total_spent, fill = status)
) +
  geom_col(width = 0.62) +
  facet_wrap(~ ordering) +
  scale_fill_manual(values = c("Accepted" = "#27AE60",
                               "Rejected" = "#C0392B")) +
  labs(x = "Transaction", y = "Amount plus fee", fill = "")
```

The figure shows why a shared ledger needs more than arithmetic. The balances
tell us that the two transfers cannot both be final. The network still needs a
mechanism for selecting one accepted history and rejecting conflicting updates.

**Blocks**

A block is a container for ledger information. In a simple teaching example,
the block contains:

* a block number;
* a timestamp;
* transaction data;
* the hash of the previous block;
* a nonce used by proof-of-work examples;
* the block's own hash.

The diagram below shows the structure.

```{r}
#| label: fig-block-structure
#| fig-cap: "A simplified block structure."
#| fig-width: 7
#| fig-height: 2.6
block_boxes <- tibble(
  label = c("Block number", "Timestamp", "Transactions",
            "Previous hash", "Nonce", "Block hash"),
  xmin = c(0, 0, 0, 3.4, 3.4, 3.4),
  xmax = c(3, 3, 3, 6.4, 6.4, 6.4),
  ymin = c(2.4, 1.2, 0, 2.4, 1.2, 0),
  ymax = c(3.2, 2, 0.8, 3.2, 2, 0.8)
)

ggplot(block_boxes) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "#EAF2F8", color = "#1B4F72", linewidth = 0.7
  ) +
  geom_text(aes(x = (xmin + xmax) / 2, y = (ymin + ymax) / 2, label = label),
            size = 3.7) +
  annotate("text", x = 3.2, y = 3.45, label = "Block", fontface = "bold",
           size = 4.2) +
  coord_cartesian(xlim = c(-0.1, 6.5), ylim = c(-0.05, 3.65), expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(0, 0, 0, 0))
```

In R, a block can be represented as a list.

```{r}
block <- list(
  number = 1,
  timestamp = "2026-01-01 09:00:00",
  data = "Alice pays Bob 10",
  parent_hash = "0",
  nonce = 0
)

tibble(
  field = names(block),
  value = vapply(block, as.character, character(1))
)
```

This object still needs a hash before it can be chained to a history.

The block gives the system a unit of agreement. The network evaluates a batch
of transactions together with the previous history that the batch extends. The
parent hash records that link: it identifies the past ledger state that the new
block is trying to update.

**Hashes**

A cryptographic hash function maps an input of arbitrary size into a fixed-size
digest. A useful hash function for a ledger has three properties:

* the same input gives the same hash;
* a small change in the input gives a very different hash;
* the input cannot be reconstructed from the hash in any practical way.

For block $i$, we can write the hash as:

$$
h_i = H(\mathrm{data}_i, \mathrm{parentHash}_i,
\mathrm{timestamp}_i, \mathrm{nonce}_i).
$$

For non-genesis blocks, the parent hash equals the hash of the previous block:

$$
\mathrm{parentHash}_i = h_{i-1}.
$$

The examples use SHA-256 through the <tt>`digest`</tt> package. SHA-256 returns
256 bits, which correspond to 32 bytes or 64 hexadecimal characters.

```{r}
hash_example <- digest("Alice pays Bob 10", algo = "sha256")

tibble(
  message = "Alice pays Bob 10",
  algorithm = "SHA-256",
  hex_characters = nchar(hash_example),
  hash_prefix = substr(hash_example, 1, 16)
)
```

A tiny change in the message changes the digest.

```{r}
hash_comparison <- tibble(
  message = c("Alice pays Bob 10", "Alice pays Bob 11"),
  sha256 = vapply(message, digest, character(1), algo = "sha256")
) |>
  mutate(hash_prefix = substr(sha256, 1, 18)) |>
  select(message, hash_prefix)

hash_comparison
```

The table shows only hash prefixes to keep the output readable. The important
point is that the two prefixes already diverge after a one-character change,
while each full SHA-256 digest remains 64 hexadecimal characters long.

The following helper creates a block and then hashes the full block content.

```{r}
hash_block <- function(block) {
  block$hash <- NULL
  digest(block, algo = "sha256")
}

create_block <- function(number, data, parent_hash = "0",
                         nonce = 0, timestamp = Sys.time()) {
  block <- list(
    number = number,
    timestamp = as.character(timestamp),
    data = data,
    parent_hash = parent_hash,
    nonce = nonce
  )
  block$hash <- hash_block(block)
  block
}

block1 <- create_block(
  number = 1,
  data = "Genesis block",
  parent_hash = "0",
  timestamp = "2026-01-01 09:00:00"
)

block2 <- create_block(
  number = 2,
  data = "Alice pays Bob 10",
  parent_hash = block1$hash,
  timestamp = "2026-01-01 09:01:00"
)

block3 <- create_block(
  number = 3,
  data = "Bob pays Carol 5",
  parent_hash = block2$hash,
  timestamp = "2026-01-01 09:02:00"
)

chain <- list(block1, block2, block3)

chain_summary <- tibble(
  block = vapply(chain, function(x) x$number, numeric(1)),
  data = vapply(chain, function(x) x$data, character(1)),
  parent_hash = vapply(chain, function(x) substr(x$parent_hash, 1, 12),
                       character(1)),
  block_hash = vapply(chain, function(x) substr(x$hash, 1, 12),
                      character(1))
)

chain_summary
```

The summary keeps the block order, data, parent commitment, and block
commitment visible without printing the full nested R objects.

The financial meaning of this construction is integrity. A block hash is a
compact cryptographic commitment to the block content, and a parent hash is a
commitment to the previous accepted history. Linking those commitments means
that the latest block implicitly refers to every earlier accepted block. If an
old transaction changes, the old hash changes, and later blocks no longer point
to the same history.

**Chain validation**

A basic validation rule checks two conditions:

* each block hash matches the content of that block;
* each non-genesis block points to the hash of the previous block.

```{r}
validate_chain <- function(chain) {
  for (i in seq_along(chain)) {
    block <- chain[[i]]
    declared_hash <- block$hash
    expected_hash <- hash_block(block)

    if (!identical(declared_hash, expected_hash)) {
      return(FALSE)
    }

    if (i == 1 && block$parent_hash != "0") {
      return(FALSE)
    }

    if (i > 1 && block$parent_hash != chain[[i - 1]]$hash) {
      return(FALSE)
    }
  }

  TRUE
}

tibble(check = "Original chain", valid = validate_chain(chain))
```

The chain can be drawn as a sequence of blocks connected by hashes.

```{r}
#| label: fig-chain-hashes
#| fig-cap: "Blocks linked by parent hashes."
#| fig-width: 7.2
#| fig-height: 1.8
chain_plot <- tibble(
  block = c("Block 1", "Block 2", "Block 3"),
  x = c(1, 4, 7),
  y = 1,
  hash = c("h1", "h2", "h3")
)

ggplot(chain_plot) +
  geom_rect(
    aes(xmin = x - 0.9, xmax = x + 0.9, ymin = y - 0.45, ymax = y + 0.45),
    fill = "#E8F6F3", color = "#117864", linewidth = 0.8
  ) +
  geom_text(aes(x = x, y = y + 0.12, label = block), fontface = "bold",
            size = 3.8) +
  geom_text(aes(x = x, y = y - 0.16, label = paste0("hash = ", hash)),
            size = 3.2) +
  geom_segment(
    data = tibble(x = c(1.9, 4.9), xend = c(3.1, 6.1), y = 1, yend = 1),
    aes(x = x, xend = xend, y = y, yend = yend),
    arrow = grid::arrow(length = grid::unit(0.18, "inches")),
    color = "#566573", linewidth = 0.8
  ) +
  annotate("text", x = 2.5, y = 1.33, label = "parent_hash = h1", size = 3) +
  annotate("text", x = 5.5, y = 1.33, label = "parent_hash = h2", size = 3) +
  coord_cartesian(xlim = c(0, 8), ylim = c(0.35, 1.6), expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(0, 0, 0, 0))
```

The arrows carry the key dependency: block 2 names the hash of block 1, and
block 3 names the hash of block 2. A change in an earlier block breaks the
labels that later blocks use to identify their parent history.

Now change an old block.

```{r}
tampered_chain <- chain
tampered_chain[[2]]$data <- "Alice pays Bob 100"

tibble(check = "Changed block 2 data", valid = validate_chain(tampered_chain))
```

The validation fails because the declared hash of block 2 no longer matches
the altered content. If block 2 is rehashed, block 3 still points to the old
block 2 hash.

```{r}
tampered_rehashed <- tampered_chain
tampered_rehashed[[2]]$hash <- hash_block(tampered_rehashed[[2]])

tibble(check = "Rehashed block 2 only",
       valid = validate_chain(tampered_rehashed))
```

This is the core of the chain idea. A historical alteration requires updating
the altered block and every later block. In a real network, the attacker would
also need the rest of the network to accept the altered history.

The teaching validation function checks structural consistency. Real systems add
transaction signatures, double-spending rules, block-size or block-weight
limits, timestamp rules, consensus-specific constraints, and network-level
acceptance. The simplified function preserves the main idea: a valid ledger
history is a sequence whose content and links agree.

**Proof-of-work**

Hashes are fast to compute, so hashing alone does not make history expensive
to rewrite. Proof-of-work adds a cost. A miner searches for a nonce such that
the block hash satisfies a difficulty condition. In the teaching rule below, the
hash must start with $d$ zeros:

$$
\operatorname{substr}(H(\mathrm{block}, \mathrm{nonce}),1,d)
=\underbrace{00\cdots0}_{d\ \mathrm{zeros}}.
$$

Verification is cheap because any node can hash the block once and check the
leading zeros. Finding a valid nonce requires repeated trials.

This asymmetry is the key idea. Producing the proof is costly; checking it is
cheap. In a financial ledger, that asymmetry can be used to make accepted
history costly to replace. The block remains visible and verifiable; the
barrier is the amount of work needed to find an alternative valid chain.

```{r}
proof_of_work <- function(block, difficulty) {
  target <- paste(rep("0", difficulty), collapse = "")
  nonce <- 0

  repeat {
    block$nonce <- nonce
    candidate_hash <- hash_block(block)

    if (substr(candidate_hash, 1, difficulty) == target) {
      return(list(nonce = nonce, hash = candidate_hash))
    }

    nonce <- nonce + 1
  }
}

pow_block <- list(
  number = 4,
  timestamp = "2026-01-01 09:03:00",
  data = "Carol pays Dave 3",
  parent_hash = block3$hash
)

pow_result_example <- proof_of_work(pow_block, difficulty = 2)

tibble(
  difficulty = 2,
  nonce = pow_result_example$nonce,
  hash_start = substr(pow_result_example$hash, 1, 16)
)
```

The number of trials tends to rise quickly with difficulty.

```{r}
difficulty_table <- tibble(difficulty = 1:4) |>
  rowwise() |>
  mutate(nonce = proof_of_work(pow_block, difficulty)$nonce) |>
  ungroup()

difficulty_table
```

```{r}
#| label: fig-pow-difficulty
#| fig-cap: "Proof-of-work trials under increasing teaching difficulty."
#| fig-width: 7
#| fig-height: 3.4
ggplot(difficulty_table, aes(x = difficulty, y = nonce)) +
  geom_line(color = "#7B241C", linewidth = 1) +
  geom_point(color = "#C0392B", size = 3) +
  scale_x_continuous(breaks = difficulty_table$difficulty) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Difficulty: required leading zeros", y = "Nonce found")
```

The exact nonce depends on the block content, but the figure shows why
difficulty matters: adding required leading zeros turns hashing into repeated
search. Verification still requires a single hash check.

Bitcoin is the canonical proof-of-work case. Miners gather valid transactions,
construct candidate blocks, search for a valid nonce, and broadcast the block
when the proof is found. Ethereum is useful as a contrast because it moved from
proof-of-work to proof-of-stake in 2022. This chapter keeps proof-of-work
because it makes the cost of rewriting history visible in a small R example.

**Transactions and fees**

Transactions are the financial content of blocks. A pending transaction has a
sender, receiver, amount, and fee. The fee compensates the block producer for
including the transaction.

In this teaching model, rejection has three sources: insufficient sender balance,
invalid amount or fee, and limited block capacity. The first is an accounting
failure, the second is a rule failure, and the third is a scarcity problem.

```{r}
pending_transactions <- tibble(
  tx_id = c("tx1", "tx2", "tx3", "tx4", "tx5"),
  sender = c("Alice", "Bob", "Carol", "Dave", "Alice"),
  receiver = c("Bob", "Carol", "Alice", "Bob", "Carol"),
  amount = c(25, 20, 12, 12, 70),
  fee = c(0.40, 0.15, 0.80, 0.30, 0.60)
)

pending_transactions
```

In many blockchain systems, transactions waiting for inclusion sit in a pending
pool. A block producer may prefer transactions with higher fees, subject to
validity and block capacity. If the block can include $m$ transactions, a
simple selection rule is:

$$
T^* = \operatorname{top}_m(\mathrm{fee}_j)
\quad \mathrm{among\ valid\ pending\ transactions}.
$$

The teaching function below processes transactions in descending fee order and
updates balances sequentially.

```{r}
select_for_block <- function(pool, balances, block_size = 3,
                             miner = "Miner") {
  ordered_pool <- arrange(pool, desc(fee))
  accepted <- ordered_pool[0, ]
  rejected <- ordered_pool[0, ]
  balances_after <- balances

  for (k in seq_len(nrow(ordered_pool))) {
    tx <- ordered_pool[k, ]

    if (nrow(accepted) >= block_size) {
      rejected <- bind_rows(rejected, tx)
      next
    }

    sender_balance <- balances_after$balance[
      balances_after$wallet == tx$sender
    ]

    valid <- length(sender_balance) == 1 &&
      sender_balance >= tx$amount + tx$fee &&
      tx$amount > 0 &&
      tx$fee >= 0

    if (valid) {
      accepted <- bind_rows(accepted, tx)
      balances_after$balance[balances_after$wallet == tx$sender] <-
        balances_after$balance[balances_after$wallet == tx$sender] -
        tx$amount - tx$fee
      balances_after$balance[balances_after$wallet == tx$receiver] <-
        balances_after$balance[balances_after$wallet == tx$receiver] +
        tx$amount
      balances_after$balance[balances_after$wallet == miner] <-
        balances_after$balance[balances_after$wallet == miner] + tx$fee
    } else {
      rejected <- bind_rows(rejected, tx)
    }
  }

  list(accepted = accepted, rejected = rejected, balances = balances_after)
}

block_selection <- select_for_block(
  pending_transactions,
  balances,
  block_size = 3,
  miner = "Miner"
)

block_decision <- bind_rows(
  mutate(block_selection$accepted, status = "Accepted"),
  mutate(block_selection$rejected, status = "Rejected")
) |>
  arrange(desc(fee))

block_decision
block_selection$balances
```

The first output records the inclusion decision; the second output records the
resulting balances. The accepted transactions are valid under the balance rule
and fit inside the block capacity. The rejected transactions fail validation or
remain outside the limited block space. The economic intuition is already
visible: users pay for inclusion, and block producers respond to those
incentives.

```{r}
#| label: fig-fee-ranking
#| fig-cap: "Fee ranking and block inclusion in a small pending pool."
#| fig-width: 7
#| fig-height: 3.2
fee_competition <- pending_transactions |>
  mutate(
    status = if_else(tx_id %in% block_selection$accepted$tx_id,
                     "Included in block", "Left out or invalid"),
    tx_id = reorder(tx_id, fee)
  )

ggplot(fee_competition, aes(x = tx_id, y = fee, fill = status)) +
  geom_col(width = 0.65) +
  coord_flip() +
  scale_fill_manual(values = c("Included in block" = "#2E86C1",
                               "Left out or invalid" = "#BFC9CA")) +
  labs(x = "Pending transaction", y = "Fee", fill = "")
```

This turns block construction into a market for scarce inclusion. The fee is
small relative to the transfer amount, but it can decide timing when many
transactions compete for limited capacity. Pending transactions, fees, and
inclusion delays therefore reveal pressure on the settlement layer.

```{r}
#| label: fig-wallet-balances
#| fig-cap: "Wallet balances before and after the mined block."
balance_plot <- bind_rows(
  mutate(balances, state = "Before block"),
  mutate(block_selection$balances, state = "After block")
)

ggplot(balance_plot, aes(x = wallet, y = balance, fill = state)) +
  geom_col(position = "dodge", width = 0.68) +
  scale_fill_manual(values = c("Before block" = "#2C7FB8",
                               "After block" = "#F39C12")) +
  labs(x = "Wallet", y = "Balance", fill = "")
```

The balance plot connects block inclusion to state change. Senders lose the
amount plus fee, receivers gain the transferred amount, and the next formula
isolates the miner's fee reward in this simplified example:

$$
\mathrm{Reward}_{\mathrm{fees}}
=\sum_{j\in T^*} \mathrm{fee}_j.
$$

```{r}
fee_reward <- sum(block_selection$accepted$fee)

tibble(component = "Transaction fees", reward = fee_reward)
```

In Bitcoin, miners can also receive a block subsidy according to the issuance
schedule. In this teaching account-based model, only transaction fees are
shown.

**Mining a transaction block**

The accepted transactions can now become block data. The block hash commits to
the transaction set, the parent hash, the timestamp, and the nonce.

```{r}
candidate_block <- list(
  number = 4,
  timestamp = "2026-01-01 09:03:00",
  data = block_selection$accepted,
  parent_hash = block3$hash
)

pow_result <- proof_of_work(candidate_block, difficulty = 2)
candidate_block$nonce <- pow_result$nonce
candidate_block$hash <- pow_result$hash

transaction_chain <- c(chain, list(candidate_block))

mined_block_summary <- tibble(
  block = candidate_block$number,
  transactions = nrow(candidate_block$data),
  nonce = candidate_block$nonce,
  parent_hash = substr(candidate_block$parent_hash, 1, 12),
  block_hash = substr(candidate_block$hash, 1, 12),
  valid_chain = validate_chain(transaction_chain)
)

mined_block_summary
```

The summary keeps the mined block readable while preserving the pieces that
matter for validation: transaction count, nonce, parent hash, block hash, and
chain status.

Changing a transaction after the block is mined breaks the chain.

```{r}
altered_transaction_chain <- transaction_chain
altered_transaction_chain[[4]]$data$amount[1] <- 250

tibble(check = "Altered mined transaction",
       valid = validate_chain(altered_transaction_chain))
```

This example connects pending transactions, balances, and the accepted ledger
state. The hash commits to the candidate block; validation rules, signatures,
balance checks, consensus, and network acceptance determine which transactions
enter the ledger.

It also separates two views of the same system. The transaction view asks which
transfers are accepted. The state view asks what the balances are after those
transfers. Financial applications need both: history for auditability and state
for valuation, reporting, and risk controls.

**Digital signatures**

A transaction also needs authorization. The ledger can define valid balances,
fees, and block rules, but it still needs evidence that the owner of the value
approved the transfer. Public-key cryptography supplies that authorization
layer.

The basic object is a key pair:

$$
(\mathrm{sk}, \mathrm{pk}),
$$

where $\mathrm{sk}$ is the private key and $\mathrm{pk}$ is the public key. The
private key must be kept secret. The public key can be shared because it is used
for verification. A simplified signature workflow is:

$$
m = \mathrm{serialize}(\mathrm{transaction}),
\qquad
h = H(m),
\qquad
\sigma = \mathrm{Sign}_{\mathrm{sk}}(h),
$$

and verification checks:

$$
\mathrm{Verify}_{\mathrm{pk}}(h,\sigma)\in\{\mathrm{TRUE},\mathrm{FALSE}\}.
$$

This gives two properties that matter for a financial ledger:

* authenticity: the signature matches the public key associated with the
  spending authority;
* integrity: changing the signed message invalidates the signature.

The ledger question is narrow and powerful: did the holder of the relevant
private key authorize this exact transaction message? That question is smaller
than legal identity, investor eligibility, tax treatment, or the existence of an
off-chain asset. Those issues require identity, legal, operational, and
governance layers around the cryptographic system.

The code below should be read as one security workflow:

* create a key pair;
* sign an exact transaction message;
* verify the signature with the public key;
* alter the message and see verification fail;
* separate signature logic from confidentiality.

The next example uses RSA keys through the <tt>`openssl`</tt> package. This is
a teaching choice. Bitcoin signatures use elliptic-curve schemes, including
ECDSA over secp256k1 and Schnorr signatures for Taproot outputs. The teaching
idea is the same: one key signs, the corresponding public key verifies, and any
change to the signed message breaks verification.

```{r}
alice_key <- rsa_keygen(bits = 2048)
alice_public_key <- alice_key$pubkey

key_roles <- tibble(
  object = c("Private key", "Public key"),
  kept_secret = c(TRUE, FALSE),
  can_sign = c(TRUE, FALSE),
  can_verify = c(TRUE, TRUE),
  financial_role = c(
    "Controls spending authority",
    "Allows others to verify authorization"
  )
)

key_roles
```

The private key is the sensitive asset. The public key can travel through the
network because it is used to check signatures. A wallet, custodian, or
exchange may manage this key material for a user, but the economic implication
is the same: control of the private key controls the ability to authorize
spending.

Now create a transaction message and sign it.

```{r}
signed_transaction <- list(
  sender = "Alice",
  receiver = "Bob",
  amount = 10,
  fee = 0.1
)

serialized_transaction <- serialize(signed_transaction, NULL)
transaction_digest <- sha256(serialized_transaction)
signature <- signature_create(serialized_transaction, sha256, key = alice_key)

tibble(
  object = c("Serialized transaction", "SHA-256 digest", "Digital signature"),
  bytes = c(length(serialized_transaction),
            length(transaction_digest),
            length(signature))
)
```

The signature is a compact object attached to the transaction data. Anyone with
Alice's public key can verify the signature.

```{r}
signature_valid <- signature_verify(
  serialized_transaction,
  signature,
  sha256,
  pubkey = alice_public_key
)

tibble(message = "Original signed transaction",
       valid_signature = signature_valid)
```

Now alter the transaction and verify the old signature against the altered
message. The amount and fee are unchanged, but the receiver changes from Bob to
Carol.

```{r}
altered_signed_transaction <- list(
  sender = "Alice",
  receiver = "Carol",
  amount = 10,
  fee = 0.1
)

altered_serialized_transaction <- serialize(altered_signed_transaction, NULL)

altered_signature_valid <- tryCatch(
  signature_verify(
    altered_serialized_transaction,
    signature,
    sha256,
    pubkey = alice_public_key
  ),
  error = function(e) FALSE
)

tibble(message = "Altered receiver",
       valid_signature = altered_signature_valid)
```

The altered message fails verification. This is why signatures are part of the
financial logic of a blockchain: the ledger should only accept transactions
authorized by the holder of the relevant private key.

**Confidentiality and encryption**

Signatures and encryption solve different problems. A signature answers who
authorized a message and whether the message changed. Encryption hides message
contents from parties that do not hold the decryption key.

In a public-key encryption system, a sender can encrypt a message with the
receiver's public key. The receiver decrypts it with the receiver's private
key:

$$
c = \mathrm{Enc}_{\mathrm{pk}_B}(m),
\qquad
m = \mathrm{Dec}_{\mathrm{sk}_B}(c).
$$

Bitcoin transactions are usually public enough for network validation, so
ordinary transaction data is visible to nodes. The encryption example below is
therefore a teaching example for confidentiality. It clarifies why a private
key is valuable: the key can authorize messages through signatures and can
decrypt messages that were encrypted to the matching public key.

```{r}
bob_key <- rsa_keygen(bits = 2048)
bob_public_key <- bob_key$pubkey

payment_note <- charToRaw("Alice pays Bob 10 with fee 0.1")

cipher_text <- rsa_encrypt(payment_note, bob_public_key)
recovered_note <- rsa_decrypt(cipher_text, bob_key)

tibble(
  object = c("Plain message", "Encrypted message", "Recovered message"),
  bytes = c(length(payment_note), length(cipher_text), length(recovered_note)),
  readable = c(rawToChar(payment_note), "<cipher text>", rawToChar(recovered_note))
)
```

The encrypted message is longer and unreadable as ordinary text. The receiver's
private key recovers the original message.

```{r}
tibble(check = "Recovered message matches original",
       valid = identical(payment_note, recovered_note))
```

In production systems, large messages are commonly encrypted with a symmetric
key, and the symmetric key is protected with public-key encryption. The small
RSA example is still useful because it separates three ideas that are often
mixed together:

| Object | Main question | Financial meaning |
|---|---|---|
| Hash | Did the data change? | Integrity and compact commitment |
| Signature | Who authorized this exact data? | Spending authority and non-repudiation in the ledger model |
| Encryption | Who can read the data? | Confidentiality and access control |
| Private-key custody | Who can use the secret key? | Operational control over value |

This separation matters for blockchain finance. A public ledger can allow
verification while still creating privacy problems. A signature can prove that
a key authorized a transaction while leaving the real-world owner of the key
uncertain. Encryption can protect message contents while leaving metadata,
timing, amounts, or network relationships exposed.

**Custody and key risk**

Private-key custody turns cryptography into a financial risk. If the private
key is stolen, an attacker can authorize transfers. If the private key is lost,
the owner may lose the ability to spend. If a custodian controls the private
key, the user faces operational, legal, and counterparty risk. These risks are
different from market risk, but they can dominate the financial outcome.

```{r}
custody_scenarios <- tibble(
  scenario = c(
    "Self-custody",
    "Lost private key",
    "Stolen private key",
    "Exchange custody",
    "Two-of-three multisignature"
  ),
  who_can_sign = c(
    "The holder of the wallet key",
    "Nobody, if the key cannot be recovered",
    "The attacker",
    "The exchange or custodian",
    "Any two authorized key holders"
  ),
  main_financial_risk = c(
    "Operational mistakes by the owner",
    "Permanent loss of spending ability",
    "Unauthorized transfer",
    "Counterparty, legal, and operational failure",
    "Coordination failure or compromised signers"
  )
)

custody_scenarios
```

Multisignature arrangements are one way to reduce single-key failure. A
two-of-three setup can require any two authorized keys to sign before funds
move. That design changes the operational problem from protecting one secret to
coordinating several secrets. It can reduce theft and loss risk, but it adds
governance and recovery procedures.

For this chapter, the important point is that security is part of the financial
model. The ledger state says who controls value according to protocol rules.
The signature says which key authorized a transaction. Custody determines who
can use that key in practice. Privacy determines how much of the transfer
history can be linked to people, institutions, or strategies.

**Network and consensus**

A blockchain ledger is distributed over a peer-to-peer network. A simplified
proof-of-work workflow is:

* users broadcast transactions;
* nodes check transaction validity;
* miners gather valid pending transactions into candidate blocks;
* miners search for a proof-of-work;
* the first valid block is broadcast;
* nodes accept the block if the proof and the transactions are valid;
* new blocks build on the accepted parent hash.

The same workflow can be represented as a graph. Let $a_{ij}=1$ when a message
can move from node $i$ to node $j$, and $a_{ij}=0$ otherwise. A simple measure
of local connectivity is the total degree:

$$
d_i = \sum_j a_{ij} + \sum_j a_{ji}.
$$

In the next figure, node size is proportional to $d_i$. Wallets are peripheral,
while relay, full-node, and mining nodes sit closer to the center of message
propagation.

```{r}
#| label: fig-network-flow
#| fig-cap: "A simplified peer-to-peer blockchain network."
#| fig-width: 8.6
#| fig-height: 4.4
network_nodes <- tibble(
  id = c("wallet_a", "wallet_b", "wallet_c", "exchange",
         "relay_1", "relay_2", "full_1", "full_2", "full_3",
         "miner_1", "miner_2", "chain"),
  label = c("Wallet A", "Wallet B", "Wallet C", "Exchange\nnode",
            "Relay\nnode 1", "Relay\nnode 2", "Full\nnode 1",
            "Full\nnode 2", "Full\nnode 3", "Miner 1", "Miner 2",
            "Accepted\nchain"),
  type = c("Wallet", "Wallet", "Wallet", "Service",
           "Relay", "Relay", "Full node", "Full node", "Full node",
           "Miner", "Miner", "Ledger"),
  x = c(0.2, 0.6, 0.2, 0.9, 2.5, 3.0, 4.3, 4.6, 4.0, 6.0, 6.3, 7.9),
  y = c(3.4, 2.2, 1.0, 4.4, 3.2, 1.5, 4.0, 2.4, 0.8, 3.5, 1.4, 2.4)
)

network_edges <- tibble(
  from = c("wallet_a", "wallet_b", "wallet_c", "exchange",
           "exchange", "relay_1", "relay_1", "relay_1", "relay_2",
           "relay_2", "full_1", "full_2", "full_3", "full_1",
           "full_2", "full_3", "miner_1", "miner_2", "miner_1",
           "miner_2", "relay_2"),
  to = c("relay_1", "relay_1", "full_3", "relay_2",
         "full_2", "full_1", "full_2", "miner_1", "full_2",
         "miner_2", "relay_2", "relay_1", "relay_2", "chain",
         "chain", "chain", "full_1", "full_2", "chain",
         "chain", "full_3"),
  relation = c("Transaction relay", "Transaction relay",
               "Transaction relay", "Transaction relay",
               "Transaction relay", "Peer relay", "Peer relay",
               "Candidate block", "Peer relay", "Candidate block",
               "Peer relay", "Peer relay", "Peer relay",
               "Validation state", "Validation state",
               "Validation state", "Block relay", "Block relay",
               "Validation state", "Validation state", "Peer relay"),
  weight = c(1.0, 1.0, 0.7, 1.1, 0.8, 0.9, 1.0, 1.4, 0.9, 1.3,
             0.8, 0.8, 0.7, 1.2, 1.2, 1.0, 1.3, 1.3, 1.5, 1.5, 0.7)
)

network_graph <- igraph::graph_from_data_frame(
  d = network_edges,
  directed = TRUE,
  vertices = network_nodes
)

igraph::V(network_graph)$degree <- igraph::degree(network_graph, mode = "all")

network_tbl_graph <- tidygraph::as_tbl_graph(network_graph)

ggraph(
  network_tbl_graph,
  layout = "manual",
  x = igraph::V(network_graph)$x,
  y = igraph::V(network_graph)$y
) +
  geom_edge_link(
    aes(edge_width = weight, edge_colour = relation),
    arrow = grid::arrow(length = grid::unit(0.11, "inches")),
    end_cap = circle(5, "mm"),
    start_cap = circle(5, "mm"),
    alpha = 0.8
  ) +
  geom_node_point(
    aes(size = degree, fill = type),
    shape = 21, color = "white", stroke = 1.1
  ) +
  geom_node_text(
    aes(label = label),
    size = 2.8, lineheight = 0.9
  ) +
  scale_edge_width(range = c(0.35, 1.9), guide = "none") +
  scale_edge_colour_manual(values = c(
    "Transaction relay" = "#2874A6",
    "Peer relay" = "#566573",
    "Candidate block" = "#B03A2E",
    "Block relay" = "#196F3D",
    "Validation state" = "#B7950B"
  )) +
  scale_fill_manual(values = c(
    "Wallet" = "#D6EAF8",
    "Service" = "#FCF3CF",
    "Relay" = "#EAECEE",
    "Full node" = "#D5F5E3",
    "Miner" = "#FADBD8",
    "Ledger" = "#E8DAEF"
  )) +
  scale_size(range = c(8, 18), guide = "none") +
  coord_cartesian(xlim = c(-0.25, 8.3), ylim = c(0.35, 4.8),
                  expand = FALSE, clip = "off") +
  guides(
    edge_colour = guide_legend(title = "Message"),
    fill = guide_legend(title = "Node type")
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    plot.margin = margin(4, 6, 4, 6)
  )
```

The diagram shows how ledger validity depends on message propagation. Wallets
and service nodes inject transactions from the edge of the graph. Relay and
full nodes move messages across the network. Miners receive candidate
transactions, propose blocks, and broadcast block information back to
validating nodes. The accepted chain is the history that nodes are willing to
extend after verification.

The useful part of this graph is the topology. The same transaction can reach
the accepted ledger through several paths, and nodes with more connections have
more influence over propagation. A valid transaction still has to move through
a communication layer before it becomes part of the history that participants
observe.

Consensus is the rule that tells nodes which valid history to extend. In
Bitcoin, the accepted history is tied to accumulated proof-of-work. The economic
security argument is that rewriting history requires controlling enough
computational work to replace the accepted chain, which is costly.

Incentives complete the mechanism. Fees and block rewards encourage miners to
spend resources on valid blocks. Invalid blocks waste effort because nodes
reject them.

The generic lab now has enough pieces to read Bitcoin more carefully:

| Toy mechanism | Bitcoin object | Financial reading |
|---|---|---|
| Account balance table | UTXO set reconstructed from unspent outputs | Spendable value controlled by keys |
| Toy transaction | Signed transaction with inputs, outputs, and fee | Payment instruction and settlement cost |
| Block with parent hash | Block header linked to the previous block | Commitment to accepted history |
| Toy proof-of-work difficulty | Network difficulty target | Cost of rewriting recent settlement |
| Fee reward | Fees plus protocol-defined subsidy | Miner incentive and block-space allocation |
| Peer-to-peer message graph | Wallets, relays, full nodes, miners, and services | Propagation, validation, and observability |

## Bitcoin case: a proof-of-work ledger

With that map in place, Bitcoin can be read as a specific monetary and payment
system built from ledger, hash, signature, network, and proof-of-work pieces.
Its design connects payments, scarce issuance, public verification, private-key
custody, and costly history rewriting [@nakamoto2008bitcoin].

The generic model and Bitcoin share the same broad workflow:

$$
\mathrm{signed\ transaction}
\rightarrow
\mathrm{pending\ transaction\ set}
\rightarrow
\mathrm{candidate\ block}
\rightarrow
\mathrm{proof\ of\ work}
\rightarrow
\mathrm{network\ validation}
\rightarrow
\mathrm{accepted\ ledger\ history}.
$$

The bridge above gives the correspondence. The main simplifications are:

| Generic model in this chapter | Bitcoin case |
|---|---|
| Account balances are stored directly for teaching clarity | Balances are inferred from unspent transaction outputs |
| Transactions move amounts from sender to receiver | Transactions consume previous outputs and create new outputs |
| The miner receives transaction fees in the teaching balance table | Miners can receive fees and a protocol-defined block subsidy |
| Difficulty is a small number of leading zeros | Difficulty adjusts so blocks arrive near the protocol target interval |
| Finality is shown as a valid chain in R | Settlement confidence grows as more blocks build on top |

A table also helps separate protocol roles from market access roles:

| Role | Protocol or market function | Financial reading |
|---|---|---|
| Wallet | Creates and signs transactions | Custody and spending authority |
| Mempool | Holds valid pending transactions before block inclusion | Fee pressure and settlement delay |
| Miner | Selects transactions and searches for proof-of-work | Incentives and block-space allocation |
| Full node | Verifies blocks and transactions against consensus rules | Independent validation |
| Accepted chain | Stores the history participants extend | Settlement reference |
| Exchange or custodian | Provides market access and often manages keys | Counterparty and operational risk |
| Explorer or analytics tool | Makes public transaction data searchable | Monitoring, attribution, and compliance |

This table separates protocol roles from financial access roles. A price, a
wallet balance, a transaction status, and an exchange account are produced by
different parts of the system. Treating them as one object makes Bitcoin look
flatter than it is.

A Bitcoin transaction also has an operational path:

| Stage | Mechanism | Financial consequence |
|---|---|---|
| Wallet construction | The wallet selects spendable outputs, creates new outputs, and signs the transaction | Spending authority and fee choice |
| Mempool relay | Nodes hold valid pending transactions according to local policy | Pending settlement and public fee pressure |
| Block construction | Miners select transactions subject to block weight and fee incentives | Scarce block space |
| Coinbase transaction | The miner includes a special transaction that pays the subsidy and collected fees | Miner revenue and issuance rule |
| Confirmation | The transaction is included in a block, then new blocks build above it | Increasing settlement confidence |

Fee-rate thinking is useful because block space is measured by transaction size
as well as by amount paid. A compact model is:

$$
\mathrm{FeeRate}_j =
\frac{\mathrm{Fee}_j}{\mathrm{VirtualBytes}_j}.
$$

Miners compare valid pending transactions under block-space constraints, so a
large transaction may need a larger absolute fee to offer the same fee rate as
a smaller transaction. The miner's block revenue combines the protocol-defined
subsidy and the fees from included transactions:

$$
\mathrm{MinerRevenue}_t =
\mathrm{Subsidy}_t + \sum_{j\in B_t}\mathrm{Fee}_j.
$$

The subsidy follows the issuance schedule, while the fee component changes with
demand for block space. This is why mempool conditions, fee rates, and
confirmation depth are financial variables, even before looking at the market
price of bitcoin.

The account-based model says that Alice has a balance and can send part of it.
Bitcoin uses a UTXO model. A UTXO is an unspent transaction output. If Alice
controls previous outputs worth 0.8 BTC and 0.4 BTC, and wants to pay 0.9 BTC,
her transaction spends enough previous outputs and creates new outputs: one for
the receiver, one for change back to Alice, and one implicit fee to the miner.
The balance is therefore reconstructed from the set of unspent outputs
controlled by Alice's keys.

The simplified accounting equation is:

$$
\sum \mathrm{Inputs} = \sum \mathrm{Outputs} + \mathrm{Fee}.
$$

The fee is the difference between the value consumed by the transaction and the
value assigned to new outputs:

$$
\mathrm{Fee} = \sum \mathrm{Inputs} - \sum \mathrm{Outputs}.
$$

In the account example above, the fee was an explicit column. Here the fee is
the unassigned remainder after inputs and outputs are compared. The economic
meaning is similar: the sender gives up value, the receiver obtains value, and
the miner has an incentive to include the transaction.

The same example can be read as a small UTXO table:

| Side | Reference | Amount (BTC) | Controlled by | Role |
|---|---:|---:|---|---|
| Input | Previous output 1 | 0.80 | Alice key | Spent source of value |
| Input | Previous output 2 | 0.40 | Alice key | Spent source of value |
| Output | New output 1 | 0.90 | Bob key | Payment to receiver |
| Output | New output 2 | 0.28 | Alice key | Change back to sender |
| Fee | Implicit remainder | 0.02 | Miner | Incentive for inclusion |

The table balances because inputs of 1.20 BTC fund outputs of 1.18 BTC and an
implicit fee of 0.02 BTC. After the transaction is accepted, the two previous
outputs are spent, and only the new outputs remain available for future
transactions.

**Bitcoin transactions as a network**

A UTXO ledger can also be read as a network. Nodes represent wallets,
addresses, services, or output clusters. Directed edges represent value moving
from one node to another. The edge weight can be the BTC amount. This view is
useful because it changes the question from a single payment to a system of
relationships: which nodes concentrate flow, which nodes mostly receive, which
nodes mostly distribute, and where fees or change outputs appear.

The next example uses a small synthetic transaction network. It is not meant to
identify real addresses. Its purpose is to show how network tools can turn
ledger data into a visual model. This is the point where the chapter moves from
blockchain as a validation mechanism to blockchain as analyzable financial
data.

```{r}
#| label: fig-utxo-transaction-network
#| fig-cap: "A synthetic UTXO transaction network with BTC-weighted flows."
#| fig-width: 8.4
#| fig-height: 4.8
utxo_nodes <- tibble(
  name = c("coinbase", "miner", "exchange", "custodian", "alice", "bob",
           "merchant", "change_a", "cold_storage", "fee_pool"),
  label = c("Coinbase\nreward", "Miner", "Exchange\nhot wallet",
            "Custodian", "Alice\nwallet", "Bob\nwallet",
            "Merchant", "Change\naddress A", "Cold\nstorage",
            "Fee\npool"),
  type = c("Protocol", "Miner", "Custody", "Custody", "Wallet", "Wallet",
           "Merchant", "Change", "Storage", "Fees"),
  x = c(0.3, 1.6, 3.0, 4.4, 5.6, 5.6, 7.0, 7.0, 8.2, 3.0),
  y = c(2.9, 2.9, 2.1, 2.9, 3.4, 1.8, 2.7, 1.1, 1.6, 0.8)
)

utxo_edges <- tibble(
  from = c("coinbase", "miner", "miner", "exchange", "exchange",
           "exchange", "custodian", "alice", "alice", "bob", "bob",
           "merchant", "change_a", "exchange"),
  to = c("miner", "exchange", "fee_pool", "alice", "bob",
         "custodian", "exchange", "merchant", "change_a", "merchant",
         "fee_pool", "exchange", "cold_storage", "fee_pool"),
  btc = c(6.25, 2.40, 0.06, 1.25, 0.80, 1.10, 0.75,
          0.42, 0.81, 0.20, 0.02, 0.55, 0.60, 0.04),
  role = c("Block reward", "Distribution", "Fee", "Withdrawal",
           "Withdrawal", "Custody", "Rebalance", "Payment", "Change",
           "Payment", "Fee", "Settlement", "Savings", "Fee")
)

utxo_graph <- igraph::graph_from_data_frame(
  d = utxo_edges,
  directed = TRUE,
  vertices = utxo_nodes
)

igraph::V(utxo_graph)$in_value <- igraph::strength(
  utxo_graph,
  mode = "in",
  weights = igraph::E(utxo_graph)$btc
)
igraph::V(utxo_graph)$out_value <- igraph::strength(
  utxo_graph,
  mode = "out",
  weights = igraph::E(utxo_graph)$btc
)
igraph::V(utxo_graph)$total_value <-
  igraph::V(utxo_graph)$in_value + igraph::V(utxo_graph)$out_value
igraph::V(utxo_graph)$degree <- igraph::degree(utxo_graph, mode = "all")

utxo_tbl_graph <- tidygraph::as_tbl_graph(utxo_graph)

ggraph(
  utxo_tbl_graph,
  layout = "manual",
  x = igraph::V(utxo_graph)$x,
  y = igraph::V(utxo_graph)$y
) +
  geom_edge_link(
    aes(edge_width = btc, edge_colour = role),
    arrow = grid::arrow(length = grid::unit(0.12, "inches")),
    end_cap = circle(5, "mm"),
    start_cap = circle(5, "mm"),
    alpha = 0.82
  ) +
  geom_node_point(
    aes(size = total_value, fill = type),
    shape = 21, color = "white", stroke = 1.1
  ) +
  geom_node_text(
    aes(label = label),
    size = 2.75, lineheight = 0.86
  ) +
  scale_edge_width(range = c(0.35, 2.7), guide = "none") +
  scale_edge_colour_manual(values = c(
    "Block reward" = "#117A65",
    "Distribution" = "#2874A6",
    "Fee" = "#922B21",
    "Withdrawal" = "#7D3C98",
    "Custody" = "#CA6F1E",
    "Rebalance" = "#566573",
    "Payment" = "#1F618D",
    "Change" = "#85929E",
    "Settlement" = "#196F3D",
    "Savings" = "#B7950B"
  )) +
  scale_fill_manual(values = c(
    "Protocol" = "#D6EAF8",
    "Miner" = "#FADBD8",
    "Custody" = "#FDEBD0",
    "Wallet" = "#D5F5E3",
    "Merchant" = "#E8DAEF",
    "Change" = "#EAECEE",
    "Storage" = "#FCF3CF",
    "Fees" = "#F5B7B1"
  )) +
  scale_size(range = c(8, 22), guide = "none") +
  coord_cartesian(xlim = c(-0.2, 8.8), ylim = c(0.35, 3.75),
                  expand = FALSE, clip = "off") +
  guides(edge_colour = guide_legend(title = "Edge role"),
         fill = guide_legend(title = "Node type")) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    plot.margin = margin(6, 6, 6, 6)
  )
```

The network view carries a different kind of explanation. The exchange and the
miner appear as high-flow nodes because several edges pass through them. Alice
creates both a payment edge and a change edge. The fee pool collects small
edges that would be easy to miss in a table. A merchant receives payments and
later settles part of the balance back to an exchange. These are financial
roles, even in a synthetic network.

This figure should be read as an accounting map. The edge width shows value,
the edge color shows transaction role, and the node size shows total flow
handled by each participant. It makes three concepts visible at once: issuance
through the coinbase reward, payments through wallet-to-merchant edges, and
operational concentration through the exchange and miner nodes.

We can summarize the same graph numerically. Let $w_{ij}$ be the BTC value sent
from node $i$ to node $j$. The incoming and outgoing value of node $i$ are:

$$
\mathrm{InValue}_i = \sum_j w_{ji},
\qquad
\mathrm{OutValue}_i = \sum_j w_{ij}.
$$

```{r}
network_node_summary <- igraph::as_data_frame(utxo_graph, what = "vertices") |>
  as_tibble() |>
  transmute(
    node = gsub("\n", " ", label),
    type,
    in_value = round(in_value, 2),
    out_value = round(out_value, 2),
    total_value = round(total_value, 2),
    degree
  ) |>
  arrange(desc(total_value))

network_node_summary
```

```{r}
#| label: fig-utxo-node-flow
#| fig-cap: "Incoming plus outgoing BTC by node in the synthetic UTXO network."
#| fig-width: 7.4
#| fig-height: 3.6
network_node_summary |>
  mutate(node = reorder(node, total_value)) |>
  ggplot(aes(x = node, y = total_value, fill = type)) +
  geom_col(width = 0.68) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Protocol" = "#D6EAF8",
    "Miner" = "#FADBD8",
    "Custody" = "#FDEBD0",
    "Wallet" = "#D5F5E3",
    "Merchant" = "#E8DAEF",
    "Change" = "#EAECEE",
    "Storage" = "#FCF3CF",
    "Fees" = "#F5B7B1"
  )) +
  labs(x = "Node", y = "Incoming plus outgoing BTC", fill = "")
```

This is where network analysis becomes useful for financial interpretation.
Large nodes can indicate custody, market access, mining income, payments, or
operational concentration. The meaning depends on the node label and the
transaction role. A high-flow exchange address has a different interpretation
from a high-flow merchant or miner address.

The bar chart forces the visual claim into a simple metric. In the synthetic
data, the miner and the exchange dominate total observed flow, while the fee
pool is small in value but important for incentives. A real blockchain dataset
would require address clustering, time windows, filtering rules, and careful
labeling, but the modeling question is the same: which parts of the transaction
graph carry the most economic activity?

Bitcoin also makes proof-of-work part of monetary security. A miner proposes a
block and searches for a valid proof. Nodes verify the block by checking the
proof, transaction validity, and the rule that spent outputs have not already
been spent. A transaction with one confirmation is included in one block. More
confirmations mean that additional blocks have been built on top of that block,
raising the cost of replacing that history.

The settlement idea is probabilistic. A payment becomes harder to reverse as
more proof-of-work accumulates above it. This is why confirmations matter in
Bitcoin payments: they are a practical measure of how deeply a transaction sits
inside the accepted history.

One simplified way to visualize the intuition is a catch-up race. Suppose an
attacker controls share $q$ of mining power, with $0<q<0.5$, and an honest
chain is already $z$ confirmations ahead. A compact approximation for the
probability that the attacker catches up is:

$$
\Pr(\mathrm{catchUp}\mid z,q)
\approx
\left(\frac{q}{1-q}\right)^z.
$$

The approximation is a teaching device. It captures the central shape: when the
attacker has less than half of total work, each additional confirmation lowers
the catch-up probability.

```{r}
#| label: fig-confirmation-risk
#| fig-cap: "Approximate catch-up probability by confirmations and attacker share."
#| fig-width: 7.2
#| fig-height: 3.5
confirmation_grid <- as_tibble(expand.grid(
  confirmations = 1:10,
  attacker_share = c(0.10, 0.20, 0.30, 0.40)
)) |>
  mutate(
    catch_up_probability =
      (attacker_share / (1 - attacker_share))^confirmations,
    attacker_label = paste0("q = ",
                            scales::percent(attacker_share, accuracy = 1))
  )

ggplot(
  confirmation_grid,
  aes(x = confirmations, y = catch_up_probability, color = attacker_label)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:10) +
  scale_y_log10(labels = scales::label_scientific(digits = 2)) +
  labs(x = "Confirmations",
       y = "Catch-up probability (log scale)", color = "")
```

The logarithmic scale makes the settlement intuition visible. Confirmations
translate settlement depth into a practical risk measure: they express the cost
of replacing recent history. The same payment can be operationally acceptable
at different confirmation depths depending on amount, counterparty risk,
liquidity needs, and the recipient's tolerance for reversal risk.

Bitcoin is financially interesting for four reasons.

First, it turns ledger maintenance into an incentive problem. Miners spend
resources because accepted blocks can pay rewards and fees. Second, it makes
custody explicit: control of private keys is control over spending authority.
Third, it creates a public audit trail, so anyone can verify the transaction
history according to protocol rules. Fourth, it introduces a digital asset with
a rule-based issuance schedule, which makes scarcity part of the protocol
design.

The R model above should be read as a simplified map. It captures blocks,
hashes, proof-of-work, fees, validation, and ledger updates. It leaves out many
Bitcoin details: UTXO selection algorithms, script rules, block propagation,
difficulty adjustment mechanics, mempool policy, fee estimation, wallet
software, and network topology. The simplification is deliberate. The goal is
to make the financial logic visible before adding production-level complexity.

## Tokenization

Tokenization extends the ledger idea from native coins to other financial or
economic claims. A token can represent a unit in a digital asset, a claim on a
cash-flow arrangement, a fund share, a loyalty point, a stablecoin balance, or
a record linked to an off-chain asset. The crucial point is that the token
record and the legal or operational claim have to be aligned.

The shift from Bitcoin to tokenization changes the financial question. In
Bitcoin, the native asset and the ledger are part of the same protocol. In
tokenization, the ledger records ownership units whose value often depends on
something outside the ledger: cash, securities, real assets, fund shares,
invoices, commodities, or contractual rights. That creates two layers:

| Layer | Question |
|---|---|
| On-chain record | Who holds how many tokens according to the ledger? |
| Off-chain claim | What legal, economic, or operational right does each token represent? |

The on-chain record can be precise while the off-chain claim still requires
legal documentation, custody, audits, servicing, redemption rules, and dispute
resolution. This is why tokenization is a finance topic as much as a technology
topic.

```{r}
#| label: fig-tokenization-layers
#| fig-cap: "Tokenization layers from off-chain asset to cash-flow rule."
#| fig-width: 8
#| fig-height: 1.1
token_layers <- tibble(
  id = c("asset", "legal", "issuer", "registry", "investors", "cashflow"),
  label = c("Underlying\nasset", "Legal\nclaim", "Issuer/\nservicer",
            "Token\nregistry", "Investors", "Cash-flow\nrule"),
  layer = c("Off-chain", "Off-chain", "Bridge",
            "On-chain", "Market", "Finance"),
  x = c(0.5, 1.9, 3.3, 4.7, 6.1, 7.5),
  y = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
)

token_edges <- tibble(
  from = c("asset", "legal", "issuer", "registry", "investors"),
  to = c("legal", "issuer", "registry", "investors", "cashflow")
) |>
  left_join(token_layers |> select(from = id, x, y), by = "from") |>
  rename(x_from = x, y_from = y) |>
  left_join(token_layers |> select(to = id, x, y), by = "to") |>
  rename(x_to = x, y_to = y) |>
  mutate(
    x_from_plot = x_from + 0.36,
    x_to_plot = x_to - 0.36
  )

ggplot() +
  geom_segment(
    data = token_edges,
    aes(x = x_from_plot, y = y_from, xend = x_to_plot, yend = y_to),
    arrow = grid::arrow(length = grid::unit(0.11, "inches")),
    color = "#7F8C8D", linewidth = 0.65
  ) +
  geom_point(
    data = token_layers,
    aes(x = x, y = y, fill = layer),
    shape = 21, size = 14, color = "white", stroke = 1.1
  ) +
  geom_text(
    data = token_layers,
    aes(x = x, y = y, label = label),
    size = 2.8, lineheight = 0.88
  ) +
  scale_fill_manual(values = c(
    "Off-chain" = "#D6EAF8",
    "Bridge" = "#FDEBD0",
    "On-chain" = "#D5F5E3",
    "Market" = "#E8DAEF",
    "Finance" = "#FCF3CF"
  )) +
  coord_cartesian(xlim = c(-0.1, 8.1), ylim = c(1.35, 1.65),
                  expand = FALSE, clip = "off") +
  guides(fill = "none") +
  theme_void() +
  theme(plot.margin = margin(8, 2, 8, 2))
```

This stack is where tokenization becomes a financial structure. The token
registry can say who holds units. Valuation and risk also depend on the asset,
legal claim, servicing process, transfer restrictions, and cash-flow rule. A
token model is incomplete unless the financial claim represented by the token
is explicit.

In a simple model, a token contract has:

* an issuer;
* a token name;
* total supply;
* holder balances;
* transfer rules;
* optional cash-flow or redemption rules.

Suppose a hypothetical issuer creates 1,000 tokens linked to participation in a
small financial asset pool.

```{r}
token_metadata <- tibble(
  token = "POOL_A_TOKEN",
  underlying = "Hypothetical asset-pool participation",
  total_supply = 1000,
  unit_claim = "1 token = 0.1% of the pool cash-flow rule"
)

token_register <- tibble(
  holder = c("Issuer", "Investor_A", "Investor_B", "Investor_C"),
  token_balance = c(1000, 0, 0, 0)
)

token_metadata
token_register
```

Token transfers are ledger updates. If holder $a$ transfers $q$ tokens to
holder $b$, the token register changes as:

$$
\mathrm{Token}_a^{\prime} = \mathrm{Token}_a-q,
\qquad
\mathrm{Token}_b^{\prime} = \mathrm{Token}_b+q,
\qquad
0 \leq q \leq \mathrm{Token}_a.
$$

```{r}
transfer_tokens <- function(register, from, to, amount) {
  from_balance <- register$token_balance[register$holder == from]

  valid <- length(from_balance) == 1 &&
    from_balance >= amount &&
    amount > 0 &&
    to %in% register$holder

  if (!valid) {
    stop("Invalid token transfer")
  }

  register$token_balance[register$holder == from] <-
    register$token_balance[register$holder == from] - amount
  register$token_balance[register$holder == to] <-
    register$token_balance[register$holder == to] + amount

  register
}

token_register_after <- token_register |>
  transfer_tokens("Issuer", "Investor_A", 400) |>
  transfer_tokens("Issuer", "Investor_B", 350) |>
  transfer_tokens("Issuer", "Investor_C", 150) |>
  transfer_tokens("Investor_A", "Investor_C", 50)

token_register_after
```

```{r}
#| label: fig-token-register
#| fig-cap: "Token ownership before and after transfers."
token_plot <- bind_rows(
  mutate(token_register, state = "Initial register"),
  mutate(token_register_after, state = "After transfers")
)

ggplot(token_plot, aes(x = holder, y = token_balance, fill = state)) +
  geom_col(position = "dodge", width = 0.68) +
  scale_fill_manual(values = c("Initial register" = "#566573",
                               "After transfers" = "#27AE60")) +
  labs(x = "Holder", y = "Token balance", fill = "")
```

The plot separates issuance from redistribution. The issuer begins with the
full supply, and later balances show which investors would receive cash flows
under a proportional allocation rule.

If the pool distributes cash flow $C$, a proportional token rule allocates:

$$
\mathrm{CashFlow}_h = \frac{\mathrm{Token}_h}{S}C,
$$

where $S$ is total token supply.

```{r}
cash_flow <- 5000

cash_distribution <- token_register_after |>
  mutate(
    ownership = token_balance / sum(token_balance),
    cash_flow = ownership * cash_flow
  )

cash_distribution
```

The same register can be connected to valuation. If the token represents a
proportional claim on expected cash flows $C_t$, and $r$ is the discount rate,
then holder $h$ has a model value:

$$
\mathrm{PV}_h =
\frac{\mathrm{Token}_h}{S}
\sum_{t=1}^T
\frac{\mathbb{E}[C_t]}{(1+r)^t}.
$$

This is a simple discounted cash-flow model. The blockchain register tells us
the token balance. The valuation model still needs expected cash flows,
discount rates, legal enforceability, and servicing assumptions.

```{r}
cash_flow_scenarios <- as_tibble(expand.grid(
  year = 1:4,
  scenario = c("Downside", "Base", "Upside")
)) |>
  mutate(
    expected_cash_flow = case_when(
      scenario == "Downside" ~ c(3200, 3400, 3500, 3600)[year],
      scenario == "Base" ~ c(4500, 4750, 5000, 5250)[year],
      TRUE ~ c(5600, 6100, 6500, 6900)[year]
    ),
    discount_rate = case_when(
      scenario == "Downside" ~ 0.14,
      scenario == "Base" ~ 0.10,
      TRUE ~ 0.08
    ),
    present_value = expected_cash_flow / (1 + discount_rate)^year
  )

scenario_value <- cash_flow_scenarios |>
  group_by(scenario, discount_rate) |>
  summarise(pool_value = sum(present_value), .groups = "drop") |>
  mutate(value_per_token = pool_value / token_metadata$total_supply)

holder_token_value <- token_register_after |>
  tidyr::crossing(scenario_value) |>
  mutate(holder_value = token_balance * value_per_token) |>
  arrange(scenario, desc(holder_value))

scenario_value
holder_token_value
```

```{r}
#| label: fig-token-scenario-value
#| fig-cap: "Scenario value of the tokenized cash-flow claim."
#| fig-width: 7.2
#| fig-height: 3.4
holder_token_value |>
  mutate(holder = reorder(holder, holder_value)) |>
  ggplot(aes(x = holder, y = holder_value, fill = scenario)) +
  geom_col(position = "dodge", width = 0.68) +
  coord_flip() +
  scale_fill_manual(values = c(
    "Downside" = "#BFC9CA",
    "Base" = "#2E86C1",
    "Upside" = "#27AE60"
  )) +
  labs(x = "Holder", y = "Modeled token value", fill = "")
```

The valuation figure is deliberately simple. It shows that a token register and
a valuation model answer different questions. The register answers who holds
units. The cash-flow model asks what those units may be worth under a set of
assumptions. Tokenization becomes financially meaningful when those two layers
are kept aligned.

This example shows why tokenization belongs in a financial modeling course. It
combines a ledger, ownership units, transfer rules, cash-flow allocation, and
valuation. The hard parts appear quickly: custody, investor rights, transfer
restrictions, identity checks, tax treatment, legal enforceability, valuation,
and operational controls.

## Privacy

Public blockchains make verification easier by exposing transaction history,
but that transparency creates privacy questions. A public address can be
pseudonymous, yet transaction patterns may still reveal relationships. If an
address is linked to a person or institution, past and future activity can
become easier to trace.

The banking model usually limits transaction information to the parties, the
bank, and regulators. A public blockchain makes more information visible to the
network. Privacy-preserving tools, address management, zero-knowledge proofs,
and permissioned ledgers are attempts to balance auditability and privacy.

For modeling purposes, privacy can be treated as a linkage problem. A public
address may start without a real-world name, but repeated use, counterparties,
service labels, timing, and transaction amounts can make the address easier to
classify. A simple exposure score for address $a$ can be written as:

$$
\mathrm{Exposure}_a =
\mathrm{Reuse}_a
+ \mathrm{Counterparties}_a
+ \mathrm{ServiceLabel}_a,
$$

where $\mathrm{ServiceLabel}_a$ equals one when the address has been linked to
a known service such as an exchange, custodian, merchant, or public donation
address.

```{r}
privacy_events <- tibble(
  address = c("addr_A", "addr_A", "addr_A", "addr_B", "addr_B",
              "addr_C", "addr_D", "addr_D", "addr_E"),
  counterparty = c("exchange_1", "merchant_1", "merchant_2", "wallet_x",
                   "wallet_y", "exchange_1", "donation_page",
                   "exchange_1", "wallet_z"),
  amount = c(0.40, 0.08, 0.12, 0.55, 0.20, 1.10, 0.03, 0.07, 0.18),
  service_label = c(TRUE, FALSE, FALSE, FALSE, FALSE,
                    TRUE, TRUE, TRUE, FALSE)
)

privacy_summary <- privacy_events |>
  group_by(address) |>
  summarise(
    reuse = n(),
    counterparties = n_distinct(counterparty),
    labeled_service = as.integer(any(service_label)),
    observed_amount = sum(amount),
    exposure_score = reuse + counterparties + labeled_service,
    .groups = "drop"
  ) |>
  arrange(desc(exposure_score), desc(observed_amount))

privacy_summary
```

```{r}
#| label: fig-privacy-exposure
#| fig-cap: "Address exposure score in a small public-ledger example."
#| fig-width: 7
#| fig-height: 3.2
privacy_summary |>
  mutate(
    address = reorder(address, exposure_score),
    service_status = if_else(labeled_service == 1,
                             "Linked to service", "Unlabeled")
  ) |>
  ggplot(aes(x = address, y = exposure_score, fill = service_status)) +
  geom_col(width = 0.68) +
  coord_flip() +
  scale_fill_manual(values = c("Linked to service" = "#B03A2E",
                               "Unlabeled" = "#7F8C8D")) +
  labs(x = "Address", y = "Exposure score", fill = "")
```

The chart makes the linkage problem easier to see. Addresses with repeated
activity, many counterparties, and service labels rise in the exposure ranking.
This kind of score can guide monitoring, clustering, and privacy-risk review.
It should not be read as proof of identity by itself.

The score is a teaching device, but the logic is important. Address reuse and
public service labels reduce ambiguity. A blockchain can be transparent enough
for audit and still expose sensitive relationships, strategies, or customer
activity. This is why privacy is a financial design issue: disclosure affects
trading behavior, institutional adoption, compliance, and personal safety.

Privacy tools change the tradeoff. Address rotation can reduce simple reuse.
Coin-control policies can reduce accidental linkage. Zero-knowledge proofs can
allow verification of selected statements while hiding underlying transaction
details. Permissioned ledgers can restrict who sees transaction data. Each
choice changes auditability, compliance cost, market confidence, and the amount
of information available for analysis.

## Advantages and limits

The chapter's main lesson is that blockchain is a design for financial records
and transfers. Its interest comes from combining records, verification,
incentives, custody, privacy, and settlement rules in one system. Bitcoin shows
the native-asset case: the ledger, asset, issuance rule, and settlement process
belong to the same protocol. Tokenization shows a different use: the ledger
records units whose value may depend on legal claims, servicing, and assets
outside the ledger.

```{r}
#| label: fig-blockchain-closing-map
#| fig-cap: "Blockchain mechanisms and financial modeling questions."
#| fig-width: 8.4
#| fig-height: 2.5
closing_nodes <- tibble(
  id = c("crypto", "ledger", "network", "incentives", "finance"),
  label = c("Cryptography\nhashes + signatures",
            "Ledger\nblocks + registers",
            "Network\nmempool + nodes",
            "Incentives\nfees + rewards",
            "Financial questions\nsettlement + custody\nvaluation + privacy"),
  layer = c("Mechanism", "Mechanism", "Mechanism", "Mechanism", "Finance"),
  x = c(0.7, 2.4, 4.1, 5.8, 7.7),
  y = c(1, 1, 1, 1, 1)
)

closing_edges <- tibble(
  from_x = closing_nodes$x[-nrow(closing_nodes)] + 0.48,
  to_x = closing_nodes$x[-1] - 0.48,
  y = 1
)

ggplot() +
  geom_segment(
    data = closing_edges,
    aes(x = from_x, xend = to_x, y = y, yend = y),
    arrow = grid::arrow(length = grid::unit(0.12, "inches")),
    color = "#566573", linewidth = 0.75
  ) +
  geom_rect(
    data = closing_nodes,
    aes(xmin = x - 0.56, xmax = x + 0.56,
        ymin = y - 0.28, ymax = y + 0.28, fill = layer),
    color = "white", linewidth = 0.9
  ) +
  geom_text(
    data = closing_nodes,
    aes(x = x, y = y, label = label),
    size = 2.7, lineheight = 0.88
  ) +
  scale_fill_manual(values = c("Mechanism" = "#D6EAF8",
                               "Finance" = "#FDEBD0")) +
  coord_cartesian(xlim = c(0.05, 8.35), ylim = c(0.62, 1.38),
                  expand = FALSE, clip = "off") +
  guides(fill = "none") +
  theme_void() +
  theme(plot.margin = margin(2, 4, 2, 4))
```

The figure summarizes the modeling move in the chapter. Technical mechanisms
matter because they change financial questions: who can authorize a transfer,
which history is accepted, how quickly transactions settle, what incentives
shape block inclusion, how token claims are valued, and how much information is
visible to others.

Potential advantages include:

* shared audit trail;
* public verification of ledger history;
* resistance to undetected alteration;
* programmable transfers and token rules;
* faster experimentation with digital assets and settlement logic.

Important limits include:

* scalability and transaction cost;
* privacy leakage;
* governance disputes;
* custody risk from lost or stolen private keys;
* smart-contract or implementation errors;
* regulatory uncertainty;
* weak links between on-chain tokens and off-chain assets.

The strongest way to read blockchain in this book is as a special application
of financial modeling. Prices, returns, portfolios, and VaR model market
behavior. Blockchain models the record-keeping and transfer layer of financial
value. A private key changes who can authorize a payment. A fee changes which
transactions may be included sooner. A confirmation changes settlement risk. A
token rule changes who receives cash flows. A privacy design changes who can
observe financial relationships. Small technical choices can therefore change
the financial meaning of the system.

This makes the chapter a bridge from market models to financial infrastructure.
It asks how value is authorized, recorded, transferred, observed, and sometimes
tokenized. That question is narrower than the full universe of digital assets,
but it is rich enough to connect computation, cryptography, networks, and
finance in one modeling problem.

## Further visual resources

The following resources are optional visual complements. They are most useful
after the R examples, when the reader already has the ledger, hash, signature,
network, and privacy objects in mind.

| Resource | Best use in this chapter |
|---|---|
| [Blockchain Explanation](https://youtu.be/J-ab9was1p0) | Review the full ledger-to-blockchain intuition |
| [How does a blockchain work - Simply Explained](https://youtu.be/SSo_EIwHSd4) | Visualize blocks, hashes, and chain linkage |
| [Asymmetric Encryption - Simply explained](https://youtu.be/AQDCe585Lnc) | Reinforce public-key logic before signatures and custody |
| [Proof-of-Stake (vs proof-of-work)](https://youtu.be/M3EFi_POhps) | Contrast consensus mechanisms after the proof-of-work example |
| [Zero Knowledge Proof - ZKP](https://youtu.be/OcmvMs4AMbM) | Extend the privacy discussion beyond address linkage |
