# Blockchain

```{r blockchain-setup, include=FALSE}
rm(list = ls())

library(digest)
library(dplyr)
library(ggplot2)
library(openssl)
library(tibble)

set.seed(13)
theme_set(theme_minimal(base_size = 12))
```

The previous chapters worked with market data, returns, trading rules, asset
pricing, portfolio allocation, and market risk. This chapter changes the object
of analysis. A blockchain is treated here as a financial-computing mechanism:
a way to maintain a shared record of transfers, validate changes to that
record, and coordinate incentives across participants that do not all rely on a
single central ledger.

The financial question is direct: how can a system record transfers of value
when many participants can read the ledger, submit transactions, and verify the
history? The answer combines several ideas:

* a ledger that records transactions;
* blocks that group transactions;
* cryptographic hashes that link blocks;
* digital signatures that authorize transactions;
* a consensus rule that tells nodes which history to accept;
* incentives that reward participants who help maintain the ledger.

The chapter has three layers. The first layer is a generic blockchain model:
we build a small ledger, define valid transactions, group transactions into
blocks, link blocks with hashes, and validate the resulting history. The second
layer uses Bitcoin as the main proof-of-work case. Bitcoin is useful because it
turns the generic mechanism into an economic system with miners, fees, scarce
issuance, public verification, and costly history rewriting. The third layer
introduces tokenization as a financial extension, where a ledger records
ownership units and cash-flow allocation rules.

The route is therefore:

$$
\mathrm{generic\ ledger\ mechanism}
\longrightarrow
\mathrm{Bitcoin\ proof\ of\ work\ case}
\longrightarrow
\mathrm{tokenized\ financial\ claims}.
$$

The examples are intentionally small. They are meant to show the logic of the
system, not to reproduce a production blockchain.

A useful way to read the chapter is through four frictions: a user may try to
spend the same value twice; block space can become scarce; settlement becomes
more reliable as confirmations accumulate; tokenized claims need a connection
between the ledger record and the legal or operational asset. These frictions
make blockchain a financial topic, because they connect technology to scarcity,
incentives, custody, and settlement.

## Generic blockchain model

The generic model abstracts from the details of any specific blockchain. It
uses an account-based ledger because balances are easy to read: each wallet has
a balance, transactions move value across wallets, fees compensate the block
producer, and the final ledger state can be checked directly. Bitcoin uses a
different accounting model, based on unspent transaction outputs, and that
difference is discussed later.

The generic model answers four questions:

| Component | Question | Financial role |
|---|---|---|
| Ledger state | Who owns what before and after transfers? | Records balances or ownership claims |
| Transaction rule | Which transfers are valid? | Prevents invalid spending and malformed transfers |
| Block structure | Which transactions enter the next ledger update? | Groups accepted transfers into an auditable batch |
| Consensus rule | Which history should nodes extend? | Coordinates participants around one accepted ledger |

The important sequence is:

$$
\mathrm{current\ ledger}
\rightarrow
\mathrm{valid\ transactions}
\rightarrow
\mathrm{candidate\ block}
\rightarrow
\mathrm{accepted\ block}
\rightarrow
\mathrm{updated\ ledger}.
$$

**A shared ledger problem**

A conventional financial ledger records who owns what and how balances change
after transactions. In a centralized system, one institution can update the
ledger, reject invalid transfers, and resolve conflicts. A blockchain moves
part of that logic into a shared protocol. Many participants can hold copies of
the ledger, check the same rules, and agree on the accepted history.

The most basic financial state is a table of balances. Suppose four wallets
start with the following balances.

```{r}
balances <- tibble(
  wallet = c("Alice", "Bob", "Carol", "Dave", "Miner"),
  balance = c(100, 35, 15, 5, 0)
)

balances
```

A transaction changes that state. If wallet $A$ sends amount $x$ to wallet
$B$ and pays fee $f$, the accounting update is:

$$
\mathrm{Balance}_A^{\prime} = \mathrm{Balance}_A - x - f,
\qquad
\mathrm{Balance}_B^{\prime} = \mathrm{Balance}_B + x,
\qquad
\mathrm{Balance}_{\mathrm{miner}}^{\prime}
= \mathrm{Balance}_{\mathrm{miner}}+f.
$$

The transfer is valid only if the sender has enough balance and the transaction
format is acceptable. The rule can be written as:

$$
\mathrm{Balance}_A \geq x+f,
\qquad
x>0,
\qquad
f \geq 0.
$$

The rest of the generic model turns this accounting idea into a blockchain
workflow. The ledger state tells us the balances at a point in time. The block
history tells us how the ledger reached that state. This distinction matters in
finance because a balance without an auditable path is weaker than a balance
that can be traced through accepted transactions.

At this point there is no Bitcoin yet. There is only a minimal transfer system:
wallets, balances, transaction rules, fees, and an accepted ledger history.

**Double-spending pressure**

The first interesting problem is conflict. If Alice has 100 units and submits
two transactions that each spend about 70 units, both transactions cannot settle
inside the same ledger history:

$$
x_1+f_1+x_2+f_2 > \mathrm{Balance}_A.
$$

The system needs a rule for deciding which transaction enters the accepted
history. In a centralized ledger the institution decides. In a blockchain
network, the accepted history comes from transaction validity, block inclusion,
and consensus.

```{r}
double_spend_attempt <- tibble(
  tx_id = c("tx_A", "tx_B"),
  sender = c("Alice", "Alice"),
  receiver = c("Bob", "Carol"),
  amount = c(70, 70),
  fee = c(0.20, 0.40)
)

apply_ledger_order <- function(pool, balances, miner = "Miner") {
  balances_after <- balances
  outcome <- pool |>
    mutate(total_spent = amount + fee, status = NA_character_)

  for (k in seq_len(nrow(outcome))) {
    tx <- outcome[k, ]
    sender_balance <- balances_after$balance[
      balances_after$wallet == tx$sender
    ]

    valid <- sender_balance >= tx$amount + tx$fee

    if (valid) {
      outcome$status[k] <- "Accepted"
      balances_after$balance[balances_after$wallet == tx$sender] <-
        balances_after$balance[balances_after$wallet == tx$sender] -
        tx$amount - tx$fee
      balances_after$balance[balances_after$wallet == tx$receiver] <-
        balances_after$balance[balances_after$wallet == tx$receiver] +
        tx$amount
      balances_after$balance[balances_after$wallet == miner] <-
        balances_after$balance[balances_after$wallet == miner] + tx$fee
    } else {
      outcome$status[k] <- "Rejected"
    }
  }

  outcome
}

double_spend_outcome <- bind_rows(
  apply_ledger_order(double_spend_attempt, balances) |>
    mutate(ordering = "tx_A arrives first"),
  apply_ledger_order(arrange(double_spend_attempt, desc(fee)), balances) |>
    mutate(ordering = "Higher fee arrives first")
)

double_spend_outcome
```

```{r}
#| fig-width: 7.2
#| fig-height: 3.1
ggplot(
  double_spend_outcome,
  aes(x = tx_id, y = total_spent, fill = status)
) +
  geom_col(width = 0.62) +
  facet_wrap(~ ordering) +
  scale_fill_manual(values = c("Accepted" = "#27AE60",
                               "Rejected" = "#C0392B")) +
  labs(x = "Transaction", y = "Amount plus fee", fill = "")
```

The figure shows why a shared ledger needs more than arithmetic. The balances
tell us that the two transfers cannot both be final. The network still needs a
mechanism for selecting one accepted history and rejecting conflicting updates.

**Blocks**

A block is a container for ledger information. In a simple teaching example,
the block contains:

* a block number;
* a timestamp;
* transaction data;
* the hash of the previous block;
* a nonce used by proof-of-work examples;
* the block's own hash.

The diagram below shows the structure.

```{r}
#| label: fig-block-structure
#| fig-cap: "A simplified block structure."
#| fig-width: 7
#| fig-height: 2.6
block_boxes <- tibble(
  label = c("Block number", "Timestamp", "Transactions",
            "Previous hash", "Nonce", "Block hash"),
  xmin = c(0, 0, 0, 3.4, 3.4, 3.4),
  xmax = c(3, 3, 3, 6.4, 6.4, 6.4),
  ymin = c(2.4, 1.2, 0, 2.4, 1.2, 0),
  ymax = c(3.2, 2, 0.8, 3.2, 2, 0.8)
)

ggplot(block_boxes) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "#EAF2F8", color = "#1B4F72", linewidth = 0.7
  ) +
  geom_text(aes(x = (xmin + xmax) / 2, y = (ymin + ymax) / 2, label = label),
            size = 3.7) +
  annotate("text", x = 3.2, y = 3.45, label = "Block", fontface = "bold",
           size = 4.2) +
  coord_cartesian(xlim = c(-0.1, 6.5), ylim = c(-0.05, 3.65), expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(0, 0, 0, 0))
```

In R, a block can be represented as a list.

```{r}
block <- list(
  number = 1,
  timestamp = "2026-01-01 09:00:00",
  data = "Alice pays Bob 10",
  parent_hash = "0",
  nonce = 0
)

block
```

This object is not yet chained to any history. The next step is to compute a
hash.

The block is useful because it gives the system a unit of agreement. Instead of
asking the network to agree on each isolated transaction, the system asks the
network to agree on a batch of transactions and on the previous history that
the batch extends. This is why the parent hash is part of the block: it tells
the reader which past ledger the new block is trying to update.

**Hashes**

A cryptographic hash function maps an input of arbitrary size into a fixed-size
digest. A useful hash function for a ledger has three properties:

* the same input gives the same hash;
* a small change in the input gives a very different hash;
* the input cannot be reconstructed from the hash in any practical way.

For block $i$, we can write the hash as:

$$
h_i = H(\mathrm{data}_i, \mathrm{parentHash}_i,
\mathrm{timestamp}_i, \mathrm{nonce}_i).
$$

For non-genesis blocks, the parent hash equals the hash of the previous block:

$$
\mathrm{parentHash}_i = h_{i-1}.
$$

The examples use SHA-256 through the <tt>`digest`</tt> package. SHA-256 returns
256 bits, which correspond to 32 bytes or 64 hexadecimal characters.

```{r}
digest("Alice pays Bob 10", algo = "sha256")
nchar(digest("Alice pays Bob 10", algo = "sha256"))
```

A tiny change in the message changes the digest.

```{r}
digest("Alice pays Bob 10", algo = "sha256")
digest("Alice pays Bob 11", algo = "sha256")
```

The following helper creates a block and then hashes the full block content.

```{r}
hash_block <- function(block) {
  block$hash <- NULL
  digest(block, algo = "sha256")
}

create_block <- function(number, data, parent_hash = "0",
                         nonce = 0, timestamp = Sys.time()) {
  block <- list(
    number = number,
    timestamp = as.character(timestamp),
    data = data,
    parent_hash = parent_hash,
    nonce = nonce
  )
  block$hash <- hash_block(block)
  block
}

block1 <- create_block(
  number = 1,
  data = "Genesis block",
  parent_hash = "0",
  timestamp = "2026-01-01 09:00:00"
)

block2 <- create_block(
  number = 2,
  data = "Alice pays Bob 10",
  parent_hash = block1$hash,
  timestamp = "2026-01-01 09:01:00"
)

block3 <- create_block(
  number = 3,
  data = "Bob pays Carol 5",
  parent_hash = block2$hash,
  timestamp = "2026-01-01 09:02:00"
)

chain <- list(block1, block2, block3)
chain
```

The financial meaning of this construction is integrity. A later block commits
to the hash of the previous block. If an old transaction changes, the old hash
changes, and the later blocks no longer point to the same history.

This is a commitment mechanism. A block hash is a compact fingerprint of the
block content. A parent hash is a compact fingerprint of the previous accepted
history. Linking those fingerprints creates a chain of commitments: the latest
block implicitly refers to every earlier accepted block through the sequence of
hashes.

**Chain validation**

A basic validation rule checks two conditions:

* each block hash matches the content of that block;
* each non-genesis block points to the hash of the previous block.

```{r}
validate_chain <- function(chain) {
  for (i in seq_along(chain)) {
    block <- chain[[i]]
    declared_hash <- block$hash
    expected_hash <- hash_block(block)

    if (!identical(declared_hash, expected_hash)) {
      return(FALSE)
    }

    if (i == 1 && block$parent_hash != "0") {
      return(FALSE)
    }

    if (i > 1 && block$parent_hash != chain[[i - 1]]$hash) {
      return(FALSE)
    }
  }

  TRUE
}

validate_chain(chain)
```

The chain can be drawn as a sequence of blocks connected by hashes.

```{r}
#| label: fig-chain-hashes
#| fig-cap: "Blocks linked by parent hashes."
#| fig-width: 7.2
#| fig-height: 1.8
chain_plot <- tibble(
  block = c("Block 1", "Block 2", "Block 3"),
  x = c(1, 4, 7),
  y = 1,
  hash = c("h1", "h2", "h3")
)

ggplot(chain_plot) +
  geom_rect(
    aes(xmin = x - 0.9, xmax = x + 0.9, ymin = y - 0.45, ymax = y + 0.45),
    fill = "#E8F6F3", color = "#117864", linewidth = 0.8
  ) +
  geom_text(aes(x = x, y = y + 0.12, label = block), fontface = "bold",
            size = 3.8) +
  geom_text(aes(x = x, y = y - 0.16, label = paste0("hash = ", hash)),
            size = 3.2) +
  geom_segment(
    data = tibble(x = c(1.9, 4.9), xend = c(3.1, 6.1), y = 1, yend = 1),
    aes(x = x, xend = xend, y = y, yend = yend),
    arrow = grid::arrow(length = grid::unit(0.18, "inches")),
    color = "#566573", linewidth = 0.8
  ) +
  annotate("text", x = 2.5, y = 1.33, label = "parent_hash = h1", size = 3) +
  annotate("text", x = 5.5, y = 1.33, label = "parent_hash = h2", size = 3) +
  coord_cartesian(xlim = c(0, 8), ylim = c(0.35, 1.6), expand = FALSE) +
  theme_void() +
  theme(plot.margin = margin(0, 0, 0, 0))
```

Now change an old block.

```{r}
tampered_chain <- chain
tampered_chain[[2]]$data <- "Alice pays Bob 100"

validate_chain(tampered_chain)
```

The validation fails because the declared hash of block 2 no longer matches
the altered content. If block 2 is rehashed, block 3 still points to the old
block 2 hash.

```{r}
tampered_rehashed <- tampered_chain
tampered_rehashed[[2]]$hash <- hash_block(tampered_rehashed[[2]])

validate_chain(tampered_rehashed)
```

This is the core of the chain idea. A historical alteration requires updating
the altered block and every later block. In a real network, the attacker would
also need the rest of the network to accept that altered history.

The toy validation function checks structural consistency. Real systems add
many more checks: transaction signatures, double-spending rules, block-size or
block-weight limits, timestamp rules, consensus-specific constraints, and
network-level acceptance. The simplified function is still useful because it
shows the main idea: a valid ledger history is a sequence whose content and
links agree.

**Proof-of-work**

Hashes are fast to compute, so hashing alone does not make history expensive
to rewrite. Proof-of-work adds a cost. A miner searches for a nonce such that
the block hash satisfies a difficulty condition. In the toy rule below, the
hash must start with $d$ zeros:

$$
\operatorname{substr}(H(\mathrm{block}, \mathrm{nonce}),1,d)
=\underbrace{00\cdots0}_{d\ \mathrm{zeros}}.
$$

Verification is cheap because any node can hash the block once and check the
leading zeros. Finding a valid nonce requires repeated trials.

This asymmetry is the key idea. Producing the proof is costly; checking it is
cheap. In a financial ledger, that asymmetry can be used to make accepted
history costly to replace. The cost does not come from secrecy. Everyone can
see the block and verify the proof. The cost comes from the amount of work
needed to find an alternative valid chain.

```{r}
proof_of_work <- function(block, difficulty) {
  target <- paste(rep("0", difficulty), collapse = "")
  nonce <- 0

  repeat {
    block$nonce <- nonce
    candidate_hash <- hash_block(block)

    if (substr(candidate_hash, 1, difficulty) == target) {
      return(list(nonce = nonce, hash = candidate_hash))
    }

    nonce <- nonce + 1
  }
}

pow_block <- list(
  number = 4,
  timestamp = "2026-01-01 09:03:00",
  data = "Carol pays Dave 3",
  parent_hash = block3$hash
)

proof_of_work(pow_block, difficulty = 2)
```

The number of trials tends to rise quickly with difficulty.

```{r}
difficulty_table <- tibble(difficulty = 1:4) |>
  rowwise() |>
  mutate(nonce = proof_of_work(pow_block, difficulty)$nonce) |>
  ungroup()

difficulty_table
```

```{r}
#| label: fig-pow-difficulty
#| fig-cap: "Proof-of-work trials under increasing toy difficulty."
#| fig-width: 7
#| fig-height: 3.4
ggplot(difficulty_table, aes(x = difficulty, y = nonce)) +
  geom_line(color = "#7B241C", linewidth = 1) +
  geom_point(color = "#C0392B", size = 3) +
  scale_x_continuous(breaks = difficulty_table$difficulty) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Difficulty: required leading zeros", y = "Nonce found")
```

Bitcoin is the canonical proof-of-work case. Miners gather valid transactions,
construct candidate blocks, search for a valid nonce, and broadcast the block
when the proof is found. Ethereum is useful as a contrast because it moved from
proof-of-work to proof-of-stake in 2022. This chapter keeps proof-of-work
because it makes the cost of rewriting history visible in a small R example.

**Transactions and fees**

Transactions are the financial content of blocks. A pending transaction has a
sender, receiver, amount, and fee. The fee compensates the block producer for
including the transaction.

The ledger should reject a transaction for three basic reasons in this toy
model. First, the sender may not have enough balance. Second, the transaction
may have an invalid amount or fee. Third, the block may already be full. These
are different failures. Insufficient balance is an accounting failure. A
negative amount or fee is a rule failure. A full block is a capacity failure.

```{r}
pending_transactions <- tibble(
  tx_id = c("tx1", "tx2", "tx3", "tx4", "tx5"),
  sender = c("Alice", "Bob", "Carol", "Dave", "Alice"),
  receiver = c("Bob", "Carol", "Alice", "Bob", "Carol"),
  amount = c(25, 20, 12, 12, 70),
  fee = c(0.40, 0.15, 0.80, 0.30, 0.60)
)

pending_transactions
```

In many blockchain systems, transactions waiting for inclusion sit in a pending
pool. A block producer may prefer transactions with higher fees, subject to
validity and block capacity. If the block can include $m$ transactions, a
simple selection rule is:

$$
T^* = \operatorname{top}_m(\mathrm{fee}_j)
\quad \mathrm{among\ valid\ pending\ transactions}.
$$

The toy function below processes transactions in descending fee order and
updates balances sequentially.

```{r}
select_for_block <- function(pool, balances, block_size = 3,
                             miner = "Miner") {
  ordered_pool <- arrange(pool, desc(fee))
  accepted <- ordered_pool[0, ]
  rejected <- ordered_pool[0, ]
  balances_after <- balances

  for (k in seq_len(nrow(ordered_pool))) {
    tx <- ordered_pool[k, ]

    if (nrow(accepted) >= block_size) {
      rejected <- bind_rows(rejected, tx)
      next
    }

    sender_balance <- balances_after$balance[
      balances_after$wallet == tx$sender
    ]

    valid <- length(sender_balance) == 1 &&
      sender_balance >= tx$amount + tx$fee &&
      tx$amount > 0 &&
      tx$fee >= 0

    if (valid) {
      accepted <- bind_rows(accepted, tx)
      balances_after$balance[balances_after$wallet == tx$sender] <-
        balances_after$balance[balances_after$wallet == tx$sender] -
        tx$amount - tx$fee
      balances_after$balance[balances_after$wallet == tx$receiver] <-
        balances_after$balance[balances_after$wallet == tx$receiver] +
        tx$amount
      balances_after$balance[balances_after$wallet == miner] <-
        balances_after$balance[balances_after$wallet == miner] + tx$fee
    } else {
      rejected <- bind_rows(rejected, tx)
    }
  }

  list(accepted = accepted, rejected = rejected, balances = balances_after)
}

block_selection <- select_for_block(
  pending_transactions,
  balances,
  block_size = 3,
  miner = "Miner"
)

block_selection$accepted
block_selection$rejected
block_selection$balances
```

The accepted transactions are valid under the balance rule and fit inside the
block capacity. The rejected transactions either fail the validity test or are
left outside the block capacity in this simple example.

This small example also shows why fees matter. If several valid transactions
compete for limited block capacity, a fee ranking can influence which transfers
enter first. In production systems the selection rule can be more complex, but
the economic intuition is already visible: users pay for inclusion, and block
producers respond to those incentives.

```{r}
#| fig-width: 7
#| fig-height: 3.2
fee_competition <- pending_transactions |>
  mutate(
    status = if_else(tx_id %in% block_selection$accepted$tx_id,
                     "Included in block", "Left out or invalid"),
    tx_id = reorder(tx_id, fee)
  )

ggplot(fee_competition, aes(x = tx_id, y = fee, fill = status)) +
  geom_col(width = 0.65) +
  coord_flip() +
  scale_fill_manual(values = c("Included in block" = "#2E86C1",
                               "Left out or invalid" = "#BFC9CA")) +
  labs(x = "Pending transaction", y = "Fee", fill = "")
```

This turns block construction into a market for scarce inclusion. The fee is
small relative to the transfer amount, but it can decide timing when many
transactions compete for limited capacity. This is one reason public blockchain
data can be financially rich: pending transactions, fees, and inclusion delays
reveal pressure on the settlement layer.

```{r}
#| label: fig-wallet-balances
#| fig-cap: "Wallet balances before and after the mined block."
balance_plot <- bind_rows(
  mutate(balances, state = "Before block"),
  mutate(block_selection$balances, state = "After block")
)

ggplot(balance_plot, aes(x = wallet, y = balance, fill = state)) +
  geom_col(position = "dodge", width = 0.68) +
  scale_fill_manual(values = c("Before block" = "#2C7FB8",
                               "After block" = "#F39C12")) +
  labs(x = "Wallet", y = "Balance", fill = "")
```

The miner balance increases by the sum of the fees included in the block:

$$
\mathrm{Reward}_{\mathrm{fees}}
=\sum_{j\in T^*} \mathrm{fee}_j.
$$

```{r}
sum(block_selection$accepted$fee)
```

In Bitcoin, miners can also receive a block subsidy according to the issuance
schedule. In this toy account-based model, only transaction fees are shown.

**Mining a transaction block**

The accepted transactions can now become block data. The block hash commits to
the transaction set, the parent hash, the timestamp, and the nonce.

```{r}
candidate_block <- list(
  number = 4,
  timestamp = "2026-01-01 09:03:00",
  data = block_selection$accepted,
  parent_hash = block3$hash
)

pow_result <- proof_of_work(candidate_block, difficulty = 2)
candidate_block$nonce <- pow_result$nonce
candidate_block$hash <- pow_result$hash

transaction_chain <- c(chain, list(candidate_block))

validate_chain(transaction_chain)
candidate_block
```

Changing a transaction after the block is mined breaks the chain.

```{r}
altered_transaction_chain <- transaction_chain
altered_transaction_chain[[4]]$data$amount[1] <- 250

validate_chain(altered_transaction_chain)
```

This example connects three financial objects: pending transactions, balances,
and the accepted ledger state. The hash does not make the transaction true by
itself. The validation rules, signatures, balance checks, consensus rule, and
network acceptance determine which transactions enter the ledger.

The example also separates two views of the same system. The transaction view
asks which transfers are accepted. The state view asks what the balances are
after those transfers. Financial applications need both views: the history is
needed for auditability, while the state is needed for valuation, reporting,
and risk controls.

**Digital signatures**

A transaction also needs authorization. Digital signatures use public-key
cryptography to connect a message with a private key. The sender signs the
transaction with a private key. Anyone with the corresponding public key can
verify that the signature matches the message.

This provides two properties:

* authenticity: the signature matches the public key;
* integrity: changing the message invalidates the signature.

Confidentiality is a separate issue. A signature verifies who authorized a
message and whether it changed. Encryption hides message contents from parties
that do not have the decryption key.

In ledger terms, a signature answers a narrow question: did the holder of the
relevant private key authorize this exact message? It does not prove that the
sender is a real-world person, that the transfer is legal, or that the asset
behind a token exists. Those questions require identity, legal, operational,
and governance layers outside the cryptographic signature.

```{r}
alice_key <- rsa_keygen()
alice_public_key <- alice_key$pubkey

signed_transaction <- list(
  sender = "Alice",
  receiver = "Bob",
  amount = 10,
  fee = 0.1
)

serialized_transaction <- serialize(signed_transaction, NULL)
signature <- signature_create(serialized_transaction, sha256, key = alice_key)

signature_verify(
  serialized_transaction,
  signature,
  sha256,
  pubkey = alice_public_key
)
```

Now alter the transaction and verify the old signature against the altered
message.

```{r}
altered_signed_transaction <- list(
  sender = "Alice",
  receiver = "Carol",
  amount = 10,
  fee = 0.1
)

altered_serialized_transaction <- serialize(altered_signed_transaction, NULL)

tryCatch(
  signature_verify(
    altered_serialized_transaction,
    signature,
    sha256,
    pubkey = alice_public_key
  ),
  error = function(e) FALSE
)
```

The altered message fails verification. This is why signatures are part of the
financial logic of a blockchain: the ledger should only accept transactions
authorized by the holder of the relevant private key.

**Network and consensus**

A blockchain ledger is distributed over a peer-to-peer network. A simplified
proof-of-work workflow is:

* users broadcast transactions;
* nodes check transaction validity;
* miners gather valid pending transactions into candidate blocks;
* miners search for a proof-of-work;
* the first valid block is broadcast;
* nodes accept the block if the proof and the transactions are valid;
* new blocks build on the accepted parent hash.

```{r}
#| label: fig-network-flow
#| fig-cap: "A simplified peer-to-peer blockchain network."
#| fig-width: 8.6
#| fig-height: 2.8
network_nodes <- tibble(
  id = c("wallet_a", "wallet_b", "wallet_c", "pool",
         "miner_1", "miner_2", "node_1", "node_2", "node_3", "ledger"),
  label = c("Wallet A", "Wallet B", "Wallet C", "Pending\npool",
            "Miner 1", "Miner 2", "Full node\n1", "Full node\n2",
            "Full node\n3", "Accepted\nledger"),
  type = c("Wallet", "Wallet", "Wallet", "Pool",
           "Miner", "Miner", "Node", "Node", "Node", "Ledger"),
  x = c(0, 0, 0, 2.2, 4.3, 4.3, 6.5, 6.5, 6.5, 8.7),
  y = c(3, 2, 1, 2, 2.7, 1.3, 3, 2, 1, 2)
)

network_edges <- tibble(
  from = c("wallet_a", "wallet_b", "wallet_c", "pool", "pool",
           "miner_1", "miner_2", "miner_1", "miner_2",
           "node_1", "node_2", "node_3"),
  to = c("pool", "pool", "pool", "miner_1", "miner_2",
         "node_1", "node_3", "node_2", "node_2",
         "ledger", "ledger", "ledger")
) |>
  left_join(network_nodes |> select(from = id, x, y), by = "from") |>
  rename(x_from = x, y_from = y) |>
  left_join(network_nodes |> select(to = id, x, y), by = "to") |>
  rename(x_to = x, y_to = y)

ggplot() +
  geom_segment(
    data = network_edges,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
    arrow = grid::arrow(length = grid::unit(0.12, "inches")),
    color = "#7F8C8D", linewidth = 0.65, alpha = 0.9
  ) +
  geom_point(
    data = network_nodes,
    aes(x = x, y = y, fill = type),
    shape = 21, size = 12, color = "white", stroke = 1.1
  ) +
  geom_text(
    data = network_nodes,
    aes(x = x, y = y, label = label),
    size = 2.95, lineheight = 0.88
  ) +
  scale_fill_manual(values = c(
    "Wallet" = "#D6EAF8",
    "Pool" = "#FCF3CF",
    "Miner" = "#FADBD8",
    "Node" = "#D5F5E3",
    "Ledger" = "#E8DAEF"
  )) +
  coord_cartesian(xlim = c(-0.45, 9.1), ylim = c(0.65, 3.35), expand = FALSE) +
  guides(fill = "none") +
  theme_void() +
  theme(plot.margin = margin(0, 0, 0, 0))
```

The diagram is a simplified network view. Wallets submit transactions to the
pending pool. Miners collect transactions and propose blocks. Full nodes verify
the candidate block and the ledger rules. The accepted ledger represents the
history that nodes are willing to extend.

Consensus is the rule that tells nodes which valid history to extend. In
Bitcoin, the accepted history is tied to accumulated proof-of-work. The economic
security argument is that rewriting history requires controlling enough
computational work to replace the accepted chain, which is costly.

This is also where incentives matter. Fees and block rewards encourage miners
to spend resources on valid blocks. Invalid blocks waste effort because nodes
reject them.

## Bitcoin case: a proof-of-work ledger

At this point the chapter has built a generic blockchain mechanism. Bitcoin is
the next step because it turns that mechanism into a specific monetary and
payment system. Its design connects payments, proof-of-work, scarce issuance,
public verification, and private-key custody [@nakamoto2008bitcoin].

The generic model and Bitcoin share the same broad workflow:

$$
\mathrm{signed\ transaction}
\rightarrow
\mathrm{pending\ transaction\ set}
\rightarrow
\mathrm{candidate\ block}
\rightarrow
\mathrm{proof\ of\ work}
\rightarrow
\mathrm{network\ validation}
\rightarrow
\mathrm{accepted\ ledger\ history}.
$$

The differences are important:

| Generic model in this chapter | Bitcoin case |
|---|---|
| Account balances are stored directly for teaching clarity | Balances are inferred from unspent transaction outputs |
| Transactions move amounts from sender to receiver | Transactions consume previous outputs and create new outputs |
| The miner receives transaction fees in the toy balance table | Miners can receive fees and a protocol-defined block subsidy |
| Difficulty is a small number of leading zeros | Difficulty adjusts so blocks arrive near the protocol target interval |
| Finality is shown as a valid chain in R | Settlement confidence grows as more blocks build on top |

The relevant Bitcoin network roles can be represented as nodes rather than as a
single line of steps. A wallet creates a signed transaction and broadcasts it.
The mempool is the waiting area for valid pending transactions. Miners convert
some of those transactions into candidate blocks. Full nodes check the proposed
block against the consensus rules. Exchanges, custodians, merchants, and block
explorers sit outside the core protocol while remaining financially important
as common points of access, monitoring, and settlement interpretation.

```{r}
#| label: fig-bitcoin-network-roles
#| fig-cap: "Relevant roles in a simplified Bitcoin network."
#| fig-width: 8.4
#| fig-height: 2.8
bitcoin_nodes <- tibble(
  id = c("payer", "receiver", "mempool", "miner", "full_nodes",
         "chain", "exchange", "explorer"),
  label = c("Payer\nwallet", "Receiver\nwallet", "Mempool", "Miner",
            "Full\nnodes", "Accepted\nchain", "Exchange/\ncustodian",
            "Explorer/\nanalytics"),
  type = c("Wallet", "Wallet", "Pool", "Miner", "Node",
           "Ledger", "Access", "Analytics"),
  x = c(0.4, 7.7, 2.3, 4.1, 5.9, 7.7, 0.4, 5.9),
  y = c(2.5, 0.8, 1.65, 2.45, 1.65, 1.65, 0.8, 0.45)
)

bitcoin_edges <- tibble(
  from = c("payer", "mempool", "miner", "full_nodes", "chain",
           "exchange", "chain", "full_nodes"),
  to = c("mempool", "miner", "full_nodes", "chain", "receiver",
         "mempool", "explorer", "exchange")
) |>
  left_join(bitcoin_nodes |> select(from = id, x, y), by = "from") |>
  rename(x_from = x, y_from = y) |>
  left_join(bitcoin_nodes |> select(to = id, x, y), by = "to") |>
  rename(x_to = x, y_to = y)

ggplot() +
  geom_segment(
    data = bitcoin_edges,
    aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
    arrow = grid::arrow(length = grid::unit(0.11, "inches")),
    color = "#7F8C8D", linewidth = 0.6, alpha = 0.9
  ) +
  geom_point(
    data = bitcoin_nodes,
    aes(x = x, y = y, fill = type),
    shape = 21, size = 12, color = "white", stroke = 1.1
  ) +
  geom_text(
    data = bitcoin_nodes,
    aes(x = x, y = y, label = label),
    size = 2.85, lineheight = 0.88
  ) +
  scale_fill_manual(values = c(
    "Wallet" = "#D6EAF8",
    "Pool" = "#FCF3CF",
    "Miner" = "#FADBD8",
    "Node" = "#D5F5E3",
    "Ledger" = "#E8DAEF",
    "Access" = "#FDEBD0",
    "Analytics" = "#EAECEE"
  )) +
  coord_cartesian(xlim = c(-0.2, 8.2), ylim = c(0.05, 2.95), expand = FALSE) +
  guides(fill = "none") +
  theme_void() +
  theme(plot.margin = margin(0, 0, 0, 0))
```

The diagram separates protocol roles from financial access roles. Wallets,
miners, full nodes, and the accepted chain describe the transaction validation
path. Exchanges and custodians describe how many users actually enter the
system. Explorers and analytics tools describe how public data becomes
observable, searchable, and useful for monitoring.

The account-based model says that Alice has a balance and can send part of it.
Bitcoin uses a UTXO model. A UTXO is an unspent transaction output. If Alice
controls previous outputs worth 0.8 BTC and 0.4 BTC, and wants to pay 0.9 BTC,
her transaction spends enough previous outputs and creates new outputs: one for
the receiver, one for change back to Alice, and one implicit fee to the miner.
The balance is therefore reconstructed from the set of unspent outputs
controlled by Alice's keys.

The simplified accounting equation is:

$$
\sum \mathrm{Inputs} = \sum \mathrm{Outputs} + \mathrm{Fee}.
$$

The fee is the difference between the value consumed by the transaction and the
value assigned to new outputs:

$$
\mathrm{Fee} = \sum \mathrm{Inputs} - \sum \mathrm{Outputs}.
$$

This differs from the account example above, where the fee was an explicit
column. The economic meaning is similar: the sender gives up value, the
receiver obtains value, and the miner has an incentive to include the
transaction.

Bitcoin also makes proof-of-work part of monetary security. A miner proposes a
block and searches for a valid proof. Nodes verify the block by checking the
proof, transaction validity, and the rule that spent outputs have not already
been spent. A transaction with one confirmation is included in one block. More
confirmations mean that additional blocks have been built on top of that block,
raising the cost of replacing that history.

The settlement idea is probabilistic. A payment becomes harder to reverse as
more proof-of-work accumulates above it. This is why confirmations matter in
Bitcoin payments: they are a practical measure of how deeply a transaction sits
inside the accepted history.

One simplified way to visualize the intuition is a catch-up race. Suppose an
attacker controls share $q$ of mining power, with $0<q<0.5$, and an honest
chain is already $z$ confirmations ahead. A compact approximation for the
probability that the attacker catches up is:

$$
\Pr(\mathrm{catchUp}\mid z,q)
\approx
\left(\frac{q}{1-q}\right)^z.
$$

The approximation is a teaching device. It captures the central shape: when the
attacker has less than half of total work, each additional confirmation lowers
the catch-up probability.

```{r}
#| fig-width: 7.2
#| fig-height: 3.5
confirmation_grid <- as_tibble(expand.grid(
  confirmations = 1:10,
  attacker_share = c(0.10, 0.20, 0.30, 0.40)
)) |>
  mutate(
    catch_up_probability =
      (attacker_share / (1 - attacker_share))^confirmations,
    attacker_label = paste0("q = ",
                            scales::percent(attacker_share, accuracy = 1))
  )

ggplot(
  confirmation_grid,
  aes(x = confirmations, y = catch_up_probability, color = attacker_label)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:10) +
  scale_y_log10(labels = scales::label_scientific(digits = 2)) +
  labs(x = "Confirmations",
       y = "Catch-up probability (log scale)", color = "")
```

The logarithmic scale makes the settlement intuition visible. Confirmations
translate settlement depth into a practical risk measure: they express the cost
of replacing recent history. The same payment can be operationally acceptable at different
confirmation depths depending on amount, counterparty risk, liquidity needs,
and the recipient's tolerance for reversal risk.

Bitcoin is financially interesting for four reasons.

First, it turns ledger maintenance into an incentive problem. Miners spend
resources because accepted blocks can pay rewards and fees. Second, it makes
custody explicit: control of private keys is control over spending authority.
Third, it creates a public audit trail, so anyone can verify the transaction
history according to protocol rules. Fourth, it introduces a digital asset with
a rule-based issuance schedule, which makes scarcity part of the protocol
design.

The R model above should be read as a simplified map. It captures blocks,
hashes, proof-of-work, fees, validation, and ledger updates. It leaves out many
Bitcoin details: UTXO selection algorithms, script rules, block propagation,
difficulty adjustment mechanics, mempool policy, fee estimation, wallet
software, and network topology. The simplification is deliberate. The goal is
to make the financial logic visible before adding production-level complexity.

## Tokenization

Tokenization extends the ledger idea from native coins to other financial or
economic claims. A token can represent a unit in a digital asset, a claim on a
cash-flow arrangement, a fund share, a loyalty point, a stablecoin balance, or
a record linked to an off-chain asset. The crucial point is that the token
record and the legal or operational claim have to be aligned.

The shift from Bitcoin to tokenization changes the financial question. In
Bitcoin, the native asset and the ledger are part of the same protocol. In
tokenization, the ledger records ownership units whose value often depends on
something outside the ledger: cash, securities, real assets, fund shares,
invoices, commodities, or contractual rights. That creates two layers:

| Layer | Question |
|---|---|
| On-chain record | Who holds how many tokens according to the ledger? |
| Off-chain claim | What legal, economic, or operational right does each token represent? |

The on-chain record can be precise while the off-chain claim still requires
legal documentation, custody, audits, servicing, redemption rules, and dispute
resolution. This is why tokenization is a finance topic as much as a technology
topic.

```{r}
#| fig-width: 8
#| fig-height: 1.1
token_layers <- tibble(
  id = c("asset", "legal", "issuer", "registry", "investors", "cashflow"),
  label = c("Underlying\nasset", "Legal\nclaim", "Issuer/\nservicer",
            "Token\nregistry", "Investors", "Cash-flow\nrule"),
  layer = c("Off-chain", "Off-chain", "Bridge",
            "On-chain", "Market", "Finance"),
  x = c(0.5, 1.9, 3.3, 4.7, 6.1, 7.5),
  y = c(1.5, 1.5, 1.5, 1.5, 1.5, 1.5)
)

token_edges <- tibble(
  from = c("asset", "legal", "issuer", "registry", "investors"),
  to = c("legal", "issuer", "registry", "investors", "cashflow")
) |>
  left_join(token_layers |> select(from = id, x, y), by = "from") |>
  rename(x_from = x, y_from = y) |>
  left_join(token_layers |> select(to = id, x, y), by = "to") |>
  rename(x_to = x, y_to = y) |>
  mutate(
    x_from_plot = x_from + 0.36,
    x_to_plot = x_to - 0.36
  )

ggplot() +
  geom_segment(
    data = token_edges,
    aes(x = x_from_plot, y = y_from, xend = x_to_plot, yend = y_to),
    arrow = grid::arrow(length = grid::unit(0.11, "inches")),
    color = "#7F8C8D", linewidth = 0.65
  ) +
  geom_point(
    data = token_layers,
    aes(x = x, y = y, fill = layer),
    shape = 21, size = 14, color = "white", stroke = 1.1
  ) +
  geom_text(
    data = token_layers,
    aes(x = x, y = y, label = label),
    size = 2.8, lineheight = 0.88
  ) +
  scale_fill_manual(values = c(
    "Off-chain" = "#D6EAF8",
    "Bridge" = "#FDEBD0",
    "On-chain" = "#D5F5E3",
    "Market" = "#E8DAEF",
    "Finance" = "#FCF3CF"
  )) +
  coord_cartesian(xlim = c(-0.1, 8.1), ylim = c(1.35, 1.65),
                  expand = FALSE, clip = "off") +
  guides(fill = "none") +
  theme_void() +
  theme(plot.margin = margin(8, 2, 8, 2))
```

This stack is where tokenization becomes a financial structure. The token
registry can say who holds units. Valuation and risk also depend on the asset,
legal claim, servicing process, transfer restrictions, and cash-flow rule. A
token model is incomplete unless the financial claim represented by the token
is explicit.

In a simple model, a token contract has:

* an issuer;
* a token name;
* total supply;
* holder balances;
* transfer rules;
* optional cash-flow or redemption rules.

Suppose a hypothetical issuer creates 1,000 tokens linked to participation in a
small financial asset pool.

```{r}
token_metadata <- tibble(
  token = "POOL_A_TOKEN",
  underlying = "Hypothetical asset-pool participation",
  total_supply = 1000,
  unit_claim = "1 token = 0.1% of the pool cash-flow rule"
)

token_register <- tibble(
  holder = c("Issuer", "Investor_A", "Investor_B", "Investor_C"),
  token_balance = c(1000, 0, 0, 0)
)

token_metadata
token_register
```

Token transfers are ledger updates. If holder $a$ transfers $q$ tokens to
holder $b$, the token register changes as:

$$
\mathrm{Token}_a^{\prime} = \mathrm{Token}_a-q,
\qquad
\mathrm{Token}_b^{\prime} = \mathrm{Token}_b+q,
\qquad
0 \leq q \leq \mathrm{Token}_a.
$$

```{r}
transfer_tokens <- function(register, from, to, amount) {
  from_balance <- register$token_balance[register$holder == from]

  valid <- length(from_balance) == 1 &&
    from_balance >= amount &&
    amount > 0 &&
    to %in% register$holder

  if (!valid) {
    stop("Invalid token transfer")
  }

  register$token_balance[register$holder == from] <-
    register$token_balance[register$holder == from] - amount
  register$token_balance[register$holder == to] <-
    register$token_balance[register$holder == to] + amount

  register
}

token_register_after <- token_register |>
  transfer_tokens("Issuer", "Investor_A", 400) |>
  transfer_tokens("Issuer", "Investor_B", 350) |>
  transfer_tokens("Issuer", "Investor_C", 150) |>
  transfer_tokens("Investor_A", "Investor_C", 50)

token_register_after
```

```{r}
#| label: fig-token-register
#| fig-cap: "Token ownership before and after transfers."
token_plot <- bind_rows(
  mutate(token_register, state = "Initial register"),
  mutate(token_register_after, state = "After transfers")
)

ggplot(token_plot, aes(x = holder, y = token_balance, fill = state)) +
  geom_col(position = "dodge", width = 0.68) +
  scale_fill_manual(values = c("Initial register" = "#566573",
                               "After transfers" = "#27AE60")) +
  labs(x = "Holder", y = "Token balance", fill = "")
```

If the pool distributes cash flow $C$, a proportional token rule allocates:

$$
\mathrm{CashFlow}_h = \frac{\mathrm{Token}_h}{S}C,
$$

where $S$ is total token supply.

```{r}
cash_flow <- 5000

cash_distribution <- token_register_after |>
  mutate(
    ownership = token_balance / sum(token_balance),
    cash_flow = ownership * cash_flow
  )

cash_distribution
```

This simple example shows why tokenization belongs in a financial modeling
course. It combines a ledger, ownership units, transfer rules, and cash-flow
allocation. The hard parts appear quickly: custody, investor rights, transfer
restrictions, identity checks, tax treatment, legal enforceability, valuation,
and operational controls.

## Privacy

Public blockchains make verification easier by exposing transaction history,
but that transparency creates privacy questions. A public address can be
pseudonymous, yet transaction patterns may still reveal relationships. If an
address is linked to a person or institution, past and future activity can
become easier to trace.

The banking model usually limits transaction information to the parties, the
bank, and regulators. A public blockchain makes more information visible to the
network. Privacy-preserving tools, address management, zero-knowledge proofs,
and permissioned ledgers are attempts to balance auditability and privacy.

## Advantages and limits

The chapter's main lesson is that blockchain is a ledger design. Its financial
interest comes from combining records, verification, incentives, and settlement
rules in one protocol.

Potential advantages include:

* shared audit trail;
* public verification of ledger history;
* resistance to undetected alteration;
* programmable transfers and token rules;
* faster experimentation with digital assets and settlement logic.

Important limits include:

* scalability and transaction cost;
* privacy leakage;
* governance disputes;
* custody risk from lost or stolen private keys;
* smart-contract or implementation errors;
* regulatory uncertainty;
* weak links between on-chain tokens and off-chain assets.

The strongest way to read blockchain in this book is as an extension of
financial modeling. Prices, returns, portfolios, and VaR model market behavior.
Blockchain models the record-keeping and transfer layer of financial value. The
technical details matter because a small change in validation, incentives,
privacy, or token rights can change the financial meaning of the system.

## Further visual resources

The following resources can be useful as visual complements to the R examples:

* [Blockchain Explanation](https://youtu.be/J-ab9was1p0)
* [How does a blockchain work - Simply Explained](https://youtu.be/SSo_EIwHSd4)
* [Asymmetric Encryption - Simply explained](https://youtu.be/AQDCe585Lnc)
* [Proof-of-Stake (vs proof-of-work)](https://youtu.be/M3EFi_POhps)
* [Zero Knowledge Proof - ZKP](https://youtu.be/OcmvMs4AMbM)
