```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  out.extra = ""
)
```

```{r book-edition, echo=FALSE, results='asis'}
source("R/book-edition.R")
write_book_edition_script()
```

```{r contact-links, echo=FALSE, results='asis'}
xfun::file_string("contact.html")
```

```{r publication-dates, echo=FALSE, results='asis'}
write_book_dates()
```

## Preface {.unnumbered}

::: {.home-panel}
This book introduces financial modeling with R for readers who want to extract,
visualize, and analyze financial data while keeping the financial idea and the
computation together. The chapters connect data workflows, prices and returns,
asset pricing, portfolio allocation, risk-return reasoning, and blockchain
examples through reproducible code, figures, tables, equations, and
interpretation.
:::

## What's new in this edition {.unnumbered}

- **Book-based reading path:** The material is now organized as a Quarto book with independent chapters and a clear reading sequence.
- **Clearer navigation:** The sidebar, chapter sections, local page menu, progress bar, and back-to-top control make it easier to move through the financial examples.
- **Reproducible publication:** The site is ready to render into `_book` and publish from GitHub Actions while keeping generated HTML out of the repository root.
