```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  out.extra = ""
)
```

```{r book-edition, echo=FALSE, results='asis'}
source("R/book-edition.R")
write_book_edition_script()
```

```{r contact-links, echo=FALSE, results='asis'}
xfun::file_string("contact.html")
```

```{r publication-dates, echo=FALSE, results='asis'}
write_book_dates()
```

## Preface {.unnumbered}

::: {.home-panel}
This book introduces financial modeling with R for readers who want to extract,
visualize, and analyze financial data while keeping the financial idea and the
computation together. The reading path starts with financial data, then moves
to prices and returns, trading rules, asset pricing, portfolio allocation, and
market risk measurement. The final blockchain chapter works as a special
application that extends the financial modeling toolkit toward distributed
records, cryptographic validation, and transaction design. Throughout the book,
the emphasis is on reproducible code, figures, tables, equations, and careful
interpretation.
:::

## What's new in this edition {.unnumbered}

- **Book-based reading path:** The material is now organized as a Quarto book with independent chapters and a clear reading sequence.
- **Applied trading workflow:** The book now includes a robo trader chapter that connects prices, returns, technical indicators, classification, and simple trading rules.
- **Portfolio risk measurement:** Value at Risk is now part of the financial modeling path, linking portfolio allocation with market risk and loss scenarios.
- **Clearer navigation:** The sidebar, chapter sections, local page menu, progress bar, and back-to-top control make it easier to move through the financial examples.
- **Reproducible publication:** The site is ready to render into `_book` and publish from GitHub Actions while keeping generated HTML out of the repository root.
