---
title: "Data, pricing, and portfolio analysis in R."
author:
- Dr. Martín Lozano.
- sites.google.com/site/mlozanoqf/
date: "`r format(Sys.time(), '%B %d, %Y. %H:%M:%S')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: yes
  word_document: default
  html_notebook: default
  html_document:
    df_print: paged
toc: yes
fontsize: 12pt
subtitle: An introductory approach of quantitative finance.
abstract: This is an introductory applied tutorial suitable for undergraduate students interested in the area of quantitative finance. The main objective is to show how to analyze financial data and extract basic insights from asset pricing and asset allocation applications using the powerful R programming language. We extend examples and codes freely available in Internet, and some of my own. Here, we do not follow a formal approach as most mathematical background is skipped to emphasize the data analysis, model logic, discussion, graphical approach and R coding (literate programming). This is still a work in progress and it is under revision.
---

------------------------------------------------------------------------

\newpage

```{r Preface, child = 'preface.Rmd'}
```

\newpage

```{r Introduction, child = 'intro.Rmd'}
```

\newpage

```{r FinancialData, child = 'findata.Rmd'}
```

\newpage

```{r AssetReturns, child = 'assetret.Rmd'}
```

\newpage

```{r AssetPricing, child = 'assetpricing.Rmd'}
```

\newpage

```{r AssetAllocation, child = 'assetallocation.Rmd'}
```

\newpage

```{r DatavizMultilevel, child = 'multilevel.Rmd'}
```

\newpage

```{r Blockchain, child = 'blockchain.Rmd'}
```

# Conclusion.

Finance is not about how to make money; it is about much more than that. Finance is about how to find and use resources, and how to assign them into projects. This is not an easy task as resources are limited and projects are risky. Moreover, sometimes the financial markets do not work well: bad projects are financed and good projects are not. However, if we rely on finance theory, quantitative models, and data analysis, we increase the chances to make good and informed decisions. Making money is a necessary step that allows us to pursue superior objectives like making firms growth, create better jobs, stimulate innovation, economic growth, and hopefully improve the standard of living of the population.

We have made some progress at implementing financial and economic models to achieve these superior objectives but we cannot say we have succeeded. Poverty levels in some countries are high, and income inequality leads not only to economic but also social problems. Fortunately, there is an increasing and genuine interest in learning finance principles by good people around the world.

```{r include=FALSE}
a <- toc()
```

This document took `r as.numeric(a$toc-a$tic)` seconds to compile in Rmarkdown.

\newpage

# R and \LaTeX References.

The following are a constant growing list of online R and \LaTeX references.

## Instalation guides.

As you may understand, versions are frequently updated. If you find a newer video for installing a newer version please share. In any case, these videos can definitely help you as a guide to install the newer version available.

-   [Download & Install R 3.6.3](https://youtu.be/3OxPMYP8lNU)
-   [Download & Install RStudio Desktop 1.3.959](https://youtu.be/uvnuQ_fKrMc)
-   [Installing R and Rstudio on MacOS.](https://youtu.be/Y20P3u3c_1c)
-   [RStudio: A Guided Tour (by Jamison Crawford).](https://youtu.be/xgPwDlAtuNI)
-   [How to Install RStudio (and Knit to PDF).](https://youtu.be/2Sovzf6lVRo)
-   [RStudio: Explaining the Interface & R Markdown.](https://youtu.be/AHAR7j-IUOw)

## Main sites and program webpages.

-   [The R Project for Statistical Computing.](https://www.r-project.org/)
-   [RStudio.](https://www.rstudio.com/)
-   [DataCamp.](https://www.datacamp.com/)
-   [Tiny\TeX: A lightweight, cross-platform, portable, and easy-to-maintain \LaTeX distribution based on \TeX Live.](https://yihui.org/tinytex/)
-   [Swirl.](https://swirlstats.com/)
-   [Compile R online.](https://rextester.com/l/r_online_compiler)
-   [\LaTeX base.](https://latexbase.com)

## Books and e-books.

Learn R by reading these books.

-   [R For Dummies.](http://sgpwe.izt.uam.mx/files/users/uami/gma/R_for_dummies.pdf)
-   [R Programming for Data Science.](https://bookdown.org/rdpeng/rprogdatascience/)
-   [Foundations of Statistics with R.](https://mathstat.slu.edu/~speegle/_book/)
-   [Introduction to Econometrics with R.](https://www.econometrics-with-r.org/)
-   [Forecasting: Principles and Practice.](https://otexts.com/fpp3/)
-   [R Markdown Cookbook.](https://bookdown.org/yihui/rmarkdown-cookbook)
-   [R for Data Science.](https://r4ds.had.co.nz/)
-   [R Markdown: The Definitive Guide.](https://bookdown.org/yihui/rmarkdown/)
-   [Authoring Books and Technical Documents with R Markdown.](https://bookdown.org/yihui/bookdown)
-   [BOOKDOWN.](https://bookdown.org/)

## Blogs and tutorials.

Here you can find questions and answers about programming in R.

-   [R-Bloggers.](https://www.r-bloggers.com)
-   [Stack Overflow.](https://stackoverflow.com)
-   [R and Data Mining.](http://www.rdatamining.com)
-   [Revolutions.](https://blog.revolutionanalytics.com/)
-   [R-ladies.](https://rladies.org/)
-   [Interactive Tutorials for R.](https://rstudio.github.io/learnr/)
-   [Social Science Data and Statistics Resources.](https://researchguides.library.tufts.edu/data/r)
-   [Stack Exchange.](https://tex.stackexchange.com/)
-   [Kaggle.](https://www.kaggle.com/)
-   [Learn Data Science.](https://www.business-science.io/)

## Others.

-   [ProjectElon - Study Motivation.](https://www.youtube.com/c/ProjectElon/)
-   [iPanda.](https://www.youtube.com/c/iPandaChannel)
-   Twitter hashtags: \#rstats, \#DataScience
-   [Why Study a Postgraduate Degree in Quantitative Finance (by one of my current MSc students.)](https://youtu.be/7xpydgOrzkw)

## Books.

-   Elton EJ, MJ Gruber, SJ Brown and WN Goetzmann (2014) Modern Portfolio Theory and Investment Analysis, 9th Edition, New York: Wiley.
-   Hull JC (2017) Fundamentals of Futures and Options Markets, 8th Edition, Harlow UK: Pearson.
-   Hull JC (2019) Machine Learning in Business : An Introduction to the World of Data Science: Fourth Printing.
-   Richard A. Brealey, Stewart C. Myers, Franklin Allen. Principles of Corporate Finance, 10th Edition.
-   Crouhy M, D Galai and R Mark (2014) The Essentials of Risk Management, 2nd Edition, New York: McGraw Hill.
-   Matthews K and J Thompson (2014) The Economics of Banking. 3rd Edition. Chichester UK, John Wiley & Sons.
-   Chris Brooks (2014) Introductory Econometrics for Finance. Cambridge UK: Cambridge University Press.
-   Damodar Gujarati and Dawn Porter (2009) Basic Econometrics, Fifth edition, New York: McGraw-Hill.
-   Bodie, Z., Kane, A. and Marcus, A. (2011), Investments, 9th Edition, McGraw Hill Irwin.
